---
title: "compile-regional-data"
author: "Salmon Watersheds Program, Steph Peacock, Hannah Hunter"
date: "`r format(Sys.time(), '%B %d, %Y')`"
bibliography: "../methods-results/references.json"
csl: "../methods-results/canadian-journal-of-fisheries-and-aquatic-sciences.csl"
link-citations: true
header-includes:
   - \usepackage{draftwatermark}
output: 
 html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: true
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set colour palette for species
# sp_cols <- viridis(5)
# sp_cols <- c("#332288", "#44AA99", "#88CCEE", "#CC6677", "#882255", "#DDCC77") #https://davidmathlogic.com/colorblind
# names(sp_cols) <- c("Chinook", "Chum", "Coho", "Pink", "Sockeye", "Steelhead")
library(dplyr)
library(viridisLite)
library(abind)
library(readxl)
library(ggplot2)
library(cowplot)
library(here)

# Get functions and colours
source(here("code/colours.R"))
source(here("code/functions.R"))
source("https://raw.githubusercontent.com/salmonwatersheds/population-indicators/refs/heads/master/code/functions_general.R")


```

<head>
<meta name="robots" content="noindex">
</head>

<style type="text/css">

.watermark {
  opacity: 0.2;
  position: fixed;
  top: 50%;
  left: 50%;
  font-size: 500%;
  color: #000000;
}

.blackbox {
  padding: 1em;
  background: black;
  color: white;
  border: 2px solid orange;
  border-radius: 10px;
}
.center {
  text-align: center;
}
.figure p.caption {
  font-style: italic;
  color: #555;
  font-size: 14px;
}

</style>


<!--- <div class="watermark">DRAFT</div> -->


# Overview 

The Pacific Salmon Foundation's [State of Salmon Report](https://stateofsalmon.ca) aims to provide the public with answers to general questions regarding the state of salmon in British Columbia (BC) and the Yukon, such as:

* How are salmon populations doing? 
* What are the major factors influencing the state of salmon? 
* What is being done - and what could be done - to mitigate threats and recover salmon populations and the habitats they depend on?

This report, first released in September 2024, takes a data-driven approach to provide historical perspective and context for current news about salmon returns. The State of Salmon Report is being updated annually to reflect new data and changes to the state of Pacific salmon populations. 

As part of this work, PSF's [Salmon Watersheds Program](https://www.salmonwatersheds.ca) has developed simple metrics to report on the status and trends in regional abundance for each species. This document outlines the  specific data sources and analytical approach that we used to estimate time series of spawner abundance and total abundance (including catch) for each species and region, as well as draft results. Our analyses are ongoing, and the information here is subject to change on a regular basis, but we will endeavor to keep it up-to-date.

# Methods

## General Approach

We report spawner abundance for each of ten regions that represent all major Pacific salmon-bearing watersheds in Canada: Yukon, Northern Transboundary, Haida Gwaii, Nass, Skeena, Central Coast, Fraser, East Vancouver Island & Mainland Inlets, West Vancouver Island, and Columbia. These regions are also used to organize data in the [Pacific Salmon Explorer](https://www.salmonexplorer.ca). Note that there was a change from the 2024 Report to split the former Vancouver Island & Mainland Inlets region into two to better reflect the unique ecology and environments of salmon between West Vancouver Island and East Vancouver Island & Mainland Inlets. There are a relatively small number of Pacific salmon that spawn in the MacKenzie River basin in Arctic Canada that are currently not included in the State of Salmon report.

We report on six species of Pacific salmon (*Oncorhynchus* spp.): Chinook, chum, coho, pink, sockeye, and steelhead. For each species, we report on state and trends based on generational averages. For pink salmon, we take the running average of even- and odd-year lineages, but do not separate assessments of even- and odd-year pink populations as is sometimes done in status assessments due to their consistent 2-year life cycle. Our approach of using generation running averages also smooths over dominant years for sockeye salmon, for which many populations display cyclic dominance. Shifts in dominance between even- and odd-year pink populations or declines in sub-dominant years of sockeye salmon are considered in a more nuanced way in our reporting when discussing how changes in abundance have been reflected in the diversity and distribution of each species within the region. 

For each of these species, where data are available, we construct an index of **spawner abundance** at the regional scale. We focus on spawner abundance because these data are more readily accessible and easily summarised at different spatial scales. Spawners represent the abundance of salmon available to reproduce and contribute to future generations, as well as to meet cultural and ecological needs within watersheds, and thus provides a measure of status relevant to communities and ecosystems. We recognize that commercial catch has historically been a substantial portion of salmon that return to the coast, and that ignoring declines in catch will underestimate the declines in overall salmon abundance. There is also increasing rates of pre-spawn mortality (i.e. mortality of salmon after their enter the river but before they successfully spawn) that mean that spawner abundance may be less than the number of fish that survive to maturity Therefore, where data are available, we include information on **total abundance** (i.e. the number of salmon that survive to maturity, calculated as spawners plus catch plus pre-spawn mortality (where applicable)). 

We have based our assessments on the best available data for each region and species. For example, stocks that are governed by international treaties may be monitored by the [Pacific Salmon Commission (PSC)](https://www.psc.org/), and tend to have reliable time series of abundance available at regional scales. These data sources are outlined in the [Region-Specific Data section](#region-specific-data). For species and regions where aggregate abundance is not reported by the PSC or DFO at the scale needed, we adapted our approach to make the best use of available data. In most cases, this meant expanding spawner abundance from stream-level estimates to get a regional scale index of spawner abundance using two types of expansion factors [@EnglishEtAl2018NorthCentralCoast]. This expansion process is described in the [Expansion Factors section](#expansion-factors).  

`R` code and data for the analyses described here are available on GitHub at https://github.com/salmonwatersheds/state-of-salmon.

## Region-Specific Data Sources

Here we describe specific data sources for the abundance of salmon and steelhead at the regional scale. 

```{r load data dependencies, echo=FALSE}

regions <- c("Yukon", "Transboundary", "Haida Gwaii", "Nass", "Skeena", "Central Coast", "East Vancouver Island & Mainland Inlets", "West Vancouver Island", "Fraser", "Columbia")

species_vec <- c("Chinook", "Chum", "Coho", "Pink", "Sockeye", "Steelhead")

# Generation length by species and region (for smoothing)
genLength <- read.csv(here("data/gen_length_regions.csv"))

# Set version of NuSEDS for source_id
nuseds_id <- "PSF_20260428"
ctc_id <- "CTC_20250714"

# Initiate dataframe
sps_data <- NULL

# Dropbox location for archiving
Dropbox_dir <- paste0(get_XDrive(), "1_PROJECTS/1_Active/State of Salmon/2_Data & Analysis/state-of-salmon/")
```

### Yukon
The Canadian portion of the Yukon River is home to Chinook, chum, and coho salmon. Border escapement and total abundance (i.e. run size) of Canadian-origin Chinook and fall chum salmon in the mainstem Yukon River are counted at the Eagle Sonar station on the Yukon/Alaska border, and are available from the [Yukon River Panel](https://www.yukonriverpanel.com/) in their Joint Technical Committee (JTC) Reports. 

There are also Chinook, chum, and coho salmon in the Canadian portion of the Porcupine River, which joins the mainstem Yukon River in Alaska. Data on escapement to the Porcupine River are more patchy and not currently included here.


#### Chinook {.unnumbered}
We used Chinook `RR Spawning escapement estimate` and `RR Canadian origin total run size estimate` from Appendix B11 of @YukonRiverJointTechnicalCommittee2025YukonRiverSalmon. 

<!-- There is no coho compiled here - why listed above? -->

```{r yukon chinook, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of Yukon River Chinook salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}

# Load data ---
ytck <- read.csv(here('data/yukon_chinook_appendixB11.csv'), na.strings = "-") 
# Note: Use Run Reconstruction (RR) estimates of spawning escapement and total
# run size, not "Historical" or Mark-Recapture estimates

# Check if years are continuous
#unique(diff(ytck$Year)) # Yes, all one year apart # commented out to avoid outputting in Rmd

# Reformat data for SPS ---
ytck_sps <- data.frame(
	region = rep("Yukon", dim(ytck)[1]),
	species = rep("Chinook", dim(ytck)[1]),
	year = ytck$Year,
	spawners = ytck$RR.Spawning.escapement.estimate, 
	smoothedSpawners = NA,
	runsize = ytck$RR.Canadian.origin.total.run.size.estimate,
	smoothedRunsize = NA,
	source_id = rep("JTC_202603", dim(ytck)[1])
) 

# Smoothing
ytck_sps$smoothedSpawners <- genSmooth(
	abund = ytck_sps$spawners,
	years = ytck_sps$year,
	genLength = genLength$gen_length[genLength$region == "Yukon" & genLength$species == "Chinook"]
	)

ytck_sps$smoothedRunsize <- genSmooth(
	abund = ytck_sps$runsize,
	years = ytck_sps$year,
	genLength = genLength$gen_length[genLength$region == "Yukon" & genLength$species == "Chinook"]
)

# Add to SPS data
sps_data <- rbind(sps_data, ytck_sps)

# Generate standard plot
plot_abund(ytck_sps)

```

#### Chum {.unnumbered}

Canadian-origin fall chum spawner abundance was taken as the `Spawning escapement estimate` in Appendix B16 and total abundance as the `Total estimated Canadian-origin run size` from Appendix B20 of @YukonRiverJointTechnicalCommittee2025YukonRiverSalmon. 

```{r yukon chum, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of Yukon River chum salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}
# Load data ---
ytcm_B16 <- read.csv(here('data/yukon_chum_appendixB16.csv')) # Historical data
names(ytcm_B16)[1] <- "Year"
ytcm_B20 <- read.csv(here('data/yukon_chum_appendixB20.csv')) # Most recent estimates

# Check if years are continuous
#unique(diff(ytcm_B16$Year)) # Yes, all one year apart
#unique(diff(ytcm_B20$Year)) # Yes, all one year apart

# Reformat data for SPS
ytcm_sps <- data.frame(
	region = rep("Yukon", dim(ytcm_B16)[1]),
	species = rep("Chum", dim(ytcm_B16)[1]),
	year = ytcm_B16$Year,
	spawners = ytcm_B16$Spawning.escapement.estimate, 
	smoothedSpawners = NA,
	runsize = ytcm_B20$Total.estimated.Canadian.origin.run.size[match(ytcm_B16$Year, ytcm_B20$Year)],
	smoothedRunsize = NA,
	source_id = rep("JTC_202603", dim(ytcm_B16)[1])
) 

# Smoothing
ytcm_sps$smoothedSpawners <- genSmooth(
	abund = ytcm_sps$spawners,
	years = ytcm_sps$year,
	genLength = genLength$gen_length[genLength$region == "Yukon" & genLength$species == "Chum"]
)

ytcm_sps$smoothedRunsize <- genSmooth(
	abund = ytcm_sps$runsize,
	years = ytcm_sps$year,
	genLength = genLength$gen_length[genLength$region == "Yukon" & genLength$species == "Chum"]
)

# Add to SPS data
sps_data <- rbind(sps_data, ytcm_sps)

# Generate standard plot
plot_abund(ytcm_sps)


```

```{r yukon chum figs, fig.cap = "**Above:**Comparison of spawning escapement data from two sources in the Yukon River Joint Technical Committee documentation. We use the spawning escapement from Appendix B16 for our spawning escapement metric, while the escapement data from Appendix B20 makes up part of the total abundance metric we use."}
# Compare data from two appendices; note that spawning escapement is sometimes different
plot(ytcm_B16$Year, ytcm_B16$Spawning.escapement.estimate*10^-3, "o", ylab = "Spawning escapement (thousands)", las = 1, bty = "l", xlab = "", pch = 21, bg = "white", main = "Canadian-origin chum salmon")
abline(v = seq(1970, 2024, 2), col = "#00000060", lwd = 0.8, lty = 3)
points(ytcm_B20$Year, ytcm_B20$Spawning.escapement*10^-3, col = 2, pch = 19, cex = 0.6)
legend("topleft", pch= c(1, 19), pt.cex = c(1, 0.6), col = c(1,2), c("Appendix B16", "Appendix B20"))
```


### Northern Transboundary
<!-- > Internal note: Processing of raw excel file from Aaron Foos to extract relevant tables and fields is in [population-data/transboundary-data/code/sos-data-extract.R](https://www.dropbox.com/scl/fi/zxgtulvx2p5iphr3rjx57/sos-data-extract.R?rlkey=f18myjpthqbee4ciz9ny0pjf9&dl=0) -->

The best-available information on salmon abundance in the Northern Transboundary region is from the PSC's [Joint Transboundary Technical Committee Reports](https://www.psc.org/publications/technical-reports/technical-committee-reports/transboundary/). The most recent TTC report [@TTC2022FinalEstimatesTransboundary] includes data through 2021. Updated report appendices were provided by Aaron Foos (DFO) on April 15, 2025 in advance of report publication, and numbers for 2023 and 2024 should be considered **preliminary**.

We recognize that our approach in the Transboundary lacks information from many un-monitored watersheds, in particular the smaller watersheds of the Chilkat, Unuk, and Whiting Rivers. In the absence of better monitoring, we choose to report available data as a proxy for regional abundance, and note that the contribution of these smaller watersheds to total regional abundance for each species is likely small. 


#### Chinook {.unnumbered}
The regional spawner abundance and total return of Chinook was calculated as the sum of escapement and total return, respectively, to the Stikine, Alsek, and Taku indicator stocks. For each of these stocks, we used escapement and exploitation rates provided by the [PSC's Chinook Technical Committee](https://www.psc.org/publications/technical-reports/technical-committee-reports/chinook/) as part of their Synoptic Evaluations of Stock Status. These data were downloaded from their [Data Application](https://psc1.shinyapps.io/ctc-shiny-app/_w_efbbad27/). Total abundance was calculated from the escapement, $E$, and exploitation rate, $r$, as $E/(1-r)$. 

At the time of publication, the Synoptic Evaluations of Stock Status were not yet available for 2024 so we used the escapement estimates for Stikine, Alsek, and Taku Chinook from Table B2 of Appendix B - Detailed Escapement, available as a downloadable dataset from the [Chinook Technical Committee](https://www.psc.org/publications/technical-reports/technical-committee-reports/chinook/ctc-data-sets/). Canadian and US harvest of Chinook was provided by Aaron Foos (DFO) as updated tables from the Transboundary Technical Committee report. Specifically, harvest for Stikine Chinook from Table B12 and harvest of Taku Chinook from Table D8. Total harvest of Alsek Chinook was 235, provided by Teresa Wallace (DFO, pers. comm.): US Commercial: 187, US Subsistence: 8, CAN (Aboriginal): 40, CAN Recreational: 0 (No CAN recreational harvest permitted in 2024).

```{r transboundary chinook, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of Transboundary chinook salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}
# Load data ---
tbrck_raw <- read.csv(here("data/CTC_Synoptic_evaluation_data_all_2025-04-16.csv")) %>%
	filter(Year > 1975) %>% # Remove 1975 -> missing Alsek data
	filter(Stock %in% c("Alsek", "Taku", "Stikine")) %>%
	dplyr::select(Stock, Year, Escapement, RateType, Rate) %>%
	mutate(TotalRun = Escapement/(1 - Rate))

tbrck_esc <- readxl::read_xlsx(here("data/TCCHINOOK-25-02-Appendix-B-Escapement-Detailed.xlsx"), sheet = "B2", range = "A5:G53", col_names = c("year", "Alsek_esc", "Alsek_cv", "Taku_esc", "Taku_cv", "Stikine_esc", "Stikine_cv"))
# Ignore 1975 because not available for Alsek

# For 2024, take harvest from TTC B12, D8, Alsek harvest provided by email (Teresa Wallace, DFO)
tbrck <- tbrck_raw %>%
	group_by(Year) %>%
	summarise(Escapement_all = sum(Escapement),
						TotalRun_all = sum(TotalRun), 
						source_id = "PSC_20250416") %>%
	bind_rows(data.frame(Year = 2024,
						Escapement_all = (tbrck_esc$Alsek_esc + tbrck_esc$Taku_esc + tbrck_esc$Stikine_esc)[tbrck_esc$year == 2024],
						TotalRun_all = 717 + (tbrck_esc$Alsek_esc + tbrck_esc$Taku_esc + tbrck_esc$Stikine_esc)[tbrck_esc$year == 2024], source_id = "Foos_20250415")) 

# Reformat data for SPS
tbrck_sps <- data.frame(
	region = rep("Northern Transboundary", length(tbrck$Year)),
	species = rep("Chinook", length(tbrck$Year)),
	year = tbrck$Year,
	spawners = tbrck$Escapement_all, 
	smoothedSpawners = NA,
	runsize = tbrck$TotalRun_all,
	smoothedRunsize = NA,
	source_id = tbrck$source_id
) 

# Smoothing
tbrck_sps$smoothedSpawners <- genSmooth(
	abund = tbrck_sps$spawners,
	years = tbrck_sps$year,
	genLength = genLength$gen_length[genLength$region == "Northern Transboundary" & genLength$species == "Chinook"]
)

tbrck_sps$smoothedRunsize <- genSmooth(
	abund = tbrck_sps$runsize,
	years = tbrck_sps$year,
	genLength = genLength$gen_length[genLength$region == "Northern Transboundary" & genLength$species == "Chinook"]
)

plot_abund(tbrck_sps)

# Add to master sps dataframe
sps_data <- rbind(sps_data, tbrck_sps)

```

```{r transboundary chinook figs, echo = FALSE, include = TRUE, results = 'asis', fig.width = 9, fig.height = 5, fig.cap = "**Above:** (a) Chinook salmon spawner abundance and (b) total abundance showing the contributions of the Alsek, Taku, and Stikine river basins."}

# Update 2024 data for each system -- different than above, this method from previous Rmd, updates each system
tbrck_raw_updated <- tbrck_raw %>%
	rbind(data.frame(
		Stock = c("Alsek", "Taku", "Stikine"),
		Year = rep(2024, 3),
		Escapement = as.numeric(tbrck_esc %>% filter(year == 2024) %>% dplyr::select("Alsek_esc", "Taku_esc", "Stikine_esc")),
		RateType = rep(NA, 3),
		Rate = rep(NA, 3),
		TotalRun = rep(NA, 3)))

tbrck_raw_updated$TotalRun[tbrck_raw_updated$Year == 2024] <- tbrck_raw_updated$Escapement[tbrck_raw_updated$Year == 2024] + c((40 + 195), (22 + 374), (49 + 37))

# Plot ---

# Spawner abundance
spwn.stack <- tbrck_raw_updated %>% ggplot() +
	geom_area(aes(x=Year, y=Escapement/1000, fill=factor(Stock, levels=c("Stikine", "Taku", "Alsek"))), 
								alpha=0.6) + 
	scale_fill_viridis_d() +
	labs(y="Spawner abundance (thousands)", fill="") + lims(y=c(0,185)) +
	theme_minimal() + theme(legend.position = "none")

# Total abundance
tot.stack <- tbrck_raw_updated %>% ggplot() +
	geom_area(aes(x=Year, y=TotalRun/1000, fill=factor(Stock, levels=c("Stikine", "Taku", "Alsek"))), 
								alpha=0.6) + 
	scale_fill_viridis_d() +
	labs(y="Total abundance (thousands)", fill="") + lims(y=c(0,185)) +
	theme_minimal() 

cowplot::plot_grid(spwn.stack, tot.stack, rel_widths = c(0.75,1), labels="auto")

```

#### Sockeye and Coho {.unnumbered}

```{r transboundary sockeye, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of Transboundary sockeye salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}

# TBR Sockeye data are the sum of the Stikine (Appendix B26) and Taku (Appendix D17) 
# In 2024, the Appendix numbers given were B21 and D15 respectively, but that's changed.
tbrse_stik <- read.csv(here('data/tbr_sockeye_appendixB26.csv'))
tbrse_taku <- read.csv(here('data/tbr_sockeye_appendixD17.csv'))

# Remove Stikine data prior to 1984
tbrse_stik <- subset(tbrse_stik, tbrse_stik$Year >= 1984)

# Reformat data for SPS
tbrse_sps <- data.frame(
	region = rep("Northern Transboundary", dim(tbrse_stik)[1]),
	species = rep("Sockeye", dim(tbrse_stik)[1]),
	year = tbrse_stik$Year,
	spawners = tbrse_stik$StikineRiver_EscapementBroodstock + tbrse_taku$Escapement, 
	smoothedSpawners = NA,
	runsize = tbrse_stik$StikineRiver_TerminalRun + tbrse_taku$Run,
	smoothedRunsize = NA, 
	source_id = rep("Foos_20250415", dim(tbrse_stik)[1])
) 

# Smoothing
tbrse_sps$smoothedSpawners <- genSmooth(
	abund = tbrse_sps$spawners,
	years = tbrse_sps$year,
	genLength = genLength$gen_length[genLength$region == "Northern Transboundary" & genLength$species == "Sockeye"]
)

tbrse_sps$smoothedRunsize <- genSmooth(
	abund = tbrse_sps$runsize,
	years = tbrse_sps$year,
	genLength = genLength$gen_length[genLength$region == "Northern Transboundary" & genLength$species == "Sockeye"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, tbrse_sps)

# Generate standard plot
plot_abund(tbrse_sps)

```

```{r transboundary coho, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of Transboundary coho salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}

# Load data ---

tbrco <- read.csv(here('data/tbr_coho_appendixD22.csv'))

# Check if years are continuous
#unique(diff(tbrco$Year)) # Yes, all one year apart

# Reformat data for SPS
tbrco_sps <- data.frame(
	region = rep("Northern Transboundary", dim(tbrco)[1]),
	species = rep("Coho", dim(tbrco)[1]),
	year = tbrco$Year,
	spawners = tbrco$Escapement, 
	smoothedSpawners = NA,
	runsize = tbrco$TerminalRun,
	smoothedRunsize = NA,
	source_id = rep("Foos_20250415", dim(tbrco)[1])
) 

# Smoothing
tbrco_sps$smoothedSpawners <- genSmooth(
	abund = tbrco_sps$spawners,
	years = tbrco_sps$year,
	genLength = genLength$gen_length[genLength$region == "Northern Transboundary" & genLength$species == "Coho"]
)

tbrco_sps$smoothedRunsize <- genSmooth(
	abund = tbrco_sps$runsize,
	years = tbrco_sps$year,
	genLength = genLength$gen_length[genLength$region == "Northern Transboundary" & genLength$species == "Coho"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, tbrco_sps)

# Generate standard plot
plot_abund(tbrco_sps)

```

#### Pink and Chum {.unnumbered}

Monitoring of pink and chum salmon in the Transboundary region is limited, and the best available information is the Canyon Island fish wheel on the Taku River. 

```{r transboundary pink, fig.cap="**Above:**Spawner abundance of Transboundary pink salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}

# Load data ---
tbrpk <- read.csv(here("data/TTC_ManualExtract_Taku_Pink.csv")) %>%
	filter(!is.na(Value))

# Check if years are continuous
#unique(diff(tbrpk$Year)) # Yes, all one year apart

# Reformat data for SPS
tbrpk_sps <- data.frame(
	region = rep("Northern Transboundary", nrow(tbrpk)), 
	species = rep("Pink", nrow(tbrpk)), 
	year = tbrpk$Year, 
	spawners = tbrpk$Value,
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep("Foos_20250415", nrow(tbrpk))
) 
# Smoothing
tbrpk_sps$smoothedSpawners <- genSmooth(
	abund = tbrpk_sps$spawners,
	years = tbrpk_sps$year,
	genLength = genLength$gen_length[genLength$region == "Northern Transboundary" & genLength$species == "Pink"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, tbrpk_sps)

# Generate standard plot
plot_abund(tbrpk_sps)

```

```{r transboundary chum, fig.cap="**Above:**Spawner abundance of Transboundary chum salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}

# Load data ---
tbrcm <- read.csv(here("data/TTC_ManualExtract_Taku_Chum.csv")) %>%
	filter(!is.na(Value))

# Reformat data for SPS
tbrcm_sps <- data.frame(
	region = rep("Northern Transboundary", nrow(tbrcm)), #length(which(tbrpkcm$species_pooled == "Chum"))),
	species = rep("Chum", nrow(tbrcm)), #length(which(tbrpkcm$species_pooled == "Chum"))),
	year = tbrcm$Year, #sort(tbrpkcm$year[tbrpkcm$species_pooled == "Chum"]),
	spawners = tbrcm$Value, #tbrpkcm$stream_observed_count[tbrpkcm$species_pooled == "Chum"][order(tbrpkcm$year[tbrpkcm$species_pooled == "Chum"])], 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep("Foos_20250415", nrow(tbrcm))
) 
# Smoothing
tbrcm_sps$smoothedSpawners <- genSmooth(
	abund = tbrcm_sps$spawners,
	years = tbrcm_sps$year,
	genLength = genLength$gen_length[genLength$region == "Northern Transboundary" & genLength$species == "Chum"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, tbrcm_sps)

# Generate standard plot
plot_abund(tbrcm_sps)

```

#### Steelhead {.unnumbered}
Steelhead trout have also been enumerated at the Canyon Island fish wheel since 1987, though recent estimates have been patchy and the timing of the fish wheel likely misses a substantial portion of the Taku summer run steelhead. Thus, we do not use the Canyon Island counts.


### Haida Gwaii

#### Chinook {.unnumbered}
The only available data for Chinook salmon in Haida Gwaii are from a single enumeration project on the Yakoun River. Monitoring on the Yakoun has recently been restored. **More information about methods**

These estimates (prior to 2006) also match the estimated escapement to Area 1 in the  @NorthernBoundaryTechnicalCommittee2023USCanadaNorthern (Appendix 30).

```{r haida gwaii chinook, fig.cap="**Above:**Spawner abundance of Haida Gwaii chinook salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average. here, purple points are from NuSEDS spawner surveys and grey are from recent enumeration project."}

# Get spawner survey data to compare
hgck_ss <- read.csv(here("data/spawner_surveys.csv")) %>%
	subset(species_name == "Chinook" & region == "Haida Gwaii") %>%
	subset(stream_name_pse == "YAKOUN RIVER")

hgck <- read.csv(here("data/Yakoun_Chinook_1952-2025.csv"))

# There are some missing years of data; want to impute in between when smoothing
#diff(hgck$year)
hgck_years <- min(hgck$year):max(hgck$year)

# Reformat data for SPS
hgck_sps <- data.frame(
	region = rep("Haida Gwaii", length(hgck_years)),
	species = rep("Chinook", length(hgck_years)),
	year = hgck_years,
	spawners = hgck$stream_observed_count[match(hgck_years, hgck$year)], 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = c(rep(hgck$source_id[1], length(hgck_years[1]:2020)), 
								rep(hgck$source_id[nrow(hgck)], length(2021:2025)))
) 



# Smoothing
hgck_sps$smoothedSpawners <- genSmooth(
	abund = hgck_sps$spawners,
	years = hgck_sps$year,
	genLength = genLength$gen_length[genLength$region == "Haida Gwaii" & genLength$species == "Chinook"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, hgck_sps)

# Generate standard plot
plot_abund(hgck_sps)
points(hgck_ss$year, hgck_ss$stream_observed_count/1000, col="purple")

```

#### Chum {.unnumbered}
For chum, coho, pink, and sockeye in Haida Gwaii, we followed the [Expansion Factors](#expansion-factors) approach described below. We compared our expansion approach to estimates of Area 1 escapement as reported in the @NorthernBoundaryTechnicalCommittee2023USCanadaNorthern (Appendix 30), but found the Area 1 estimates were much lower for chum (see "alternate" series in figure below).

```{r haida gwaii expansion data}
# Load data ---
region_spawners <- sp <- readRDS(here("output/expanded-spawners/Haida Gwaii-spawners.rds"))
yrs <- as.numeric(dimnames(sp)[[3]])
expansion_factors <- readRDS(here("output/expanded-spawners/Haida Gwaii-expansion-factors.rds"))

# PSC Northern Boundary data (Appendix 30) for comparison 
hg_table30 <- read.csv(here("data/TCNB-23-01_Table30_Area1escapement.csv"))
names(hg_table30) <- c("year", "Sockeye", "Coho", "Pink", "Chum", "Chinook") # Change names to be consistent with species
```


```{r haida gwaii chum, results = 'asis', fig.width = 7, fig.height = 4, fig.cap="**Above:** Spawner abundance of Haida Gwaii chum salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}

# Data loaded above
hg.s <- sp[2, "Chum", ]

# Reformat data for SPS
hg.s_sps <- data.frame(
	region = rep("Haida Gwaii", length(hg.s)),
	species = rep("Chum", length(hg.s)),
	year = as.numeric(names(hg.s)),
	spawners = as.numeric(hg.s), 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(nuseds_id, length(hg.s))
) 

# Smoothing
hg.s_sps$smoothedSpawners <- genSmooth(
	abund = hg.s_sps$spawners,
	years = hg.s_sps$year,
	genLength = genLength$gen_length[genLength$region == "Haida Gwaii" & genLength$species == "Chum"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, hg.s_sps)

# Generate standard figure
plot_abund(hg.s_sps)

```


```{r haida gwaii chum figs, results = 'asis', fig.width = 7, fig.height = 6, fig.cap = "**Above:** (a) Time series of spawner abundances showing raw sum of counts for indicator streams (purple) and expanded estimates of spawner abudnance to indicator and non-indicator streams (black). This expanded time series is what is used to assess the current state and trends in spawner abundance for the region and species. Alternative time series from the PSC's Northern Boundary Technical Committee Report (Appendix 30) is shown in blue. (b) The values of Expansion Factor 1 that accounts for unmonitored indicator streams in a given year, and is calculated based on the proportional contribution of the unmonitored indicator stream(s) to the sum of all indicator streams in a (reference) decade. See section on Expansion Factors below for further details."}

# Additional series to plot
added_series.s <- hg_table30 %>% 
		select(year, "Chum") %>%
		rename("value" = "Chum")
	
# Expansion plot(s)
plot_expansions(speciesname = "Chum",
								regionname = "Haida Gwaii",
								region_spawners = region_spawners,
								expansion_factors = expansion_factors,
								added_series = added_series.s)

```


#### Coho {.unnumbered}
For chum, coho, pink, and sockeye in Haida Gwaii, we followed the [Expansion Factors](#expansion-factors) approach described below. We compared our expansion approach to estimates of escapement to Area 1 reported in the @NorthernBoundaryTechnicalCommittee2023USCanadaNorthern (Appendix 30) but found the Area 1 estimates were much lower for coho.
```{r haida gwaii coho, results = 'asis', fig.width = 7, fig.height = 4, fig.cap="**Above:** Spawner abundance of Haida Gwaii coho salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}

# Data loaded above
hg.s <- sp[2, "Coho", ]

# Reformat data for SPS
hg.s_sps <- data.frame(
	region = rep("Haida Gwaii", length(hg.s)),
	species = rep("Coho", length(hg.s)),
	year = as.numeric(names(hg.s)),
	spawners = as.numeric(hg.s), 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(nuseds_id, length(hg.s))
) 

# Smoothing
hg.s_sps$smoothedSpawners <- genSmooth(
	abund = hg.s_sps$spawners,
	years = hg.s_sps$year,
	genLength = genLength$gen_length[genLength$region == "Haida Gwaii" & genLength$species == "Coho"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, hg.s_sps)

# Generate standard figure
plot_abund(hg.s_sps)

```


```{r haida gwaii coho figs, results = 'asis', fig.width = 7, fig.height = 6, fig.cap = "**Above:** (a) Time series of spawner abundances showing raw sum of counts for indicator streams (purple) and expanded estimates of spawner abudnance to indicator and non-indicator streams (black). This expanded time series is what is used to assess the current state and trends in spawner abundance for the region and species. Alternative time series from the PSC's Northern Boundary Technical Committee Report (Appendix 30) is shown in blue. (b) The values of Expansion Factor 1 that accounts for unmonitored indicator streams in a given year, and is calculated based on the proportional contribution of the unmonitored indicator stream(s) to the sum of all indicator streams in a (reference) decade. See section on Expansion Factors below for further details."}

# Additional series to plot
added_series.s <- hg_table30 %>% 
		select(year, "Coho") %>%
		rename("value" = "Coho")
	
# Expansion plot(s)
plot_expansions(speciesname = "Coho",
								regionname = "Haida Gwaii",
								region_spawners = region_spawners,
								expansion_factors = expansion_factors,
								added_series = added_series.s)

```


#### Pink {.unnumbered}
For chum, coho, pink, and sockeye in Haida Gwaii, we followed the [Expansion Factors](#expansion-factors) approach described below. We compared our expansion approach to estimates of Area 1 escapement as reported in the @NorthernBoundaryTechnicalCommittee2023USCanadaNorthern (Appendix 30), but found the Area 1 estimates were much lower for pink (see "alternate" series in figure below).  
```{r haida gwaii pink, results = 'asis', fig.width = 7, fig.height = 4, fig.cap="**Above:** Spawner abundance of Haida Gwaii pink salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}

# Data loaded above
hg.s <- sp[2, "Pink", ]

# Reformat data for SPS
hg.s_sps <- data.frame(
	region = rep("Haida Gwaii", length(hg.s)),
	species = rep("Pink", length(hg.s)),
	year = as.numeric(names(hg.s)),
	spawners = as.numeric(hg.s), 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(nuseds_id, length(hg.s))
) 

# Smoothing
hg.s_sps$smoothedSpawners <- genSmooth(
	abund = hg.s_sps$spawners,
	years = hg.s_sps$year,
	genLength = genLength$gen_length[genLength$region == "Haida Gwaii" & genLength$species == "Pink"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, hg.s_sps)

# Generate standard figure
plot_abund(hg.s_sps)

```


```{r haida gwaii pink figs, results = 'asis', fig.width = 7, fig.height = 6, fig.cap = "**Above:** (a) Time series of spawner abundances showing raw sum of counts for indicator streams (purple) and expanded estimates of spawner abudnance to indicator and non-indicator streams (black). This expanded time series is what is used to assess the current state and trends in spawner abundance for the region and species. Alternative time series from the PSC's Northern Boundary Technical Committee Report (Appendix 30) is shown in blue. (b) The values of Expansion Factor 1 that accounts for unmonitored indicator streams in a given year, and is calculated based on the proportional contribution of the unmonitored indicator stream(s) to the sum of all indicator streams in a (reference) decade. See section on Expansion Factors below for further details."}

# Additional series to plot
added_series.s <- hg_table30 %>% 
		select(year, "Pink") %>%
		rename("value" = "Pink")
	
# Expansion plot(s)
plot_expansions(speciesname = "Pink",
								regionname = "Haida Gwaii",
								region_spawners = region_spawners,
								expansion_factors = expansion_factors,
								added_series = added_series.s)

```

#### Sockeye {.unnumbered}
For chum, coho, pink, and sockeye in Haida Gwaii, we followed the [Expansion Factors](#expansion-factors) approach described below. We compared our expansion approach to estimates of Area 1 escapement as reported in the @NorthernBoundaryTechnicalCommittee2023USCanadaNorthern (Appendix 30), but found the Area 1 estimates were much lower for sockeye (see "alternate" series in figure below).  
```{r haida gwaii sockeye, results = 'asis', fig.width = 7, fig.height = 4, fig.cap="**Above:** Spawner abundance of Haida Gwaii sockeye salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}

# Data loaded above
hg.s <- sp[2, "Sockeye", ]

# Reformat data for SPS
hg.s_sps <- data.frame(
	region = rep("Haida Gwaii", length(hg.s)),
	species = rep("Sockeye", length(hg.s)),
	year = as.numeric(names(hg.s)),
	spawners = as.numeric(hg.s), 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(nuseds_id, length(hg.s))
) 

# Smoothing
hg.s_sps$smoothedSpawners <- genSmooth(
	abund = hg.s_sps$spawners,
	years = hg.s_sps$year,
	genLength = genLength$gen_length[genLength$region == "Haida Gwaii" & genLength$species == "Sockeye"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, hg.s_sps)

# Generate standard figure
plot_abund(hg.s_sps)

```


```{r haida gwaii sockeye figs, results = 'asis', fig.width = 7, fig.height = 6, fig.cap = "**Above:** (a) Time series of spawner abundances showing raw sum of counts for indicator streams (purple) and expanded estimates of spawner abudnance to indicator and non-indicator streams (black). This expanded time series is what is used to assess the current state and trends in spawner abundance for the region and species. Alternative time series from the PSC's Northern Boundary Technical Committee Report (Appendix 30) is shown in blue. (b) The values of Expansion Factor 1 that accounts for unmonitored indicator streams in a given year, and is calculated based on the proportional contribution of the unmonitored indicator stream(s) to the sum of all indicator streams in a (reference) decade. See section on Expansion Factors below for further details."}

# Additional series to plot
added_series.s <- hg_table30 %>% 
		select(year, "Sockeye") %>%
		rename("value" = "Sockeye")
	
# Expansion plot(s)
plot_expansions(speciesname = "Sockeye",
								regionname = "Haida Gwaii",
								region_spawners = region_spawners,
								expansion_factors = expansion_factors,
								added_series = added_series.s)

```


### Nass

#### Chinook {.unnumbered}

Chinook spawners and total return were taken from the [PSC's Chinook Technical Committee (CTC) data sets](https://www.psc.org/publications/technical-reports/technical-committee-reports/chinook/ctc-data-sets/) and are reported on in @ChinookTechnicalCommittee2025AnnualReportCatch. Specifically, Nass River (Area 3) escapement and terminal run are provided in Table B3 (fields `Esc` and `t.run` respectively). In the figure below, these CTC estimates correspond closely to estimates of Nass Chinook available from the [Nisga'a Fisheries Nass Stock Assessment Updates](https://www.dropbox.com/scl/fo/ahrrza0aplzcnamatapar/h?rlkey=n51s6qgu0yze0vrphue8dk4ap&e=2&st=n0g2ox45&dl=0), especially in recent years. Historically, the Nisga'a estimates of total abundance tended to be higher than the CTC estimates. 

```{r nass chinook, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of Nass chinook salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}
# Load data ---
# TCCHINOOK Table B3
nass_ctc <- readxl::read_xlsx(here("data/TCCHINOOK-25-02-Appendix-B-Escapement-Detailed.xlsx"), sheet = "B3", range = "A5:D54", col_types = "numeric", col_names = c("Year", "Above_Gitwinksihlkw2", "Esc", "t.run"))

# TEMPORARY - use Nisgaa Data until CTC is ready # Compare them
nass_nisgaa <- read.csv(here("data/2025NassStockAssessment_Table20+22.csv"))

# Reformat data for SPS
nassck_sps <- data.frame(
	region = rep("Nass", length(nass_nisgaa$Year)),
	species = rep("Chinook", length(nass_nisgaa$Year)),
	year = nass_nisgaa$Year,
	spawners = nass_nisgaa$ESC_Chinook, 
	smoothedSpawners = NA,
	runsize = nass_nisgaa$TOTAL_RUN_Chinook,
	smoothedRunsize = NA,
	source_id = rep("Nisgaa_20251202", length(nass_nisgaa$Year))
	#source_id = rep(ctc_id, length(nass_nisgaa$Year))
) 

# Smoothing
nassck_sps$smoothedSpawners <- genSmooth(
	abund = nassck_sps$spawners,
	years = nassck_sps$year,
	genLength = genLength$gen_length[genLength$region == "Nass" & genLength$species == "Chinook"]
)

nassck_sps$smoothedRunsize <- genSmooth(
	abund = nassck_sps$runsize,
	years = nassck_sps$year,
	genLength = genLength$gen_length[genLength$region == "Nass" & genLength$species == "Chinook"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, nassck_sps)

# Generate standard plot
plot_abund(nassck_sps)
```



#### Sockeye {.unnumbered}

Nass sockeye escapement and total return were derived from the Northern Boundary Sockeye Run Reconstruction (NBSRR) Model that estimates total escapement and total return for the Skeena and Nass Rivers [@EnglishEtAl2017SkeenaSockeyeInriver; @EnglishEtAl2004AssessmentCanadianAlaskan]. Specifically, we used fields `TE` (Total Escapement) and `Total run` for Area 3 sockeye from the 2022 update to the North and Central Coast (NCC) Salmon Database Version 2, maintained by LGL Ltd. with support from PSF [@EnglishEtAl2018NorthCentralCoast; @ChallengerEtAl2018NorthCentralCoast]. For 2023-2025, we obtained the same model output from @NisgaaFisheriesandWildlifeDepartment20242024NassRiver. As can be seen in the figure below, historical numbers from these two sources closely matched, giving us confidence that it was justifiable to combine the two sources.

```{r nass sockeye, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of Nass sockeye salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}

# Load Nisgaa data
nass_nisgaa <- read.csv(here("data/2025NassStockAssessment_Table20+22.csv"))

# use NBSRR data provided by LGL up to 2022 and Nisga'a Post-season for 2023, 2024
nassse <- read.csv(here("data/nassskeena_sockeye_lgl.csv")) %>%
	subset(Region == "Nass") %>%
	mutate(source_id = "English_20230930") %>%
	bind_rows(
		data.frame(Year = c(2023:2025),
							 TE = nass_nisgaa$ESC_Sockeye[nass_nisgaa$Year %in% c(2023:2025)],
							 Total.Run = nass_nisgaa$TOTAL_RUN_Sockeye[nass_nisgaa$Year %in% c(2023:2025)],
							 source_id = rep("Nisgaa_20251202", 3)
		)
	)


# Reformat data for SPS
nassse_sps <- data.frame(
	region = rep("Nass", dim(nassse)[1]),
	species = rep("Sockeye", dim(nassse)[1]),
	year = nassse$Year,
	spawners = nassse$TE, 
	smoothedSpawners = NA,
	runsize = nassse$Total.Run,
	smoothedRunsize = NA, 
	source_id = nassse$source_id
) 

# Smoothing
nassse_sps$smoothedSpawners <- genSmooth(
	abund = nassse_sps$spawners,
	years = nassse_sps$year,
	genLength = genLength$gen_length[genLength$region == "Nass" & genLength$species == "Sockeye"]
)

nassse_sps$smoothedRunsize <- genSmooth(
	abund = nassse_sps$runsize,
	years = nassse_sps$year,
	genLength = genLength$gen_length[genLength$region == "Nass" & genLength$species == "Sockeye"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, nassse_sps)

# Generate standard plot 
plot_abund(nassse_sps)

```



#### Coho {.unnumbered}

Nass coho escapement and total return were available from the Nisga'a Fisheries and Wildlife Department (@NisgaaFisheriesandWildlifeDepartment20242024NassRiver) for 1992 to 2024. These data closely matched previous estimates from @EnglishEtAl2023PreliminaryAreaSalmon used in the 2024 report. Spawner surveys in NuSEDS tended to generally underestimate spawner abudnance to the region, and were not used because other data were available.

```{r nass coho, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of Nass coho salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}

# Reformat data for SPS
nassco_sps <- data.frame(
	region = rep("Nass", length(nass_nisgaa$Year)),
	species = rep("Coho", length(nass_nisgaa$Year)),
	year = nass_nisgaa$Year,
	spawners = nass_nisgaa$ESC_Coho, 
	smoothedSpawners = NA,
	runsize = nass_nisgaa$TOTAL_RUN_Coho,
	smoothedRunsize = NA,
	source_id = rep("Nisgaa_20251202",length(nass_nisgaa$Year))
) 

# Smoothing
nassco_sps$smoothedSpawners <- genSmooth(
	abund = nassco_sps$spawners,
	years = nassco_sps$year,
	genLength = genLength$gen_length[genLength$region == "Nass" & genLength$species == "Coho"]
)

nassco_sps$smoothedRunsize <- genSmooth(
	abund = nassco_sps$runsize,
	years = nassco_sps$year,
	genLength = genLength$gen_length[genLength$region == "Nass" & genLength$species == "Coho"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, nassco_sps)

# Generate standard plot
plot_abund(nassco_sps)

```



#### Pink and Chum {.unnumbered}

Chum and pink salmon spawner abundances were expanded from available spawner surveys from 1950 to 1991, as described below in [Expansion Factors](#expansion-factors). Note that the designation of indicator streams for the Nass region was based on @EnglishEtAl2018NorthCentralCoast rather than the indicator designation in NuSEDS. From 1992 to 2024, estimates of spawner and total abundance were provided in the 2024 post-season report from the Nisga'a Fisheries and Wildlife Department (@NisgaaFisheriesandWildlifeDepartment20242024NassRiver).

```{r nass pink, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of Nass pink salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}
# Use expanded spawner abundance to 1991 then Nisga'a final report data
sp <- readRDS(here("output/expanded-spawners/Nass-spawners.rds"))
yrs <- as.numeric(dimnames(sp)[[3]])

nasspk <- data.frame(
	year = c(yrs, 2025),
	esc = c(round(sp["expanded", "Pink", which(yrs %in% c(1950:1991))]), nass_nisgaa$ESC_Pink),
	run = c(rep(NA, length(c(1950:1991))), nass_nisgaa$TOTAL_RUN_Pink),
	source_id = c(rep(nuseds_id, length(c(1950:1991))), rep("Nisgaa_20251202", length(nass_nisgaa$TOTAL_RUN_Pink)))
)

# Reformat data for SPS
nasspk_sps <- data.frame(
	region = rep("Nass", length(nasspk$year)),
	species = rep("Pink", length(nasspk$year)),
	year = nasspk$year,
	spawners = nasspk$esc,
	smoothedSpawners = NA,
	runsize = nasspk$run,
	smoothedRunsize = NA,
	source_id = nasspk$source_id
)

# Smoothing
nasspk_sps$smoothedSpawners <- genSmooth(
	abund = nasspk_sps$spawners,
	years = nasspk_sps$year,
	genLength = genLength$gen_length[genLength$region == "Nass" & genLength$species == "Pink"]
)

nasspk_sps$smoothedRunsize <- genSmooth(
	abund = nasspk_sps$runsize,
	years = nasspk_sps$year,
	genLength = genLength$gen_length[genLength$region == "Nass" & genLength$species == "Pink"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, nasspk_sps)

# Generate standard figure
plot_abund(nasspk_sps)
	
```

```{r nass chum, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of Nass chum salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}

nasscm <- data.frame(
	year = c(yrs, 2025),
	esc = c(round(sp["expanded", "Chum", which(yrs %in% c(1950:1991))]), nass_nisgaa$ESC_Chum),
	run = c(rep(NA, length(c(1950:1991))), nass_nisgaa$TOTAL_RUN_Chum),
	source_id = c(rep(nuseds_id, length(c(1950:1991))), rep("Nisgaa_20251202", length(nass_nisgaa$TOTAL_RUN_Chum)))
)

# Reformat data for SPS
nasscm_sps <- data.frame(
	region = rep("Nass", length(nasscm$year)),
	species = rep("Chum", length(nasscm$year)),
	year = nasscm$year,
	spawners = nasscm$esc,
	smoothedSpawners = NA,
	runsize = nasscm$run,
	smoothedRunsize = NA,
	source_id = nasscm$source_id
)

# Smoothing
nasscm_sps$smoothedSpawners <- genSmooth(
	abund = nasscm_sps$spawners,
	years = nasscm_sps$year,
	genLength = genLength$gen_length[genLength$region == "Nass" & genLength$species == "Chum"]
)

nasscm_sps$smoothedRunsize <- genSmooth(
	abund = nasscm_sps$runsize,
	years = nasscm_sps$year,
	genLength = genLength$gen_length[genLength$region == "Nass" & genLength$species == "Chum"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, nasscm_sps)
	
#Generate standard figure
plot_abund(nasscm_sps)

```


#### Steelhead {.unnumbered}

The index of Nass steelhead spawner abundance and total return are for the Nass Summer CU, developed in collaboration with the Nisga’a Fish and Wildlife Department and LGL Ltd. [@EnglishEtAl2023PreliminaryAreaSalmon, @NisgaaFisheriesandWildlifeDepartment20242024NassRiver]. There is another steelhead CU in the Nass region - Nass Winter - which is not well monitored and therefore not included in our index of spawner abundance.

```{r nass steelhead, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of Nass steelhead salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}

# Reformat data for SPS
nasssh_sps <- data.frame(
	region = rep("Nass", length(nass_nisgaa$Year)),
	species = rep("Steelhead", length(nass_nisgaa$Year)),
	year = nass_nisgaa$Year,
	spawners = nass_nisgaa$ESC_Steelhead, 
	smoothedSpawners = NA,
	runsize = nass_nisgaa$TOTAL_RUN_Steelhead,
	smoothedRunsize = NA,
	source_id = rep("Nisgaa_20251202",length(nass_nisgaa$Year))
) 

# Smoothing
nasssh_sps$smoothedSpawners <- genSmooth(
	abund = nasssh_sps$spawners,
	years = nasssh_sps$year,
	genLength = genLength$gen_length[genLength$region == "Nass" & genLength$species == "Steelhead"]
)

nasssh_sps$smoothedRunsize <- genSmooth(
	abund = nasssh_sps$runsize,
	years = nasssh_sps$year,
	genLength = genLength$gen_length[genLength$region == "Nass" & genLength$species == "Steelhead"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, nasssh_sps)

# Generate standard figure
plot_abund(nasssh_sps)

```

### Skeena

#### Chinook {.unnumbered}
Chinook spawners were taken from the [PSC's Chinook Technical Committee (CTC) data sets](https://www.psc.org/publications/technical-reports/technical-committee-reports/chinook/ctc-data-sets/) and are reported on in @ChinookTechnicalCommittee2025AnnualReportCatch. Specifically, Skeena River (Area 4) escapement is provided in Table B3 (field `GSI esc`). There is no total return for the Skeena River provided in the CTC data sets.

```{r skeena chinook, fig.cap="**Above:**Spawner abundance of Skeena chinook salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}
# Load data ---
# TCCHINOOK Table B3
sk_ctc <- readxl::read_xlsx(here("data/TCCHINOOK-25-02-Appendix-B-Escapement-Detailed.xlsx"), sheet = "B3", range = "A5:G54", col_types = "numeric", col_names = c("Year", "Nass_Above_Gitwinksihlkw2", "Nass_Esc", "Nass_t.run", "Skeena_Total esc.", "Skeena_GSI esc", "Skeena_GSI SD"))

# Reformat data for SPS
skck_sps <- data.frame(
	region = rep("Skeena", length(sk_ctc$Year)),
	species = rep("Chinook", length(sk_ctc$Year)),
	year = sk_ctc$Year,
	spawners = sk_ctc$`Skeena_GSI esc`, 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(ctc_id, length(sk_ctc$Year))
) 

# Smoothing
skck_sps$smoothedSpawners <- genSmooth(
	abund = skck_sps$spawners,
	years = skck_sps$year,
	genLength = genLength$gen_length[genLength$region == "Skeena" & genLength$species == "Chinook"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, skck_sps)

#Generate standard figure
plot_abund(skck_sps)

```



#### Sockeye {.unnumbered}
Skeena sockeye escapement and total return were derived from the Northern Boundary Sockeye Run Reconstruction (NBSRR) Model that estimates total escapement and total return for the Skeena and Nass Rivers [@EnglishEtAl2017SkeenaSockeyeInriver; @EnglishEtAl2004AssessmentCanadianAlaskan]. Specifically, we used fields `TE` (Total Escapement) and `Total run` for Area 4 sockeye from the 2022 update to the North and Central Coast (NCC) Salmon Database Version 2, maintained by LGL Ltd. with support from PSF [@EnglishEtAl2018NorthCentralCoast; @ChallengerEtAl2018NorthCentralCoast]. Data for 2023 were provided by Charmaine Carr-Harris (DFO) on request. **Data for 2024 are preliminary numbers from Joint Northern Boundary Technical Report [@JointNorthernBoundaryTechnicalCommittee2026USCanadaNorthern]**

```{r skeena sockeye, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of Skeena sockeye salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}

# Compare to NCC run reconstruction: https://github.com/LGLLimited/nccdbv2/tree/master/run/2022-nass-update/data/KarlEnglish-2023-05-29
skse_LGL <- read.csv(here("data/nassskeena_sockeye_lgl.csv")) %>%
	subset(Region == "Skeena") %>%
	mutate(source_id = "English_20230930")

# From PSC Run Size app:
skse_ch2 <- read.csv(here("data/Chapter2_Data_2025_07_22.csv")) %>%
	filter(Region == "Skeena River") %>%
	mutate(source_id = "PSC_20250722")

# Add 2023 data from Charmaine (pers. comm. July 24 2025)
skse_2023 <- data.frame(
	Year = 2023,
	TE = 1437788,
	catch = 656984
) %>%
	mutate(Total.Run = TE + catch,
	source_id = "CarrHarris_20250724")

# Add 2024 data from Joint Northern Boundary Tech Report
skse_2024 <- data.frame(
	Year = 2024,
	TE = 2683467,
	catch = NA
) %>%
	mutate(Total.Run = 2807467,
	source_id = NA)

# combine sources
skse <- skse_ch2[ , c("Year", "Spawners", "Returns", "source_id")] %>%
	rename(TE = Spawners,
				 Total.Run  = Returns) %>%
	bind_rows(skse_LGL[which(skse_LGL$Year < 1982), c("Year", "TE", "Total.Run", "source_id")]) %>%
	bind_rows(skse_2023[ c("Year", "TE", "Total.Run", "source_id")]) %>%
	bind_rows(skse_2024[ c("Year", "TE", "Total.Run", "source_id")]) %>%
	arrange(Year)

	
#unique(diff(skse$Year)) # Yes, all one year apart

# Reformat data for SPS
skse_sps <- data.frame(
	region = rep("Skeena", dim(skse)[1]),
	species = rep("Sockeye", dim(skse)[1]),
	year = skse$Year,
	spawners = skse$TE, 
	smoothedSpawners = NA,
	runsize = skse$Total.Run,
	smoothedRunsize = NA,
	source_id = skse$source_id
) 

# Smoothing
skse_sps$smoothedSpawners <- genSmooth(
	abund = skse_sps$spawners,
	years = skse_sps$year,
	genLength = genLength$gen_length[genLength$region == "Skeena" & genLength$species == "Sockeye"]
)

skse_sps$smoothedRunsize <- genSmooth(
	abund = skse_sps$runsize,
	years = skse_sps$year,
	genLength = genLength$gen_length[genLength$region == "Skeena" & genLength$species == "Sockeye"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, skse_sps)

# Generate standard figure 
plot_abund(skse_sps)
```



#### Steelhead {.unnumbered}
The index of Skeena steelhead spawner abundance is derived from estimated escapement of Skeena Summer steelhead at the Tyee Test Fishery (1956 - present), provided by the Province via email. As for other regions, these estimates may not capture winter-run steelhead, for which data are not available.

```{r skeena steelhead, fig.cap="**Above:**Spawner abundance of Skeena steelhead salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}
# Load data ---
# Use Skeena steelhead run size from BC Updates
sksh <- read.csv(here("data/SkeenaSteelhead_1956-2025.csv"))


# Reformat data for SPS
sksh_sps <- data.frame(
	region = rep("Skeena", dim(sksh)[1]),
	species = rep("Steelhead", dim(sksh)[1]),
	year = sksh$Year,
	spawners = sksh$TyeeEscapement, 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = "ProvBC_20250903"
) 

# Smoothing
sksh_sps$smoothedSpawners <- genSmooth(
	abund = sksh_sps$spawners,
	years = sksh_sps$year,
	genLength = genLength$gen_length[genLength$region == "Skeena" & genLength$species == "Steelhead"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, sksh_sps)

# Generate standard figure
plot_abund(sksh_sps)
```


#### Chum {.unnumbered}
Chum salmon spawner abundances in the Skeena were expanded from available spawner surveys, as described below in [Expansion Factors](#expansion-factors). Note that the designation of indicator streams for the Skeena region was based on @EnglishEtAl2018NorthCentralCoast rather than the indicator designation in NuSEDS.

```{r skeena expansion data}
region_spawners <- sp <- readRDS(here("output/expanded-spawners/Skeena-spawners.rds"))
yrs <- as.numeric(dimnames(sp)[[3]])
expansion_factors <- readRDS(here("output/expanded-spawners/Skeena-expansion-factors.rds"))

sk_table32 <- read.csv(here("data/TCNB-23-01_Table32_Area4escapement.csv"))
names(sk_table32) <- c("year", "Sockeye", "Coho", "Pink", "Chum", "Chinook") # Change names to be consistent with species

```


```{r skeena chum, results = 'asis', fig.width = 7, fig.height = 4, fig.cap="**Above:**Spawner abundance of Skeena chum salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}

## Process data 
skeena.s <- sp[2, "Chum", ]
skeena.s <- skeena.s[!is.na(skeena.s)]

# Reformat data for SPS
skeena.s_sps <- data.frame(
	region = rep("Skeena", length(skeena.s)),
	species = rep("Chum", length(skeena.s)),
	year = as.numeric(names(skeena.s)),
	spawners = as.numeric(skeena.s), 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(nuseds_id, length(skeena.s))
) 

# Smoothing
skeena.s_sps$smoothedSpawners <- genSmooth(
	abund = skeena.s_sps$spawners,
	years = skeena.s_sps$year,
	genLength = genLength$gen_length[genLength$region == "Skeena" & genLength$species == "Chum"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, skeena.s_sps)

# Generate standard figure
plot_abund(skeena.s_sps)
```


```{r skeena chum figs, echo = FALSE, include = TRUE, results = 'asis', fig.width = 7, fig.height = 6, fig.cap = "**Above:** (a) Time series of spawner abundances showing raw sum of counts for indicator streams (purple) and expanded estimates of spawner abudnance to indicator and non-indicator streams (black). This expanded time series is what is used to assess the current state and trends in spawner abundance for the region and species. Alternative time series from the PSC's Northern Boundary Technical Committee Report (Appendix 32) is shown in blue. (b) The values of Expansion Factor 1 that accounts for unmonitored indicator streams in a given year, and is calculated based on the proportional contribution of the unmonitored indicator stream(s) to the sum of all indicator streams in a (reference) decade. See section on Expansion Factors below for further details."}

added_series.s <- sk_table32 %>% 
		select(year, "Chum") %>%
		rename("value" = "Chum") 
	
plot_expansions(speciesname = "Chum",
									regionname = "Skeena",
									region_spawners = region_spawners,
									expansion_factors = expansion_factors,
									added_series = added_series.s)

```



#### Coho {.unnumbered}
Coho salmon spawner abundances in the Skeena were expanded from available spawner surveys, as described below in [Expansion Factors](#expansion-factors). Note that the designation of indicator streams for the Skeena region was based on @EnglishEtAl2018NorthCentralCoast rather than the indicator designation in NuSEDS.
```{r skeena coho, results = 'asis', fig.width = 7, fig.height = 4, fig.cap="**Above:**Spawner abundance of Skeena coho salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}

## Process data 
skeena.s <- sp[2, "Coho", ]
skeena.s <- skeena.s[!is.na(skeena.s)]

# Reformat data for SPS
skeena.s_sps <- data.frame(
	region = rep("Skeena", length(skeena.s)),
	species = rep("Coho", length(skeena.s)),
	year = as.numeric(names(skeena.s)),
	spawners = as.numeric(skeena.s), 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(nuseds_id, length(skeena.s))
) 

# Smoothing
skeena.s_sps$smoothedSpawners <- genSmooth(
	abund = skeena.s_sps$spawners,
	years = skeena.s_sps$year,
	genLength = genLength$gen_length[genLength$region == "Skeena" & genLength$species == "Coho"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, skeena.s_sps)

# Generate standard figure
plot_abund(skeena.s_sps)
```


```{r skeena coho figs, echo = FALSE, include = TRUE, results = 'asis', fig.width = 7, fig.height = 6, fig.cap = "**Above:** (a) Time series of spawner abundances showing raw sum of counts for indicator streams (purple) and expanded estimates of spawner abudnance to indicator and non-indicator streams (black). This expanded time series is what is used to assess the current state and trends in spawner abundance for the region and species. Alternative time series from the PSC's Northern Boundary Technical Committee Report (Appendix 32) is shown in blue. (b) The values of Expansion Factor 1 that accounts for unmonitored indicator streams in a given year, and is calculated based on the proportional contribution of the unmonitored indicator stream(s) to the sum of all indicator streams in a (reference) decade. See section on Expansion Factors below for further details."}

added_series.s <- sk_table32 %>% 
		select(year, "Coho") %>%
		rename("value" = "Coho") 
	
plot_expansions(speciesname = "Coho",
									regionname = "Skeena",
									region_spawners = region_spawners,
									expansion_factors = expansion_factors,
									added_series = added_series.s)

```



#### Pink {.unnumbered}
Skeena pink salmon spawner abundances were expanded from available spawner surveys, as described below in [Expansion Factors](#expansion-factors). Note that the designation of indicator streams for the Skeena region was based on @EnglishEtAl2018NorthCentralCoast rather than the indicator designation in NuSEDS.
```{r skeena pink, results = 'asis', fig.width = 7, fig.height = 4, fig.cap="**Above:**Spawner abundance of Haida Gwaii chinook salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}

## Process data 
skeena.s <- sp[2, "Pink", ]
skeena.s <- skeena.s[!is.na(skeena.s)]

# Reformat data for SPS
skeena.s_sps <- data.frame(
	region = rep("Skeena", length(skeena.s)),
	species = rep("Pink", length(skeena.s)),
	year = as.numeric(names(skeena.s)),
	spawners = as.numeric(skeena.s), 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(nuseds_id, length(skeena.s))
) 

# Smoothing
skeena.s_sps$smoothedSpawners <- genSmooth(
	abund = skeena.s_sps$spawners,
	years = skeena.s_sps$year,
	genLength = genLength$gen_length[genLength$region == "Skeena" & genLength$species == "Pink"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, skeena.s_sps)

# Generate standard figure
plot_abund(skeena.s_sps)
```


```{r skeena pink figs, echo = FALSE, include = TRUE, results = 'asis', fig.width = 7, fig.height = 6, fig.cap = "**Above:** (a) Time series of spawner abundances showing raw sum of counts for indicator streams (purple) and expanded estimates of spawner abudnance to indicator and non-indicator streams (black). This expanded time series is what is used to assess the current state and trends in spawner abundance for the region and species. Alternative time series from the PSC's Northern Boundary Technical Committee Report (Appendix 32) is shown in bkue. (b) The values of Expansion Factor 1 that accounts for unmonitored indicator streams in a given year, and is calculated based on the proportional contribution of the unmonitored indicator stream(s) to the sum of all indicator streams in a (reference) decade. See section on Expansion Factors below for further details."}

added_series.s <- sk_table32 %>% 
		select(year, "Pink") %>%
		rename("value" = "Pink") 
	
plot_expansions(speciesname = "Pink",
									regionname = "Skeena",
									region_spawners = region_spawners,
									expansion_factors = expansion_factors,
									added_series = added_series.s)

```


### Central Coast

#### Chinook {.unnumbered}

For Central Coast Chinook, the PSC's Chinook Technical Committee (CTC) reported escapement for two indicator stocks on the Central Coast: the Atnarko River and Rivers Inlet [@ChinookTechnicalCommittee2025AnnualReportCatch]. Rivers Inlet is monitored "using an index of escapements to the Wannock, Kilbella, and Chuckwalla rivers" [@ChinookTechnicalCommittee2025AnnualReportCatch]. Because monitoring of Central Coast Chinook reported in NuSEDS has recently declined, expansions from spawner surveys are unreliable. As such, in 2025 we moved to using a composite indicator equal to the sum of the Atnarko River (`Total esc.`) and Rivers Inlet index reported by the CTC in Appendix Table B3.

```{r central coast chinook, fig.cap="**Above:**Spawner abundance of Central Coast chinook salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}

## Load data ---
# TCCHINOOK Table B3
cc_ctc <- readxl::read_xlsx(here("data/TCCHINOOK-25-02-Appendix-B-Escapement-Detailed.xlsx"), sheet = "B3", range = "H5:K54", col_types = "numeric", col_names = c("Atnarko_Total_esc", "Atnarko_CV", "Atnarko_Wild", "Rivers_Inlet")) %>%
	mutate(Year = c(1975:2024))

# Reformat data for SPS
ccck_sps <- data.frame(
	region = rep("Central Coast", length(cc_ctc$Year)),
	species = rep("Chinook", length(cc_ctc$Year)),
	year = cc_ctc$Year,
	spawners = cc_ctc$Atnarko_Total_esc + cc_ctc$Rivers_Inlet, 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(ctc_id, length(cc_ctc$Year))
) 

# Smoothing
ccck_sps$smoothedSpawners <- genSmooth(
	abund = ccck_sps$spawners,
	years = ccck_sps$year,
	genLength = genLength$gen_length[genLength$region == "Central Coast" & genLength$species == "Chinook"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, ccck_sps)

#Generate standard figure
plot_abund(ccck_sps)

```


```{r central coast chinook figs, fig.cap = "**Above:** Comparison of Chinook escapements on the Central Coast. The sum of Atnarko escapement (red) and Rivers Inlet escapement index (blue) from the CTC (purple points) was used in 2025. The expanded data from NuSEDS, not used, are in black with expansion factors used shown by grey 'x'"}
## Code to generate index comparison plot:
# Expanded spawner estimate
ccck_ex <- readRDS(here("output/expanded-spawners/Central Coast-spawners.rds"))[2, "Chinook", ]
yrs <- as.numeric(names(ccck_ex))

e <- readRDS(here("output/expanded-spawners/Central Coast-expansion-factors.rds"))[[1]]

# TCCHINOOK Table B3
cc_ctc <- readxl::read_xlsx(here("data/TCCHINOOK-25-02-Appendix-B-Escapement-Detailed.xlsx"), sheet = "B3", 
														range = "H5:K54", col_types = "numeric", 
														col_names = c("Atnarko_Total_esc", "Atnarko_CV", 
																					"Atnarko_Wild", "Rivers_Inlet")) %>%
	mutate(Year = c(1975:2024))

par(mar = c(3,4,2,4))
plot(yrs, e$exp1, pch = 4, bty = "n", xaxt = "n", yaxt = "n", xlab = "", ylab = "", col = grey(0.6), ylim = c(0, 30), lwd = 2)
axis(side = 4, col = grey(0.6), las = 1)
mtext(side = 4, line = 2.5, "Expansion factor x", col = grey(0.6))
par(new = TRUE)

plot(yrs, ccck_ex*10^-3, "o", pch = 21, bg = "white", bty = "l", xlab = "", ylab = "Spawner abundance (thousands)", ylim = c(0, 150), las = 1, yaxs = "i")
abline(v = seq(1950, 2025, 5), lty = 3, col = grey(0.8))
abline(h = seq(0, 150, 50), lty = 3, col = grey(0.8))
points(cc_ctc$Year, cc_ctc$Atnarko_Total_esc*10^-3, "o", pch = 21, bg = "white", col = 2)
points(cc_ctc$Year, cc_ctc$Rivers_Inlet*10^-3, "o", pch = 21, bg = "white", col = 4)
points(cc_ctc$Year, cc_ctc$Atnarko_Total_esc*10^-3 + cc_ctc$Rivers_Inlet*10^-3, pch = 19, col = "purple", cex = 0.6)

legend("topleft", col = c(1,2,4, "purple"), c("NuSEDS expansion", "CTC Atnarko Total esc.", "CTC Rivers Inlet index esc.", "CTC sum"), pch = c(rep(21, 3), 19), pt.bg = "white", lwd = c(rep(1, 3),NA), bty = "n")

```



#### Steelhead {.unnumbered}

There are two spawner surveys for Steelhead trout on the Central Coast that we report in the Pacific Salmon Explorer. However, these surveys have been patchy through time with the most recent estimates from 2016. Therefore, spawner abundance and total return for Central Coast steelhead was considered unknown.


#### Chum {.unnumbered}

Estimates of spawner abundance for chum salmon on the Central Coast were expanded from available spawner surveys, as described below in [Expansion Factors](#expansion-factors). Monitoring and reporting of observed spawners on the Central Coast has declined in recent decades, thus requiring larger expansion factors to arrive at an area-wide estimate that can be compared through time. As such, estimates of spawner abundance for the Central Coast over the past decade are more uncertain and should be interpreted with caution. 

```{r central coast expansion data}
region_spawners <- sp <- readRDS(here("output/expanded-spawners/Central Coast-spawners.rds"))
yrs <- as.numeric(dimnames(sp)[[3]])
expansion_factors <- readRDS(here("output/expanded-spawners/Central Coast-expansion-factors.rds"))

```


```{r central coast chum, results = 'asis', fig.width = 7, fig.height = 4, fig.cap="**Above:** Spawner abundance of Central Coast chum salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}

## Process data 
cc.s <- sp[2, "Chum", ]

# Reformat data for SPS
cc.s_sps <- data.frame(
	region = rep("Central Coast", length(cc.s)),
	species = rep("Chum", length(cc.s)),
	year = as.numeric(names(cc.s)),
	spawners = as.numeric(cc.s), 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(nuseds_id, length(cc.s)))

# Smoothing
cc.s_sps$smoothedSpawners <- genSmooth(
	abund = cc.s_sps$spawners,
	years = cc.s_sps$year,
	genLength = genLength$gen_length[genLength$region == "Central Coast" & genLength$species == "Chum"])

# Add to master sps dataframe
sps_data <- rbind(sps_data, cc.s_sps)

# Generate standard figure
plot_abund(cc.s_sps)
```


```{r central coast chum figs, echo = FALSE, include = TRUE, results = 'asis', fig.width = 7, fig.height = 6, fig.cap = "**Above:** (a) Time series of spawner abundances showing raw sum of counts for indicator streams (red) and expanded estimates of spawner abudnance to indicator and non-indicator streams (black). This expanded time series is what is used to assess the current state and trends in spawner abundance for the region and species. (b) The values of Expansion Factor 1 that accounts for unmonitored indicator streams in a given year, and is calculated based on the proportional contribution of the unmonitored indicator stream(s) to the sum of all indicator streams in a (reference) decade."}

plot_expansions(speciesname = "Chum",
								regionname = "Central Coast",
								region_spawners = region_spawners,
								expansion_factors = expansion_factors,
								added_series = NULL)

```

#### Coho {.unnumbered}

Estimates of spawner abundance for coho salmon on the Central Coast were expanded from available spawner surveys, as described below in [Expansion Factors](#expansion-factors). Monitoring and reporting of observed spawners on the Central Coast has declined in recent decades, thus requiring larger expansion factors to arrive at an area-wide estimate that can be compared through time. As such, estimates of spawner abundance for the Central Coast over the past decade are more uncertain and should be interpreted with caution. 

```{r central coast coho, results = 'asis', fig.width = 7, fig.height = 4, fig.cap="**Above:** Spawner abundance of Central Coast coho salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}

## Process data 
cc.s <- sp[2, "Coho", ]

# Reformat data for SPS
cc.s_sps <- data.frame(
	region = rep("Central Coast", length(cc.s)),
	species = rep("Coho", length(cc.s)),
	year = as.numeric(names(cc.s)),
	spawners = as.numeric(cc.s), 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(nuseds_id, length(cc.s)))

# Smoothing
cc.s_sps$smoothedSpawners <- genSmooth(
	abund = cc.s_sps$spawners,
	years = cc.s_sps$year,
	genLength = genLength$gen_length[genLength$region == "Central Coast" & genLength$species == "Coho"])

# Add to master sps dataframe
sps_data <- rbind(sps_data, cc.s_sps)

# Generate standard figure
plot_abund(cc.s_sps)
```


```{r central coast coho figs, echo = FALSE, include = TRUE, results = 'asis', fig.width = 7, fig.height = 6, fig.cap = "**Above:** (a) Time series of spawner abundances showing raw sum of counts for indicator streams (red) and expanded estimates of spawner abudnance to indicator and non-indicator streams (black). This expanded time series is what is used to assess the current state and trends in spawner abundance for the region and species. (b) The values of Expansion Factor 1 that accounts for unmonitored indicator streams in a given year, and is calculated based on the proportional contribution of the unmonitored indicator stream(s) to the sum of all indicator streams in a (reference) decade."}

plot_expansions(speciesname = "Coho",
								regionname = "Central Coast",
								region_spawners = region_spawners,
								expansion_factors = expansion_factors,
								added_series = NULL)

```

#### Pink {.unnumbered}

Estimates of spawner abundance for pink salmon on the Central Coast were expanded from available spawner surveys, as described below in [Expansion Factors](#expansion-factors). Monitoring and reporting of observed spawners on the Central Coast has declined in recent decades, thus requiring larger expansion factors to arrive at an area-wide estimate that can be compared through time. As such, estimates of spawner abundance for the Central Coast over the past decade are more uncertain and should be interpreted with caution. 

```{r central coast pink, results = 'asis', fig.width = 7, fig.height = 4, fig.cap="**Above:** Spawner abundance of Central Coast pink salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}

## Process data 
cc.s <- sp[2, "Pink", ]

# Reformat data for SPS
cc.s_sps <- data.frame(
	region = rep("Central Coast", length(cc.s)),
	species = rep("Pink", length(cc.s)),
	year = as.numeric(names(cc.s)),
	spawners = as.numeric(cc.s), 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(nuseds_id, length(cc.s)))

# Smoothing
cc.s_sps$smoothedSpawners <- genSmooth(
	abund = cc.s_sps$spawners,
	years = cc.s_sps$year,
	genLength = genLength$gen_length[genLength$region == "Central Coast" & genLength$species == "Pink"])

# Add to master sps dataframe
sps_data <- rbind(sps_data, cc.s_sps)

# Generate standard figure
plot_abund(cc.s_sps)
```


```{r central coast pink figs, echo = FALSE, include = TRUE, results = 'asis', fig.width = 7, fig.height = 6, fig.cap = "**Above:** (a) Time series of spawner abundances showing raw sum of counts for indicator streams (red) and expanded estimates of spawner abudnance to indicator and non-indicator streams (black). This expanded time series is what is used to assess the current state and trends in spawner abundance for the region and species. (b) The values of Expansion Factor 1 that accounts for unmonitored indicator streams in a given year, and is calculated based on the proportional contribution of the unmonitored indicator stream(s) to the sum of all indicator streams in a (reference) decade."}

plot_expansions(speciesname = "Pink",
								regionname = "Central Coast",
								region_spawners = region_spawners,
								expansion_factors = expansion_factors,
								added_series = NULL)

```


#### Sockeye {.unnumbered}

Estimates of spawner abundance for sockeye salmon on the Central Coast were expanded from available spawner surveys, as described below in [Expansion Factors](#expansion-factors). Monitoring and reporting of observed spawners on the Central Coast has declined in recent decades, thus requiring larger expansion factors to arrive at an area-wide estimate that can be compared through time. As such, estimates of spawner abundance for the Central Coast over the past decade are more uncertain and should be interpreted with caution. 

```{r central coast sockeye, results = 'asis', fig.width = 7, fig.height = 4, fig.cap="**Above:** Spawner abundance of Central Coast sockeye salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}

## Process data 
cc.s <- sp[2, "Sockeye", ]

# Reformat data for SPS
cc.s_sps <- data.frame(
	region = rep("Central Coast", length(cc.s)),
	species = rep("Sockeye", length(cc.s)),
	year = as.numeric(names(cc.s)),
	spawners = as.numeric(cc.s), 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(nuseds_id, length(cc.s)))

# Smoothing
cc.s_sps$smoothedSpawners <- genSmooth(
	abund = cc.s_sps$spawners,
	years = cc.s_sps$year,
	genLength = genLength$gen_length[genLength$region == "Central Coast" & genLength$species == "Sockeye"])

# Add to master sps dataframe
sps_data <- rbind(sps_data, cc.s_sps)

# Generate standard figure
plot_abund(cc.s_sps)
```


```{r central coast sockeye figs, echo = FALSE, include = TRUE, results = 'asis', fig.width = 7, fig.height = 6, fig.cap = "**Above:** (a) Time series of spawner abundances showing raw sum of counts for indicator streams (red) and expanded estimates of spawner abudnance to indicator and non-indicator streams (black). This expanded time series is what is used to assess the current state and trends in spawner abundance for the region and species. (b) The values of Expansion Factor 1 that accounts for unmonitored indicator streams in a given year, and is calculated based on the proportional contribution of the unmonitored indicator stream(s) to the sum of all indicator streams in a (reference) decade."}

plot_expansions(speciesname = "Sockeye",
								regionname = "Central Coast",
								region_spawners = region_spawners,
								expansion_factors = expansion_factors,
								added_series = NULL)

```


<!-- Note that the designation of indicator streams for the Central Coast region was based on @EnglishEtAl2018NorthCentralCoast rather than the indicator designation in NuSEDS. SP: Not the case for 2025 update. -->

### East Vancouver Island & Mainland Inlets

In the 2025 State of Salmon report, we have split the former "Vancouver Island & Mainland Inlets" region into two: East Vancouver Island & Mainland Inlets and West Vancouver Island. This split was made to better reflect the different ecology and habitats of salmon between the regions that may lead to different status and trends.

#### Chinook {.unnumbered}

PSC's Chinook Technical Committee (CTC) reports escapement (spawner abundance) and run size (total abundance) for three indicator stocks in the Strait of Georgia: Nanaimo, Cowichan, and Phillips [@ChinookTechnicalCommittee2025AnnualReportCatch]. We report spawner abundance and total abundance as the sum of `Esc` and `t.run`, respectively, across these three indicator stocks as reported in Appendix Table B4 available from the [PSC's Chinook Technical Committee (CTC) data sets](https://www.psc.org/publications/technical-reports/technical-committee-reports/chinook/ctc-data-sets/).

```{r east VI chinook, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of East Vancouver Island & Mainland Inlets chinook salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}
## Load data --
# (1) CTC data
evimi_ctc <- readxl::read_xlsx(here("data/TCCHINOOK-25-02-Appendix-B-Escapement-Detailed.xlsx"), sheet = "B4", range = "A5:G54", col_types = "numeric", col_names = c("Year", "Nanaimo_esc", "Nanaimo_trun", "Cowichan_esc", "Cowichan_trun", "Phillips_esc", "Phillips_trun")) # Southern BC

evimi_ctc_summed <- evimi_ctc %>%
	group_by(Year) %>%
	summarise(esc_sum = sum(Nanaimo_esc, Cowichan_esc, Phillips_esc),
						trun_sum = sum(Nanaimo_trun, Cowichan_trun, Phillips_trun))

# Put in SPS format
evimick_sps <- data.frame(
	region = rep("East Vancouver Island & Mainland Inlets", dim(evimi_ctc_summed)[1]),
	species = rep("Chinook", dim(evimi_ctc_summed)[1]),
	year = evimi_ctc_summed$Year,
	spawners = evimi_ctc_summed$esc_sum, 
	smoothedSpawners = NA,
	runsize = evimi_ctc_summed$trun_sum,
	smoothedRunsize = NA,
	source_id = rep(ctc_id, dim(evimi_ctc_summed)[1])
) 

# Smoothing
evimick_sps$smoothedSpawners <- genSmooth(
	abund = evimick_sps$spawners,
	years = evimick_sps$year,
	genLength = genLength$gen_length[genLength$region == "East Vancouver Island & Mainland Inlets" & genLength$species == "Chinook"]
)

evimick_sps$smoothedRunsize <- genSmooth(
	abund = evimick_sps$runsize,
	years = evimick_sps$year,
	genLength = genLength$gen_length[genLength$region == "East Vancouver Island & Mainland Inlets" & genLength$species == "Chinook"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, evimick_sps)

#Generate standard figures
plot_abund(evimick_sps)
```


```{r east VI chinook figs, echo = FALSE, include = TRUE, results = 'asis', fig.width = 9, fig.height = 5, fig.cap = "**Above:** (a) Chinook salmon spawner abundance and (b) total abundance showing the contributions of the Cowichan, Nanaimo, and Philips indicator stocks."}

evimi_ctc <- readxl::read_xlsx(here("data/TCCHINOOK-25-02-Appendix-B-Escapement-Detailed.xlsx"), sheet = "B4", range = "A5:G54", col_types = "numeric", col_names = c("Year", "Nanaimo_esc", "Nanaimo_trun", "Cowichan_esc", "Cowichan_trun", "Phillips_esc", "Phillips_trun")) # Southern BC
	

# Spawner abundance
spwn.stack <- evimi_ctc %>% 
	tidyr::pivot_longer(
    cols = -Year,
    names_to = c("stock", ".value"),  # .value keeps variable names (height, weight)
    names_sep = "_") %>%
	ggplot() +
	geom_area(aes(x=Year, y=esc/1000, fill=stock), 
								alpha=0.6) + 
	scale_fill_viridis_d() +
	labs(y="Spawner abundance (thousands)", fill="") + lims(y=c(0,40)) +
	theme_minimal() + theme(legend.position = "none")

# Total abundance
tot.stack <- evimi_ctc %>% 
	tidyr::pivot_longer(
    cols = -Year,
    names_to = c("stock", ".value"),  # .value keeps variable names (height, weight)
    names_sep = "_") %>%
	ggplot() +
	geom_area(aes(x=Year, y=trun/1000, fill=stock), 
								alpha=0.6) + 
	scale_fill_viridis_d() +
	labs(y="Spawner abundance (thousands)", fill="") + lims(y=c(0,40)) +
	theme_minimal() + theme(legend.position = "right")

suppressWarnings(
cowplot::plot_grid(spwn.stack, tot.stack, rel_widths = c(0.75,1), labels="auto") )

```


#### Chum {.unnumbered}
Initial time series of spawner abundances for chum salmon in East Vancouver Island & Mainland Inlets were expanded from available spawner surveys, as described below in [Expansion Factors](#expansion-factors). For the assessments, we truncated the time series of spawner abundance to years 1953+ because earlier years had high expansion factor values indicating lack of monitoring on key indicator streams (see figures below). 

```{r east VI expansion data}
region_spawners <- sp <- readRDS(here("output/expanded-spawners/East Vancouver Island & Mainland Inlets-spawners.rds"))
yrs <- as.numeric(dimnames(sp)[[3]])
expansion_factors <- readRDS(here("output/expanded-spawners/East Vancouver Island & Mainland Inlets-expansion-factors.rds"))
```

```{r east VI chum, results = 'asis', fig.width = 7, fig.height = 4, fig.cap="**Above:** Spawner abundance of East Vancouver Island & Mainland Inlets chum salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}
## Process data
evimi.s <- sp[2, "Chum", ]

# Get rid of NAs (esp in 2024)
evimi.s <- evimi.s[1:max(which(!is.na(evimi.s)))]

# For all VIMI species, start in 1953 - earlier expansion factors are large
evimi.s	<- evimi.s[which(as.numeric(names(evimi.s)) >= 1953)]

# Reformat data for SPS
evimi.s_sps <- data.frame(
	region = rep("East Vancouver Island & Mainland Inlets", length(evimi.s)),
	species = rep("Chum", length(evimi.s)),
	year = as.numeric(names(evimi.s)),
	spawners = as.numeric(evimi.s), 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(nuseds_id, length(evimi.s))
) 

# Smoothing
evimi.s_sps$smoothedSpawners <- genSmooth(
	abund = evimi.s_sps$spawners,
	years = evimi.s_sps$year,
	genLength = genLength$gen_length[genLength$region == "East Vancouver Island & Mainland Inlets" & genLength$species == "Chum"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, evimi.s_sps)

# Generate standard figure
plot_abund(evimi.s_sps)

```

```{r east VI chum figs, results = 'asis', fig.width = 7, fig.height = 6, fig.cap = "**Above:** (a) Time series of spawner abundances showing raw sum of counts for indicator streams (red) and expanded estimates of spawner abudnance to indicator and non-indicator streams (black). This expanded time series is what is used to assess the current state and trends in spawner abundance for the region and species. (b) The values of Expansion Factor 1 that accounts for unmonitored indicator streams in a given year, and is calculated based on the proportional contribution of the unmonitored indicator stream(s) to the sum of all indicator streams in a (reference) decade."}

added_series.s <- sps_data %>% 
	filter(region == "East Vancouver Island & Mainland Inlets", species == "Chum") %>%
	select(year, spawners) %>%
	rename("value" = "spawners")

plot_expansions(speciesname = "Chum",
								regionname = "East Vancouver Island & Mainland Inlets",
								region_spawners = region_spawners,
								expansion_factors = expansion_factors,
								added_series = added_series.s)
	abline(v = 1952.5)
	text(1953, 5, "Data prior to 1953 were discarded", adj = 0)
```

#### Coho {.unnumbered}
Initial time series of spawner abundances for coho salmon in East Vancouver Island & Mainland Inlets were expanded from available spawner surveys, as described below in [Expansion Factors](#expansion-factors). For the assessments, we truncated the time series of spawner abundance to years 1953+ because earlier years had high expansion factor values indicating lack of monitoring on key indicator streams (see figures below). 

```{r east VI coho, results = 'asis', fig.width = 7, fig.height = 4, fig.cap="**Above:** Spawner abundance of East Vancouver Island & Mainland Inlets coho salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}
## Process data
evimi.s <- sp[2, "Coho", ]

# Get rid of NAs (esp in 2024)
evimi.s <- evimi.s[1:max(which(!is.na(evimi.s)))]

# For all VIMI species, start in 1953 - earlier expansion factors are large
evimi.s	<- evimi.s[which(as.numeric(names(evimi.s)) >= 1953)]

# Reformat data for SPS
evimi.s_sps <- data.frame(
	region = rep("East Vancouver Island & Mainland Inlets", length(evimi.s)),
	species = rep("Coho", length(evimi.s)),
	year = as.numeric(names(evimi.s)),
	spawners = as.numeric(evimi.s), 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(nuseds_id, length(evimi.s))
) 

# Smoothing
evimi.s_sps$smoothedSpawners <- genSmooth(
	abund = evimi.s_sps$spawners,
	years = evimi.s_sps$year,
	genLength = genLength$gen_length[genLength$region == "East Vancouver Island & Mainland Inlets" & genLength$species == "Coho"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, evimi.s_sps)

# Generate standard figure
plot_abund(evimi.s_sps)

```

```{r east VI coho figs, results = 'asis', fig.width = 7, fig.height = 6, fig.cap = "**Above:** (a) Time series of spawner abundances showing raw sum of counts for indicator streams (red) and expanded estimates of spawner abudnance to indicator and non-indicator streams (black). This expanded time series is what is used to assess the current state and trends in spawner abundance for the region and species. (b) The values of Expansion Factor 1 that accounts for unmonitored indicator streams in a given year, and is calculated based on the proportional contribution of the unmonitored indicator stream(s) to the sum of all indicator streams in a (reference) decade."}

added_series.s <- sps_data %>% 
	filter(region == "East Vancouver Island & Mainland Inlets", species == "Coho") %>%
	select(year, spawners) %>%
	rename("value" = "spawners")

plot_expansions(speciesname = "Coho",
								regionname = "East Vancouver Island & Mainland Inlets",
								region_spawners = region_spawners,
								expansion_factors = expansion_factors,
								added_series = added_series.s)
	abline(v = 1952.5)
	text(1953, 5, "Data prior to 1953 were discarded", adj = 0)
```

#### Pink {.unnumbered}
Initial time series of spawner abundances for pink salmon in East Vancouver Island & Mainland Inlets were expanded from available spawner surveys, as described below in [Expansion Factors](#expansion-factors). For the assessments, we truncated the time series of spawner abundance to years 1953+ because earlier years had high expansion factor values indicating lack of monitoring on key indicator streams (see figures below). 

```{r east VI pink, results = 'asis', fig.width = 7, fig.height = 4, fig.cap="**Above:** Spawner abundance of East Vancouver Island & Mainland Inlets pink salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}
## Process data
evimi.s <- sp[2, "Pink", ]

# Get rid of NAs (esp in 2024)
evimi.s <- evimi.s[1:max(which(!is.na(evimi.s)))]

# For all VIMI species, start in 1953 - earlier expansion factors are large
evimi.s	<- evimi.s[which(as.numeric(names(evimi.s)) >= 1953)]

# Reformat data for SPS
evimi.s_sps <- data.frame(
	region = rep("East Vancouver Island & Mainland Inlets", length(evimi.s)),
	species = rep("Pink", length(evimi.s)),
	year = as.numeric(names(evimi.s)),
	spawners = as.numeric(evimi.s), 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(nuseds_id, length(evimi.s))
) 

# Smoothing
evimi.s_sps$smoothedSpawners <- genSmooth(
	abund = evimi.s_sps$spawners,
	years = evimi.s_sps$year,
	genLength = genLength$gen_length[genLength$region == "East Vancouver Island & Mainland Inlets" & genLength$species == "Pink"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, evimi.s_sps)

# Generate standard figure
plot_abund(evimi.s_sps)

```

```{r east VI pink figs, results = 'asis', fig.width = 7, fig.height = 6, fig.cap = "**Above:** (a) Time series of spawner abundances showing raw sum of counts for indicator streams (red) and expanded estimates of spawner abudnance to indicator and non-indicator streams (black). This expanded time series is what is used to assess the current state and trends in spawner abundance for the region and species. (b) The values of Expansion Factor 1 that accounts for unmonitored indicator streams in a given year, and is calculated based on the proportional contribution of the unmonitored indicator stream(s) to the sum of all indicator streams in a (reference) decade."}

added_series.s <- sps_data %>% 
	filter(region == "East Vancouver Island & Mainland Inlets", species == "Pink") %>%
	select(year, spawners) %>%
	rename("value" = "spawners")

plot_expansions(speciesname = "Pink",
								regionname = "East Vancouver Island & Mainland Inlets",
								region_spawners = region_spawners,
								expansion_factors = expansion_factors,
								added_series = added_series.s)
	abline(v = 1952.5)
	text(1953, 5, "Data prior to 1953 were discarded", adj = 0)
```

#### Sockeye {.unnumbered}
Initial time series of spawner abundances for sockeye salmon in East Vancouver Island & Mainland Inlets were expanded from available spawner surveys, as described below in [Expansion Factors](#expansion-factors). For the assessments, we truncated the time series of spawner abundance to years 1953+ because earlier years had high expansion factor values indicating lack of monitoring on key indicator streams (see figures below). 

```{r east VI sockeye, results = 'asis', fig.width = 7, fig.height = 4, fig.cap="**Above:** Spawner abundance of East Vancouver Island & Mainland Inlets sockeye salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}
## Process data
evimi.s <- sp[2, "Sockeye", ]

# Get rid of NAs (esp in 2024)
evimi.s <- evimi.s[1:max(which(!is.na(evimi.s)))]

# For all VIMI species, start in 1953 - earlier expansion factors are large
evimi.s	<- evimi.s[which(as.numeric(names(evimi.s)) >= 1953)]

# Reformat data for SPS
evimi.s_sps <- data.frame(
	region = rep("East Vancouver Island & Mainland Inlets", length(evimi.s)),
	species = rep("Sockeye", length(evimi.s)),
	year = as.numeric(names(evimi.s)),
	spawners = as.numeric(evimi.s), 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(nuseds_id, length(evimi.s))
) 

# Smoothing
evimi.s_sps$smoothedSpawners <- genSmooth(
	abund = evimi.s_sps$spawners,
	years = evimi.s_sps$year,
	genLength = genLength$gen_length[genLength$region == "East Vancouver Island & Mainland Inlets" & genLength$species == "Sockeye"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, evimi.s_sps)

# Generate standard figure
plot_abund(evimi.s_sps)

```

```{r east VI sockeye figs, results = 'asis', fig.width = 7, fig.height = 6, fig.cap = "**Above:** (a) Time series of spawner abundances showing raw sum of counts for indicator streams (red) and expanded estimates of spawner abudnance to indicator and non-indicator streams (black). This expanded time series is what is used to assess the current state and trends in spawner abundance for the region and species. (b) The values of Expansion Factor 1 that accounts for unmonitored indicator streams in a given year, and is calculated based on the proportional contribution of the unmonitored indicator stream(s) to the sum of all indicator streams in a (reference) decade."}

added_series.s <- sps_data %>% 
	filter(region == "East Vancouver Island & Mainland Inlets", species == "Sockeye") %>%
	select(year, spawners) %>%
	rename("value" = "spawners")

plot_expansions(speciesname = "Sockeye",
								regionname = "East Vancouver Island & Mainland Inlets",
								region_spawners = region_spawners,
								expansion_factors = expansion_factors,
								added_series = added_series.s)
	abline(v = 1952.5)
	text(1953, 5, "Data prior to 1953 were discarded", adj = 0)
```


#### Steelhead {.unnumbered}
Initial time series of spawner abundances for steelhead salmon in East Vancouver Island & Mainland Inlets were expanded from available spawner surveys, as described below in [Expansion Factors](#expansion-factors). For the assessments, we truncated the time series of spawner abundance to years 1953+ because earlier years had high expansion factor values indicating lack of monitoring on key indicator streams (see figures below). 

```{r east VI steelhead, results = 'asis', fig.width = 7, fig.height = 4, fig.cap="**Above:** Spawner abundance of East Vancouver Island & Mainland Inlets steelhead salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}
## Process data
evimi.s <- sp[2, "Steelhead", ]

# Get rid of NAs (esp in 2024)
evimi.s <- evimi.s[1:max(which(!is.na(evimi.s)))]

# For all VIMI species, start in 1953 - earlier expansion factors are large
evimi.s	<- evimi.s[which(as.numeric(names(evimi.s)) >= 1953)]

# Reformat data for SPS
evimi.s_sps <- data.frame(
	region = rep("East Vancouver Island & Mainland Inlets", length(evimi.s)),
	species = rep("Steelhead", length(evimi.s)),
	year = as.numeric(names(evimi.s)),
	spawners = as.numeric(evimi.s), 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(nuseds_id, length(evimi.s))
) 

# Smoothing
evimi.s_sps$smoothedSpawners <- genSmooth(
	abund = evimi.s_sps$spawners,
	years = evimi.s_sps$year,
	genLength = genLength$gen_length[genLength$region == "East Vancouver Island & Mainland Inlets" & genLength$species == "Steelhead"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, evimi.s_sps)

# Generate standard figure
plot_abund(evimi.s_sps)

```

```{r east VI steelhead figs, results = 'asis', fig.width = 7, fig.height = 6, fig.cap = "**Above:** (a) Time series of spawner abundances showing raw sum of counts for indicator streams (red) and expanded estimates of spawner abudnance to indicator and non-indicator streams (black). This expanded time series is what is used to assess the current state and trends in spawner abundance for the region and species. (b) The values of Expansion Factor 1 that accounts for unmonitored indicator streams in a given year, and is calculated based on the proportional contribution of the unmonitored indicator stream(s) to the sum of all indicator streams in a (reference) decade."}

added_series.s <- sps_data %>% 
	filter(region == "East Vancouver Island & Mainland Inlets", species == "Steelhead") %>%
	select(year, spawners) %>%
	rename("value" = "spawners")

plot_expansions(speciesname = "Steelhead",
								regionname = "East Vancouver Island & Mainland Inlets",
								region_spawners = region_spawners,
								expansion_factors = expansion_factors,
								added_series = added_series.s)
	abline(v = 1952.5)
	text(1953, 5, "Data prior to 1953 were discarded", adj = 0)
```


### West VI

In the 2025 State of Salmon report, we have split the former "Vancouver Island & Mainland Inlets" region into two: East Vancouver Island & Mainland Inlets and West Vancouver Island. This split was made to better reflect the different ecology and habitats of salmon between the regions that may lead to different status and trends. 

#### Chinook {.unnumbered}
For Chinook spawner abundance, we used a 17-stream index of escapements from DFO. Streams were selected to represent WVI Chinook populations with minimal hatchery influence and infilling was done using data from neighbouring streams. We compare this below to the Chinook Technical Committee 14-stream index [@ChinookTechnicalCommittee2025AnnualReportCatch] used in last year's report and find them similar [@ChinookTechnicalCommittee2025AnnualReportCatch].

Chinook total abundance was reported by DFO in 2025 [@DFO2025WestCoastVancouver]. The data shown in Figure 3 of @DFO2025WestCoastVancouver are available for download at https://zenodo.org/records/14847020 [accessed 25-Aug-2025]. We assumed that total abundance is the sum of the (non-terminal) ocean catch, hatchery-origin terminal return, and natural-origin terminal return. We note that this estimate of total abundance represents a total count of all Chinook from the region, and thus cannot be compared to the index of spawner abundance, which represents only a proportion of those fish. 


```{r west VI chinook, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of West Vancouver Island chinook salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}
## Load data ---
# 17 stream index from Nicholas Brown @ DFO (pers comm.)
wvi_ck <- read.csv(here("data/R-OUT_infilled_indicators_escapement_timeseries.csv")) %>% 
	group_by(year) %>% summarize(tot_esc = sum(escapement))

# 14 Stream index from CTC
wvi_ck_ctc <- readxl::read_xlsx(here("data/TCCHINOOK-25-02-Appendix-B-Escapement-Detailed.xlsx"), sheet = "B5", range = "K24:K53", col_names = "WCVI_14Stream_Index") %>%
	mutate(Year = c(1995:2024)) # Escapement methods changed in 1995; only use earlier data for assessments

# Compare the CTC and DFO indices
wvi_ck %>% ggplot(aes(x=year, y=tot_esc/1000)) + geom_line(aes(col="DFO 17-Stream Index")) +
	geom_line(data=wvi_ck_ctc, aes(x=Year, y=WCVI_14Stream_Index/1000, col="CTC 14-Stream Index")) +
	scale_colour_manual(values = c("DFO 17-Stream Index" = "blue", "CTC 14-Stream Index" = "red")) +
	labs(col="Source", y="Spawner escapement (000s)", x="Year")
	
# Total abundance from Brown et al. 2025
wvi_ck_tot <- read_xlsx(here("data/WCVI_Chinook_total_abundance.xlsx"))

wvi_ck_tot$total_abundance <-  wvi_ck_tot$`Hatchery origin Terminal Return` + wvi_ck_tot$`Natural origin Terminal Return` +  wvi_ck_tot$`non-terminal Ocean catch`

yrs <- sort(unique(c(wvi_ck_tot$`Return year`, wvi_ck$Year)))

# Put in SPS format
wvick_sps <- data.frame(
	region = rep("West Vancouver Island", length(yrs)),
	species = rep("Chinook", length(yrs)),
	year = yrs,
	spawners = wvi_ck$tot_esc[match(yrs, wvi_ck$year)], 
	smoothedSpawners = NA,
	runsize = wvi_ck_tot$total_abundance[match(yrs, wvi_ck_tot$`Return year`)],
	smoothedRunsize = NA,
	source_id = rep(paste("Brown_20260421", "Brown_20250703", sep = ", "), length(yrs)) # update source ID 
) 

# Smoothing
wvick_sps$smoothedSpawners <- genSmooth(
	abund = wvick_sps$spawners,
	years = wvick_sps$year,
	genLength = genLength$gen_length[genLength$region == "West Vancouver Island" & genLength$species == "Chinook"]
)

wvick_sps$smoothedRunsize <- genSmooth(
	abund = wvick_sps$runsize,
	years = wvick_sps$year,
	genLength = genLength$gen_length[genLength$region == "West Vancouver Island" & genLength$species == "Chinook"]
)


if(1 == 2){ # Visualize the time peries put on the same scale
	avg <- c(spawners = exp(mean(log(wvick_sps$spawners), na.rm = TRUE)),
	total = exp(mean(log(wvick_sps$runsize), na.rm = TRUE)))

	plot(wvick_sps$year, (wvick_sps$smoothedRunsize - avg[2])/avg[2], "o", pch = 19)
	abline(h= avg[2])
	points(wvick_sps$year, (wvick_sps$smoothedSpawners - avg[1])/avg[1], "o", col = grey(0.6))

}
# Add to master sps dataframe
sps_data <- rbind(sps_data, wvick_sps)

# Generate standard plot
plot_abund(wvick_sps)

```


#### Sockeye {.unnumbered}
Reconstructions of sockeye spawners and recruitment are compelted by DFO (Nicholas Brown, pers. comm.) for three Conservation Units on West Vancouver Island: Sproat, Great Central, and Hucuktlis (formerly Henderson). We used the sum of spawners and total return for these three Conservation Units as the index of sockeye spawner and total abundance for WVI. We acknowledge that this index does not account for lake-type sockeye in other watersheds (e.g., Muchalat, Kennedy, Clayoquot) or river-type sockeye in West Vancouver Island and Northwest Vancouver Island CUs, which tend to be in much lower abundance.

```{r west VI sockeye, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of West Vancouver Island sockeye salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}

# Sum of CU abundance from Sproat, Great Central, and Henderson
## Load data ---
wvi_se_raw <- readxl::read_xlsx(here("data/Barkley_Sockeye_stock-recruit_infilled.xlsx"), sheet = "S-R data")

wvi_se_summed <- wvi_se_raw %>%
	dplyr::select(year, stock, S, N) %>% # S = escapement, N = annual terminal run size
	filter(year >= 1977) %>% # Remove 1972 to 2076 for HUC because not available for other stocks
	group_by(year) %>%
	summarise(totalS = sum(S),
						totalN = sum(N))

# plot(wvi_se_summed$year, wvi_se_summed$totalN, "o", ylim = c(0, 2e6), main = "WVI Sockeye", xlab = "", ylab = "")
# points(wvi_se_summed$year, wvi_se_summed$totalS, "o",  col = 2)
# points(yrs, sp[2, "Sockeye", ], "o", col = 4)
# legend("topleft", col = c(1,2,4), c("3 CU - Total run", "3 CU - Escape.", "NuSEDS expanded"), pch = 1, lty = 1)

# Reformat data for SPS
wvise_sps <- data.frame(
	region = rep("West Vancouver Island", dim(wvi_se_summed)[1]),
	species = rep("Sockeye", dim(wvi_se_summed)[1]),
	year = wvi_se_summed$year,
	spawners = wvi_se_summed$totalS, 
	smoothedSpawners = NA,
	runsize = wvi_se_summed$totalN,
	smoothedRunsize = NA,
	source_id = rep("Brown_20260420", dim(wvi_se_summed)[1])
) 

# Smoothing
wvise_sps$smoothedSpawners <- genSmooth(
	abund = wvise_sps$spawners,
	years = wvise_sps$year,
	genLength = genLength$gen_length[genLength$region == "West Vancouver Island" & genLength$species == "Sockeye"]
)

wvise_sps$smoothedRunsize <- genSmooth(
	abund = wvise_sps$runsize,
	years = wvise_sps$year,
	genLength = genLength$gen_length[genLength$region == "West Vancouver Island" & genLength$species == "Sockeye"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, wvise_sps)

# Generate standard figure
plot_abund(wvise_sps)
```


#### Chum {.unnumbered}
For West Vancouver Island chum salmon, spawner abundance was expanded from available spawner surveys, as described below in [Expansion Factors](#expansion-factors). The latest year of data (2025) were extracted from West Coast of Vancouver Island Stock Assessment bulletin dated November 15, 2024, and should be considered preliminary. We truncated the time series of spawner abundance to years 1953+ because earlier years had high expansion factor values indicating lack of monitoring on key indicator streams (see figures below). 

```{r west VI expansion data}

region_spawners <- sp <- readRDS(here("output/expanded-spawners/West Vancouver Island-spawners.rds"))
yrs <- as.numeric(dimnames(sp)[[3]])
expansion_factors <- readRDS(here("output/expanded-spawners/West Vancouver Island-expansion-factors.rds"))

```

```{r west VI chum, fig.width = 7, fig.height = 6, fig.cap="**Above:**Spawner abundance of West Vancouver Island chum salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}
## Process data ---
wvi.s <- sp[2, "Chum", ]
# Get rid of NAs (esp in 2024)
wvi.s <- wvi.s[1:max(which(!is.na(wvi.s)))]

# For all VIMI species, start in 1953 - earlier expansion factors are large
wvi.s	<- wvi.s[which(as.numeric(names(wvi.s)) >= 1953)]

# Reformat data for SPS
wvi.s_sps <- data.frame(
	region = rep("West Vancouver Island", length(wvi.s)),
	species = rep("Chum", length(wvi.s)),
	year = as.numeric(names(wvi.s)),
	spawners = as.numeric(wvi.s), 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(nuseds_id, length(wvi.s))
) 

# Smoothing
wvi.s_sps$smoothedSpawners <- genSmooth(
	abund = wvi.s_sps$spawners,
	years = wvi.s_sps$year,
	genLength = genLength$gen_length[genLength$region == "West Vancouver Island" & genLength$species == "Chum"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, wvi.s_sps)

# Generate standard figure
plot_abund(wvi.s_sps) 
```

```{r west VI chum figs, results = 'asis', fig.width = 7, fig.height = 6, fig.cap = "**Above:** (a) Time series of spawner abundances showing raw sum of counts for indicator streams (red) and expanded estimates of spawner abudnance to indicator and non-indicator streams (black). This expanded time series is what is used to assess the current state and trends in spawner abundance for the region and species. (b) The values of Expansion Factor 1 that accounts for unmonitored indicator streams in a given year, and is calculated based on the proportional contribution of the unmonitored indicator stream(s) to the sum of all indicator streams in a (reference) decade."}

# added_series.s <- sps_data %>%
# 	filter(region == "West Vancouver Island", species == "Chum") %>%
# 	select(year, spawners) %>%
# 	rename("value" = "spawners")
	
plot_expansions(speciesname = "Chum",
								regionname = "West Vancouver Island",
								region_spawners = region_spawners,
								expansion_factors = expansion_factors)#,
								# added_series = added_series.s)
	abline(v = 1952.5)
	text(1953, par('usr')[4]/3, "Data prior to 1953 were discarded", adj = 0)

#rm(added_series.s)
```


#### Coho {.unnumbered}
For West Vancouver Island coho salmon, spawner abundance was expanded from available spawner surveys, as described below in [Expansion Factors](#expansion-factors). The latest year of data (2025) were extracted from West Coast of Vancouver Island Stock Assessment bulletin dated November 15, 2024, and should be considered preliminary. We truncated the time series of spawner abundance to years 1953+ because earlier years had high expansion factor values indicating lack of monitoring on key indicator streams (see figures below). 
```{r west VI coho, fig.width = 7, fig.height = 6, fig.cap="**Above:**Spawner abundance of West Vancouver Island coho salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}
## Process data ---
wvi.s <- sp[2, "Coho", ]
# Get rid of NAs (esp in 2024)
wvi.s <- wvi.s[1:max(which(!is.na(wvi.s)))]

# For all VIMI species, start in 1953 - earlier expansion factors are large
wvi.s	<- wvi.s[which(as.numeric(names(wvi.s)) >= 1953)]

# Reformat data for SPS
wvi.s_sps <- data.frame(
	region = rep("West Vancouver Island", length(wvi.s)),
	species = rep("Coho", length(wvi.s)),
	year = as.numeric(names(wvi.s)),
	spawners = as.numeric(wvi.s), 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(nuseds_id, length(wvi.s))
) 

# Smoothing
wvi.s_sps$smoothedSpawners <- genSmooth(
	abund = wvi.s_sps$spawners,
	years = wvi.s_sps$year,
	genLength = genLength$gen_length[genLength$region == "West Vancouver Island" & genLength$species == "Coho"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, wvi.s_sps)

# Generate standard figure
plot_abund(wvi.s_sps) 
```

```{r west VI coho figs, results = 'asis', fig.width = 7, fig.height = 6, fig.cap = "**Above:** (a) Time series of spawner abundances showing raw sum of counts for indicator streams (red) and expanded estimates of spawner abudnance to indicator and non-indicator streams (black). This expanded time series is what is used to assess the current state and trends in spawner abundance for the region and species. (b) The values of Expansion Factor 1 that accounts for unmonitored indicator streams in a given year, and is calculated based on the proportional contribution of the unmonitored indicator stream(s) to the sum of all indicator streams in a (reference) decade."}

# added_series.s <- sps_data %>%
# 	filter(region == "West Vancouver Island", species == "Coho") %>%
# 	select(year, spawners) %>%
# 	rename("value" = "spawners")
	
plot_expansions(speciesname = "Coho",
								regionname = "West Vancouver Island",
								region_spawners = region_spawners,
								expansion_factors = expansion_factors)#,
								# added_series = added_series.s)
	abline(v = 1952.5)
	text(1953, par('usr')[4]/3, "Data prior to 1953 were discarded", adj = 0)

#rm(added_series.s)
```

#### Pink {.unnumbered}

For West Vancouver Island pink salmon, spawner abundance was expanded from available spawner surveys, as described below in [Expansion Factors](#expansion-factors). The latest year of data (2025) were extracted from West Coast of Vancouver Island Stock Assessment bulletin dated November 15, 2024, and should be considered preliminary. We truncated the time series of spawner abundance to years 1953+ because earlier years had high expansion factor values indicating lack of monitoring on key indicator streams (see figures below). 

For chum, coho, and steelhead, we relied on NuSEDS' designations of indicator or non-indicator streams, but for pink salmon in this region we designated indicator streams as those with at least 10 years of data in the most recent 20 years. As a result, expansion factors were higher in historical years, but data from both indicator and non-indicator streams show markedly higher counts in the period 1960-1980 then in subsequent decades and it is unlikely that the observed decline is a result of uncertain expansions. 

```{r west VI pink, fig.width = 7, fig.height = 6, fig.cap="**Above:**Spawner abundance of West Vancouver Island pink salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}
## Process data ---
wvi.s <- sp[2, "Pink", ]
# Get rid of NAs 
wvi.s <- wvi.s[1:max(which(!is.na(wvi.s)))]

# For all VIMI species, start in 1953 - earlier expansion factors are large
wvi.s	<- wvi.s[which(as.numeric(names(wvi.s)) >= 1953)]

# Reformat data for SPS
wvi.s_sps <- data.frame(
	region = rep("West Vancouver Island", length(wvi.s)),
	species = rep("Pink", length(wvi.s)),
	year = as.numeric(names(wvi.s)),
	spawners = as.numeric(wvi.s), 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(nuseds_id, length(wvi.s))
) 

# Smoothing
wvi.s_sps$smoothedSpawners <- genSmooth(
	abund = wvi.s_sps$spawners,
	years = wvi.s_sps$year,
	genLength = genLength$gen_length[genLength$region == "West Vancouver Island" & genLength$species == "Pink"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, wvi.s_sps)

# Generate standard figure
plot_abund(wvi.s_sps) 
```

```{r west VI pink figs, results = 'asis', fig.width = 7, fig.height = 6, fig.cap = "**Above:** (a) Time series of spawner abundances showing raw sum of counts for indicator streams (red) and expanded estimates of spawner abudnance to indicator and non-indicator streams (black). This expanded time series is what is used to assess the current state and trends in spawner abundance for the region and species. (b) The values of Expansion Factor 1 that accounts for unmonitored indicator streams in a given year, and is calculated based on the proportional contribution of the unmonitored indicator stream(s) to the sum of all indicator streams in a (reference) decade."}

# added_series.s <- sps_data %>%
# 	filter(region == "West Vancouver Island", species == "Pink") %>%
# 	select(year, spawners) %>%
# 	rename("value" = "spawners")
	
plot_expansions(speciesname = "Pink",
								regionname = "West Vancouver Island",
								region_spawners = region_spawners,
								expansion_factors = expansion_factors)#,
								# added_series = added_series.s)
	abline(v = 1952.5)
	text(1953, par('usr')[4]/3, "Data prior to 1953 were discarded", adj = 0)

#rm(added_series.s)
```

#### Steelhead {.unnumbered}
For West Vancouver Island steelhead, spawner abundance was expanded from available spawner surveys, as described below in [Expansion Factors](#expansion-factors). The latest year of data (2025) were extracted from West Coast of Vancouver Island Stock Assessment bulletin dated November 15, 2024, and should be considered preliminary. We truncated the time series of spawner abundance to years 1953+ because earlier years had high expansion factor values indicating lack of monitoring on key indicator streams (see figures below). 

```{r west VI steelhead, fig.width = 7, fig.height = 6, fig.cap="**Above:**Spawner abundance of West Vancouver Island steelhead salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}
## Process data ---
wvi.s <- sp[2, "Steelhead", ]
# Get rid of NAs (esp in 2024)
wvi.s <- wvi.s[1:max(which(!is.na(wvi.s)))]

# For all VIMI species, start in 1953 - earlier expansion factors are large
wvi.s	<- wvi.s[which(as.numeric(names(wvi.s)) >= 1953)]

# Reformat data for SPS
wvi.s_sps <- data.frame(
	region = rep("West Vancouver Island", length(wvi.s)),
	species = rep("Steelhead", length(wvi.s)),
	year = as.numeric(names(wvi.s)),
	spawners = as.numeric(wvi.s), 
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep(nuseds_id, length(wvi.s))
) 

# Smoothing
wvi.s_sps$smoothedSpawners <- genSmooth(
	abund = wvi.s_sps$spawners,
	years = wvi.s_sps$year,
	genLength = genLength$gen_length[genLength$region == "West Vancouver Island" & genLength$species == "Steelhead"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, wvi.s_sps)

# Generate standard figure
plot_abund(wvi.s_sps) 
```

```{r west VI steelhead figs, results = 'asis', fig.width = 7, fig.height = 6, fig.cap = "**Above:** (a) Time series of spawner abundances showing raw sum of counts for indicator streams (red) and expanded estimates of spawner abudnance to indicator and non-indicator streams (black). This expanded time series is what is used to assess the current state and trends in spawner abundance for the region and species. (b) The values of Expansion Factor 1 that accounts for unmonitored indicator streams in a given year, and is calculated based on the proportional contribution of the unmonitored indicator stream(s) to the sum of all indicator streams in a (reference) decade."}

# added_series.s <- sps_data %>%
# 	filter(region == "West Vancouver Island", species == "Steelhead") %>%
# 	select(year, spawners) %>%
# 	rename("value" = "spawners")
	
plot_expansions(speciesname = "Steelhead",
								regionname = "West Vancouver Island",
								region_spawners = region_spawners,
								expansion_factors = expansion_factors)#,
								# added_series = added_series.s)
	abline(v = 1952.5)
	text(1953, par('usr')[4]/3, "Data prior to 1953 were discarded", adj = 0)

#rm(added_series.s)
```

### Fraser

#### Chinook {.unnumbered}
<!-- River-level spawner abundance for Fraser Chinook found in NuSEDS can be highly unreliable, thus we based our expansions on only the most intensively monitored streams that have been deemed as reliable estimates of abundance by @BrownEtAl20202020SummaryAbundance. These streams are those shown in the [Pacific Salmon Explorer](https://www.salmonexplorer.ca), and the associated data can be accessed from the stream survey data in the [SWP Data Library](https://data.salmonwatersheds.ca/result?datasetid=1). We also truncated the Chinook index of abundance to 1984 and more recent data, as earlier years have been deemed unreliable [@BrownEtAl20202020SummaryAbundance] and from 1984 onward spawner estimates mirrored those reported by @AtlasEtAl2023TrendsChinookSalmon. For Chinook total return, we used the sum of total run for four Fraser populations (Harrison, Nicola, Lower Shuswap, and South Thompson - all) reported in @AtlasEtAl2023TrendsChinookSalmon and available from the associated [GitHub repository](https://github.com/willatlas/ChinookEscapement-RunSize). -->

Escapement of Fraser Chinook is provided in the [PSC's Chinook Technical Committee (CTC) data sets](https://www.psc.org/publications/technical-reports/technical-committee-reports/chinook/ctc-data-sets/) and are reported on in @ChinookTechnicalCommittee2025AnnualReportCatch. We took the sum of escapement to five major stocks from Table B6 (Fraser Spring 4-2 (age 1.2), Fraser Spring 5-2 (age 1.3), Fraser Summer 5-2 (age 1.3), Fraser Summer 4-1 (age 0.3), and Harrison (Fraser Fall 4-1, age 0.3)) as an index of spawner abundance. Total run size is also reported in Table B6 for the Spring/Summer runs, and we use this as an index of total abundance for Fraser Chinook, noting that it excludes the Fall 4-1 stock.

```{r fraser chinook, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of Fraser chinook salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}
# Data from Chuck Parken (shared 23-Oct-2025 via email)
frck_cp <- readxl::read_xlsx(here("data/1975-2025 Fraser Chinook escapement and terminal run by SMU - for PSF.xlsx"), skip=1)
frck_smu <- readxl::read_xlsx(here("data/1975-2025 Fraser Chinook escapement and terminal run by SMU - for PSF.xlsx"), range="A1:K1", col_names=F) # get smu names
frck_smu <- rep(stringr::str_remove_all(frck_smu, "Fraser|age|-run|[[:punct:]]|[[:space:]]")[!is.na(frck_smu)], each=2)
names(frck_cp)[2:ncol(frck_cp)] <- paste(frck_smu, stringr::str_extract_all(names(frck_cp)[2:ncol(frck_cp)], "Esc|Run"), sep="_") # stitch together smu and col names
names(frck_cp) <- stringr::str_to_lower(names(frck_cp))

# Cut data to only 1984 onward when Harrison data were available.
frck_cp <- frck_cp %>% dplyr::filter(year >= 1984)

# Numbers match CTC except that t.run from CTC doesn't include Harrison fall. So need to add that in

# Sum across SMUs
frck_cp_esc <- round(apply(frck_cp[, grep("esc", names(frck_cp))], 1, sum))
frck_cp_run <- round(apply(frck_cp[, grep("run", names(frck_cp))], 1, sum))

yrs <- frck_cp$year

# Put in SPS format
frck_sps <- data.frame(
	region = rep("Fraser", length(yrs)),
	species = rep("Chinook", length(yrs)),
	year = yrs,
	spawners = frck_cp_esc,
	smoothedSpawners = NA,
	runsize = frck_cp_run,
	smoothedRunsize = NA,
	source_id = rep("Parken_20251023", length(yrs))
) 
	
	# Smoothing
frck_sps$smoothedSpawners <- genSmooth(
		abund = frck_sps$spawners,
		years = frck_sps$year,
		genLength = genLength$gen_length[genLength$region == "Fraser" & genLength$species == "Chinook"]
	)

frck_sps$smoothedRunsize <- genSmooth(
	abund = frck_sps$runsize,
	years = frck_sps$year,
	genLength = genLength$gen_length[genLength$region == "Fraser" & genLength$species == "Chinook"]
)

#(tail(frck_sps$smoothedRunsize, 1) - exp(mean(log(frck_sps$runsize), na.rm = TRUE)))/exp(mean(log(frck_sps$runsize), na.rm = TRUE))

# Add to master sps dataframe
sps_data <- rbind(sps_data, frck_sps)

# Generate standard figure
plot_abund(frck_sps)

```


```{r fraser chinook figs, echo = FALSE, include = TRUE, results = 'asis', fig.width = 9, fig.height = 6, fig.cap = "**Above:** (a) Aggregate escapement of Fraser Chinook with the contribution of escapement to each of five stocks shown as coloured regions. Note that the Harrison River (Fall 4-1) time series begins later (1984) than the other four stocks (1975), and thus we only report spawner abundance from 1984 onward. (b) The proportional contribution of each of five stocks to aggregate Fraser Chinook escapement."}
suppressMessages(
ctc_fr <- readxl::read_xlsx(here("data/TCCHINOOK-25-02-Appendix-B-Escapement-Detailed.xlsx"), sheet = "B6", range = "A3:P53", col_types = "numeric", ) )
names(ctc_fr) <- c("Year", "Fraser Spring Age 1.2 Esc", "Fraser Spring Age 1.3 Esc", "Fraser Summer Age 0.3 Esc", "Fraser Summer Age 1.3 Esc", "Fraser Spring/Summer t.run", "Harrison Esc", "Harrison CV", "Lower Shuswap Esc", "Lower Shuswap CV", "Nicola Esc", "Nicola CV", "Lower Chilcotin Esc", "Lower Chilcotin CV", "Chilko Esc", "Chilko CV")
ctc_esc <- ctc_fr %>% dplyr::select(c("Year", "Fraser Spring Age 1.2 Esc", "Fraser Spring Age 1.3 Esc", "Fraser Summer Age 0.3 Esc", "Fraser Summer Age 1.3 Esc", "Harrison Esc"))
ctc_trun <- ctc_fr %>% dplyr::select(c(Year, ))
ctc_esc_sum <- apply(ctc_esc[, c("Fraser Spring Age 1.2 Esc", "Fraser Spring Age 1.3 Esc", "Fraser Summer Age 0.3 Esc", "Fraser Summer Age 1.3 Esc", "Harrison Esc")], 1, sum)

ctc_trun_sum <- ctc_fr$`Fraser Spring/Summer t.run`

col_frck <- viridisLite::viridis(n = 5)

# Plot ---

# Spawner abundance
esc.stacked <- ctc_esc %>% 
	rename('Spring 4-2' = 'Fraser Spring Age 1.2 Esc',
				 'Spring 5-2' = 'Fraser Spring Age 1.3 Esc',
				 'Summer 5-2' = 'Fraser Summer Age 1.3 Esc',
				 'Summer 4-1' = 'Fraser Summer Age 0.3 Esc', 
				 'Fall 4-1 (Harrison)' = 'Harrison Esc') %>% 
	tidyr::pivot_longer(-1, names_to="Stock", values_to="Escapement") %>% 
	ggplot() +
	geom_area(aes(x=Year, y=Escapement/1000, fill=factor(Stock, levels=rev(unique(Stock)))), 
								alpha=0.6) + 
	scale_fill_viridis_d() +
	labs(y="Spawner abundance (thousands)", fill="") +
	theme_minimal() + theme(legend.position = "none")




esc.sum <- ctc_esc %>% 
	rename('Spring 4-2' = 'Fraser Spring Age 1.2 Esc',
				 'Spring 5-2' = 'Fraser Spring Age 1.3 Esc',
				 'Summer 5-2' = 'Fraser Summer Age 1.3 Esc',
				 'Summer 4-1' = 'Fraser Summer Age 0.3 Esc', 
				 'Fall 4-1 (Harrison)' = 'Harrison Esc') %>% 
	tidyr::pivot_longer(-1, names_to="Stock", values_to="Escapement") %>% 
	filter(Year >= 1984) %>% 
	ggplot() +
	geom_area(aes(x=Year, y=Escapement/1000, fill=factor(Stock, levels=rev(unique(Stock)))), 
								alpha=0.6, position="fill") + 
	scale_fill_viridis_d() +
	labs(y="Proportion of spawner abundance (thousands)", fill="") +
	theme_minimal() + theme(legend.position = "right")

suppressWarnings(
cowplot::plot_grid(esc.stacked, esc.sum, labels="auto", rel_widths=c(0.75, 1))
)
```

#### Chum {.unnumbered}
Fraser chum catch and escapement are reported by PSC's Chum Technical Committee [@20232019PostSeasonSummary], and were provided to us on request to Brittany Jenewin (DFO, pers. comm.). For 2023 and 2024, estimates were extracted from the [Fraser Management Council post-season review](https://frasersalmon.ca/wp-content/uploads/2025/01/2024-Fraser-Coho-and-Chum-Post-season-Review-2025-01-28-FINAL-1.pdf).

```{r fraser chum, echo=FALSE, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of Fraser Chum salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}
# (2) Use TCCHUM Fraser data
suppressMessages(
frcm_esc <- readxl::read_xlsx(here("data/ChumTC Report Tables - full time series up to 2022.xlsx"), sheet = "3-11", range = "A4:B29") %>%
	rename(year = "...1", esc = "Fraser River") )

suppressMessages(
frcm_catch1 <- readxl::read_xlsx(here("data/ChumTC Report Tables - full time series up to 2022.xlsx"), sheet = "3-8", range = "A3:G30") %>%
	rename(year = "Year", catch = "Total") %>%
	dplyr::select(year, catch) )

suppressMessages(
frcm_catch2 <- readxl::read_xlsx(here("data/ChumTC Report Tables - full time series up to 2022.xlsx"), sheet = "3-9", range = "A4:G31") %>%
	rename(year = "...1", catch = "Total") %>%
	dplyr::select(year, catch) )


# Find years when catch and escapement are reported
yrs_tcchum <- c(max(c(min(frcm_esc$year), min(frcm_catch1$year), min(frcm_catch2$year))):min(c(max(frcm_esc$year), max(frcm_catch1$year), max(frcm_catch2$year)))) #1998:2022

frcm_run <- frcm_esc$esc[match(yrs_tcchum, frcm_esc$year)] + frcm_catch1$catch[match(yrs_tcchum, frcm_catch1$year)] + frcm_catch2$catch[match(yrs_tcchum, frcm_catch2$year)] 
	
# Data extracted from Fraser Management council post-season review presentation
# https://frasersalmon.ca/wp-content/uploads/2025/01/2024-Fraser-Coho-and-Chum-Post-season-Review-2025-01-28-FINAL-1.pdf
frcm_FMC <- read.csv(here("data/FraserChum_FMC.csv"))

# Data from PSC post-season (Brittany Jenewin)
frcm_PSC <- read.csv(here("data/FraserChum_PSC.csv"))

frcm <- frcm_esc %>%
	mutate(runsize = frcm_run) %>%
	bind_rows(data.frame(
		year = c(2023, 2024),
		esc = frcm_PSC$escapement[match(c(2023, 2024), frcm_PSC$year)],
		runsize = frcm_PSC$total_run[match(c(2023, 2024), frcm_PSC$year)]
	))


# Put in SPS format
frcm_sps <- data.frame(
	region = rep("Fraser", length(frcm$year)), 
	species = rep("Chum", length(frcm$year)), 
	year = frcm$year, 
	spawners = frcm$esc,  
	smoothedSpawners = NA,
	runsize = frcm$runsize,
	smoothedRunsize = NA,
	source_id = c(rep("Jenewin_20240416", nrow(frcm_esc)), rep("DFO_20250128", 2))
) 

# Smoothing
frcm_sps$smoothedSpawners <- genSmooth(
	abund = frcm_sps$spawners,
	years = frcm_sps$year,
	genLength = genLength$gen_length[genLength$region == "Fraser" & genLength$species == "Chum"]
)

frcm_sps$smoothedRunsize <- genSmooth(
	abund = frcm_sps$runsize,
	years = frcm_sps$year,
	genLength = genLength$gen_length[genLength$region == "Fraser" & genLength$species == "Chum"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, frcm_sps)

# Generate standard plot
plot_abund(frcm_sps)
```


#### Coho {.unnumbered}
For interior Fraser coho CUs, spawner abundance and total return to 2022 were provided by DFO on data request (pers. comm. Marissa Glavas, Data Manager, Fraser River Stock Assessment). We report total return as the sum of "Total Pre-Fishery Abundance" to all CUs and spawner abundance as the sum of "Total Return" (i.e., final spawner estimate plus fish removed from the system by DFO Salmon Enhancement Program or First Nations) to all CUs (we note the term "total return" is used differently in this data set to represent spawners!). We recognize that using these estimates ignores coastal coho populations, for which monitoring has been patchy through time. Spawner and run size for 2023 and preliminary spawner abundance for 2024 were taken from the presentation of [2025/26 Southern B.C. Salmon: Chum and Coho](https://frasersalmon.ca/wp-content/uploads/2025/02/2025-Forum-2_Richmond_Chum-and-Coho.pdf) dated February 2025. Preliminary 2024 spawner abundance is for natural-origin spawners only, and is a lower estimate of the total 2024 spawner abundance. However, the numbers of hatchery-origin spawners have been low (< 10\%) over the past two decades.

```{r fraser coho, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of Fraser coho salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}
## Load data ---
frco <- read.csv(here("data/IFC_data_compiled.csv"))
# Put in SPS format
frco_sps <- data.frame(
	region = rep("Fraser", length(frco$year)),
	species = rep("Coho", length(frco$year)),
	year = frco$year,
	spawners = frco$spawners, # Note this is the fish that returned to spawn, NOT a mistake!
	smoothedSpawners = NA,
	runsize = frco$runsize,
	smoothedRunsize = NA,
	source_id = c(rep("Glavas_20231006", length(frco$year) - 2), rep("DFO_20250225", 2))
) 

# Smoothing
frco_sps$smoothedSpawners <- genSmooth(
	abund = frco_sps$spawners,
	years = frco_sps$year,
	genLength = genLength$gen_length[genLength$region == "Fraser" & genLength$species == "Coho"]
)

frco_sps$smoothedRunsize <- genSmooth(
	abund = frco_sps$runsize,
	years = frco_sps$year,
	genLength = genLength$gen_length[genLength$region == "Fraser" & genLength$species == "Coho"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, frco_sps)

# Generate standard figure
plot_abund(frco_sps)

```


#### Pink and Sockeye {.unnumbered}
Data on total return and spawner abundance of pink salmon and sockeye salmon in the Fraser region are provided by the PSC and accessed through the Fraser Panel Annual Report: Data Application [@PacificSalmonCommission2024PSCAnnualReport]. Total escapement for Fraser sockeye was not yet available for 2025, and 2025 total return is considered preliminary. We note that Fraser River pink salmon are only counted in the dominant, odd-year run. 

```{r fraser pink, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of Fraser pink salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}
## Load data ---
# Note: Only odd year data, so no change in some years
# From PSC 
frpk <- read.csv(here("data/pink_run_size_2026-04-20.csv"))

# Put in SPS format
frpk_sps <- data.frame(
	region = rep("Fraser", length(frpk$Year)),
	species = rep("Pink", length(frpk$Year)),
	year = frpk$Year,
	spawners = frpk$Escapement, # Note this is the fish that returned to spawn, NOT a mistake!
	smoothedSpawners = NA,
	runsize = frpk$Run.Size,
	smoothedRunsize = NA, 
	source_id = rep("PSC_20260417", length(frpk$Year))
) 

# Smoothing - don't do this for pink salmon in the Fraser - no even year data!
frpk_sps$smoothedSpawners <- frpk_sps$spawners
frpk_sps$smoothedRunsize <- frpk_sps$runsize

# Add to master sps dataframe
sps_data <- rbind(sps_data, frpk_sps)

#Generate standard figure
plot_abund(frpk_sps)

```

```{r fraser sockeye, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of Fraser sockeye salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}
## Load data ---
# From PSC
frse <- read.csv(here("data/Fraser Sockeye Run Size_2026-04-20.csv")) %>%
	filter(Management.Group == "Total Fraser")

# Put in SPS format
frse_sps <- data.frame(
	region = rep("Fraser", length(frse$Year)),
	species = rep("Sockeye", length(frse$Year)),
	year = frse$Year,
	spawners = frse$Spawning.Escapement, # Note this is the fish that returned to spawn, NOT a mistake!
	smoothedSpawners = NA,
	runsize = frse$Run.Size,
	smoothedRunsize = NA,
	source_id = rep("PSC_20260417", length(frse$Year))
) 

# Fill in run size with in-season estimates if available
if(sum(is.na(frse_sps$runsize)) > 0){
	if(sum(!is.na(frse$In.season.Run.Size[which(frse$Year %in% frse_sps$year[is.na(frse_sps$runsize)])])) > 0){
		frse_sps$runsize[which(frse$Year %in% frse_sps$year[is.na(frse_sps$runsize)])] <- frse$In.season.Run.Size[which(frse$Year %in% frse_sps$year[is.na(frse_sps$runsize)])]
	}
}

# Smoothing
frse_sps$smoothedSpawners <- genSmooth(
	abund = frse_sps$spawners,
	years = frse_sps$year,
	genLength = genLength$gen_length[genLength$region == "Fraser" & genLength$species == "Sockeye"]
)

frse_sps$smoothedRunsize <- genSmooth(
	abund = frse_sps$runsize,
	years = frse_sps$year,
	genLength = genLength$gen_length[genLength$region == "Fraser" & genLength$species == "Sockeye"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, frse_sps)

# Generate standard figure
# plot_abund(frse_sps[frse$Year >= 1980,]); abline(v = 2024)
plot_abund(frse_sps)
```


#### Steelhead {.unnumbered}
Steelhead trout in the Fraser are monitored by the Province in at least 10 different streams, but these data are not readily available. Relatively reliable estimates of steelhead spawner abundance are available at the CU-level for interior Fraser steelhead from the Thompson Summer CU (monitored at the Thompson River) and Mid Fraser Summer CU (primarily monitored at the Chilcotin River; data in the [Pacific Salmon Explorer](https://salmonexplorer.ca)). We used the sum of CU-level spawner abundance for these two CUs as an index of Fraser steelhead abundance. We note that this approach does not include more coastal populations, such as the Lower Fraser Summer steelhead monitored in the Coquihalla River or Boundary Bay Winter steelhead, which may not have declined to the same extent over the past decade. However, a lack of publicly accessible data on coastal Fraser steelhead has limited our ability to include these CUs in our index of abundance.

```{r fraser steelhead, fig.cap="**Above:**Spawner abundance of Fraser steelhead salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}
## Load data ---
# Use sum of two main Fraser steelhead CUs
frsh <- read.csv(here("data/dataset1cu_steelhead_20250117.csv")) %>% 
	filter(region == "Fraser", species_name == "Steelhead") %>%
	filter(cu_name_pse %in% c("Thompson Summer", "Mid Fraser Summer")) %>%
	filter(!is.na(estimated_count) & estimated_count != -989898)

frsh_sum <- tapply(as.numeric(frsh$estimated_count), frsh$year, sum)
frsh_sum <- frsh_sum[min(which(!is.na(frsh_sum))): max(which(!is.na(frsh_sum)))]

# Put in SPS format
frsh_sps <- data.frame(
	region = rep("Fraser", length(frsh_sum)),
	species = rep("Steelhead", length(frsh_sum)),
	year = as.numeric(names(frsh_sum)),
	spawners = frsh_sum, # Note this is the fish that returned to spawn, NOT a mistake!
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep("Bison_20241129", length(frsh_sum))
) 

# Smoothing
frsh_sps$smoothedSpawners <- genSmooth(
	abund = frsh_sps$spawners,
	years = frsh_sps$year,
	genLength = genLength$gen_length[genLength$region == "Fraser" & genLength$species == "Steelhead"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, frsh_sps)

# Generate standard figure
plot_abund(frsh_sps)
```


### Columbia

#### Chinook {.unnumbered}

We used CU-level estimates of spawner abundance (run reconstructions) sourced from DFO (Athena Ogden, pers. comm.) for Chinook and sockeye.

```{r columbia chinook, fig.cap="**Above:**Spawner abundance of Columbia chinook salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}
colck <- read.csv(here("data/Chinook_Okanagan_NEW.csv")) %>% filter(Year >= 2006)
# Downloaded from https://github.com/SOLV-Code/Scanner-Data-Processing/blame/main/DATA_IN/Chinook_Okanagan_NEW.csv
# on March 18, 2025

# Put in SPS format
colck_sps <- data.frame(
	region = rep("Columbia", length(colck$Year)),
	species = rep("Chinook", length(colck$Year)),
	year = colck$Year,
	spawners = as.numeric(colck$NatOrigSpn + colck$HatchOrigSpn), # Note this is the fish that returned to spawn, NOT a mistake!
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep("DFO_20250318", length(colck$Year))
) 

# Smoothing
colck_sps$smoothedSpawners <- genSmooth(
	abund = colck_sps$spawners,
	years = colck_sps$year,
	genLength = genLength$gen_length[genLength$region == "Columbia" & genLength$species == "Chinook"]
)

# Add to master sps dataframe
sps_data <- rbind(sps_data, colck_sps)

# Generate standard figure
plot_abund(colck_sps)
```

#### Sockeye {.unnumbered}

We used CU-level estimates of spawner abundance (run reconstructions) sourced from DFO (Athena Ogden, pers. comm.) for Chinook and sockeye. Spawner abundance for Columbia sockeye was the sum of estimates for Osoyoos and, more recently, Skaha Lakes and were the most up-to-date estimates as of June 2025 (Athena Ogden, pers. comm.). Total abundance for sockeye was extracted from DFO's 2024 Wild Salmon Policy assessment for Osoyoos sockeye [@DFO2024WildSalmonPolicy] from 1980-2023.

```{r columbia sockeye, fig.cap="**Above:**Spawner abundance (grey) and total abundance (black) of Columbia sockeye salmon. For both, the raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average for each."}

# Load data ---
colse_old <- read.csv(here("data/spawner_abundance.csv")) %>% 
	subset(region == "Columbia" &  species_name == "Lake sockeye") %>%
	subset(!is.na(estimated_count))

colse <- read.csv(here("data/OkanaganSockeye.csv"))

colse2 <- readxl::read_xlsx(here("data/Okanagan Sockeye Esc for KR_revJn2025.xlsx"), sheet = "CAN Escapement", range = "A1:E65")

# Put in SPS format
yrs <- sort(colse2$Year) # don't use pre-2004 (not comparable; DFO 2024)
colse_sps <- data.frame(
	region = rep("Columbia", length(yrs)),
	species = rep("Sockeye", length(yrs)),
	year = yrs,
	spawners = round(colse2$`Total spawners`[match(yrs, colse2$Year)]), 
	smoothedSpawners = NA,
	runsize = colse$runsize[match(yrs, colse$year)],
	smoothedRunsize = NA, 
	source_id = c(rep("Ogden_20250627, Ogden_20250901", length(yrs) - 1), "Ogden_20250627")
) 

# Smoothing
colse_sps$smoothedSpawners <- genSmooth(
	abund = colse_sps$spawners,
	years = colse_sps$year,
	genLength = genLength$gen_length[genLength$region == "Columbia" & genLength$species == "Sockeye"]
)
colse_sps$smoothedRunsize <- genSmooth(
	abund = colse_sps$runsize,
	years = colse_sps$year,
	genLength = genLength$gen_length[genLength$region == "Columbia" & genLength$species == "Sockeye"]
)


# Add to master sps dataframe
sps_data <- rbind(sps_data, colse_sps)

# Generate standard figure
plot_abund(colse_sps)
```

#### Steelhead {.unnumbered}

There is no monitoring of steelhead trout in the Canadian portion of the Columbia region, but the Okanagan Nation Alliance does enumerate steelhead in akskwəkwant (Inkaneep Creek) and estimate a Canadian portion of steelhead spawning abundance. These data can be found in Table 8 of the associated report (e.g. @OBMEP20252024OkanoganSubbasin) and are available for the Mid Columbia Summer Conservation Unit [in the Pacific Salmon Explorer](https://salmonexplorer.ca/explore/data/spawner-abundance/columbia/steelhead/mid-columbia-summer-1380/).

```{r columbia steelhead, fig.cap="**Above:**Spawner abundance of Columbia steelhead salmon. The raw data is shown by points and the smoothed metric by solid lines. Dashed horizonal lines represent the long-term average."}
colsh <- read.csv(here("data/Columbia_Steelhead_OBMEP.csv"))

# Put in SPS format
colsh_sps <- data.frame(
	region = rep("Columbia", length(colsh$year)),
	species = rep("Steelhead", length(colsh$year)),
	year = colsh$year,
	spawners = colsh$natural_spawners, # Note this is the fish that returned to spawn, NOT a mistake!
	smoothedSpawners = NA,
	runsize = NA,
	smoothedRunsize = NA,
	source_id = rep("OBMEP_20250228", length(colsh$year))
) 

# Smoothing
colsh_sps$smoothedSpawners <- genSmooth(
	abund = colsh_sps$spawners,
	years = colsh_sps$year,
	genLength = genLength$gen_length[genLength$region == "Columbia" & genLength$species == "Steelhead"]
)


# Add to master sps dataframe
sps_data <- rbind(sps_data, colsh_sps)

# Generate standard figure
plot_abund(colsh_sps)

```


<!--- This code chunk must go after the last data processing step --->
```{r save processed data}
write.csv(sps_data, here("output/sps-data.csv"), row.names = FALSE) # Always have most recent # Not until this script is confirmed to be working 
write.csv(sps_data, paste0(Dropbox_dir, "output/archive/", "sps-data", Sys.Date(), ".csv"), row.names = FALSE) # Archive with date

```


## Expansion Factors 

For species and regions that lacked reliable data on spawner abundance and total return at the appropriate scale, we estimated regional-scale abundance from stream-level surveys. We started with spawner survey data shown in the [Pacific Salmon Explorer](https://www.salmonexplorer.ca). Spawner surveys were each assigned to one of the nine regions we considered based on their geographic location. We note that this is slightly different from how spawner survey data are organized in the Pacific Salmon Explorer, where data are organized by Conservation Units (CUs) that may span regional boundaries (e.g. for pink salmon that have relatively geographically large CUs). In cases of trans-regional CUs, the spawner surveys appear in both regions in the Pacific Salmon Explorer, whereas here we assign spawner surveys to the region in which they fall geographically, regardless of the CU boundary. 

Spawner survey data are largely derived from river-level estimates in  DFO’s [New Salmon Escapement Database System (NuSEDS)](https://open.canada.ca/data/en/dataset/c48669a3-045b-400d-b730-48aafe8c5ee6), but are cleaned up to address issues of, for example, inconsistent naming of streams through time or duplicate data. The spawner survey abundance is equal to the `MAX_ESTIMATE` in NUSEDS for each year and river population, calculated as the maximum of all fields containing spawner abundance data (e.g. natural adult spawners, natural jack spawners, total broodstock removals). Each of these river populations has been designated as an **indicator** stream or **non-indicator**. Indicator streams are observed more consistently in recent decades, tend to have higher spawner abundance, and tend to be monitored using more intensive methods that provide greater accuracy [@EnglishEtAl2018NorthCentralCoast]. For further information on the compilation of spawner survey data, see the [Pacific Salmon Explorer Technical Report](https://bookdown.org/salmonwatersheds/tech-report/analytical-approach.html#spawner-surveys).

Expansion Factor 1, $F_{1,y/d}$, expands the observed spawner abundances in indicator streams to account for indicator streams that are not monitored in a given year. It is calculated for each year $y$ of the spawner time series, and relies on a decadal contribution of each indicator stream to the total escapement to all indicator streams, $P_{d,i}$ in decade $d$ [@EnglishEtAl2018NorthCentralCoast]. The calculation of this decadal contribution requires at least one estimate from each indicator stream for the decade. If a decade does not contain sufficient information (i.e. one or more indicator streams are not monitored at all in a decade), then a reference decade is used to calculate $P_{d,i}$. This reference decade is chosen to be: (1) the closest decade (historical or future) with sufficient information, or failing (1), (2) the 20-year period from 1980-1999 [@ChallengerEtAl2018NorthCentralCoast].

For each decade (or reference decade if insufficient information) $d$, the average number of spawners returning to indicator stream $i$ is calculated as:

$$\bar{S}_{d,i} = \sum_{y = 1}^{Y_{d,i}} \frac{\hat{S}_{y/d, i}}{Y_{d,i}} $$
where $Y_{d,i}$ is the number of years for which spawner estimates are available within decade $d$ for stream $i$.  From the average number of spawners for all indicator streams, the decadal proportional contribution of each indicator stream is calculated as:

$$P_{d,i} = \frac{\bar{S}_{d,i}}{\sum_{i=1}^{I} \bar{S}_{d,i}}$$
where $I$ is the total number of indicator streams.

Expansion Factor 1 is then calculated for each year within the decade $y/d$ based on the decadal contributions and which streams were monitored or not in a given year:

$$F_{1,y/d}=\left( \sum_{i=1}^I P_{d,i} w_{y/d,i} \right)$$
where $w_{y/d,i}$ is 1 if stream $i$ was monitored in year $y$ and 0 if stream $i$ was not monitored in year $y$. Expansion Factor 1 is then multiplied by the sum of the observed spawners in all indicator stream to yield the expanded estimate of spawner abundances in all indicator streams in the region:

$$S'_{y} = F_{1,y/d} \sum_{i=1}^I \hat{S}_{y,i}$$

Expansion Factor 2 $F_{2,d}$ expands the spawner abundance to all indicator streams, $S'_{y}$, to account for non-indcator streams. Unlike Expansion Factor 1, this is calculated for each decade (rather than each year) and then applied to all years within a decade. Like Expansion Factor 1, there needs to be sufficient information within the given decade in order to calculate $F_{2,d}$, or else a reference decade is chosen. See @EnglishEtAl2018NorthCentralCoast for detailed on how reference decades are chosen in that case.

Expansion Factor 2 is calculated as:

$$F_{2,d} = \frac{\sum_{i = 1}^I \bar{S}_{d,i} + \sum_{j = 1}^{J} \bar{S}_{d,j}}{\sum_{i = 1}^I \bar{S}_{d,i}}$$
where $\bar{S}_{d,i}$ and $\bar{S}_{d,j}$ are the deacdal average number of spawners in indicator and non-indicator streams, respectively, calculated above.  $J$ is the total number of non-indicator streams.  The adjusted total number of spawners in both indicator and non-indicator streams is then calculated as:
$$ S''_{y} = F_{2,d} S'_{y} $$

Note that when expanding spawner abundance for spawner-recruit analysis, a third expansion factor is applied to account for streams that are never monitored and for observer (in)efficiency [@PeacockEtAl2020EvaluatingConsequencesCommon]. We did not apply this third expansion factor because it is highly undertain and we are interested in relative changes in abundance through time, so we do not require to expand to absolute abundance.

## Smoothing
We smoothed time series of spawner abundance and total return using a right-aligned running geometric mean over the length of a generation. This reduces the influence of dominant years and produces an index that is less sensitive to stochastic interannual variability that is common in salmon population dynamics. The generation length was based on the dominant life-history type for each species in a particular region. The smoothed spawner abundance in year $y$ given a generation length $g$ was calculated as:
$$ \bar{S_{y}} = \left( \prod_{t = y-g+1}^y {S''_t} \right)^{1/g} $$

We smoothed abundance from the first year to the most recent year of raw abundance data. If there were intermediate years with missing data, the smoothed abundance was still calculated using the available years (i.e. ignoring the missing data, with the exponent $1/g$ adjusted so that $g$ reflected the number of years with data in the generation). If all years in a generation were missing data, than the smoothed abundance was `NA`.

When plotting, we show the smoothed abundance relative to the long-term historical average, so that species that have vastly different abundances within a region can be plotted on the same y-axis for comparison.

## Quantifying change
We summarize the time series of smoothed spawner abundance or total return at the regional scale using three different metrics:

1. **Current status** relative to historical average;
2. **Long-term trend** over the entire time series; and
3. **Short-term trend** over the most recent three generations.

The currents status provides information on how the most recent spawner abundance compares to past values, while the two trend metrics provide information on the average direction of change and are less sensitive to the magnitude of current abundance.

The **current status** relative to historical average is calculated as:
$$ (\bar{S_{y}} - \bf{S}) / \bf{S} $$
where $y$ is the most recent year for which the index of spawner abundance could be calculated and $\bf{S}$ is the average smoothed spawner abundance over the entire time series.

The **long-term trend** is calculated following the recommendations of @DEon-EggertsonEtAl2015ReliableIdentificationDeclining, who found that the correct identification of declines in salmon population abundance may be most reliable when considering the entire time series, and applying regression-based estimates of change calculated from log-transformed and smoothed spawner abundances. As such, we fit a simple linear model to the time series of $\log ( \bar{S}_{y} )$ over $y$. The resulting slope, $m$, gives an average estimate of annual change:

$$ m = \log \left( \frac{\bar{S}_{y}}{\bar{S}_{y-1}} \right)$$.

With some rearranging, we can calculate the average annual percent change as:
$$ \frac{\bar{S}_{y} - \bar{S}_{y-1}}{\bar{S}_{y-1}} = \frac{ e^m \bar{S}_{y-1} - \bar{S}_{y-1}}{\bar{S}_{y-1}} = e^m - 1$$.
If the slope $m$ is not significantly different from zero ($p \geq 0.05$), then the trend is classified as "stable" regardless of the magnitude of the estimated slope. If the slope is significantly different from zero ($p < 0.05$), then the trend is classified as increasing if $m > 0$ or decreasing if $m < 0$.

The **short-term trend** is calculated as described above, but only using the time series over the most recent three generations.

```{r, echo = FALSE, include = TRUE, fig.width = 8, fig.height = 3.5, fig.cap = "**Above:** Simulated time series of spawner abundance (grey line), smoothed with a running geometric mean over the generation length (4 years; black line). The three metrics are illustrated: **Current status** relative to historical average (black);  **Long-term trend** over the entire time series (green); and **Short-term trend** over the most recent three generations (clay). The shaded regions of the regression lines indicate the 95% prediction intervals. In this example, the short-term trend is not statistically significant (p > 0.05), and so the trend is characterized as 'stable' despite the negative slope."}

# Simulate time series over 72 years (1950 - 2021)
S <- numeric(72)
set.seed(23435543)
S[1] <- 12
for(i in 2:72){
	S[i] <- rlnorm(n = 1, meanlog = log(S[i-1] * exp(1.3 - S[i-1]/seq(14, 10, length.out = 72)[i])), sd = 0.2)
}
S[72] <- 12

g <- 4 # generation length
SS <- numeric(72)
for(i in 1:72){
	SS[i] <- prod(S[max(1, (i - g + 1)):i])^(1/length(max(1, (i - g + 1)):i))
}

# Plot
par(mfrow = c(1,1), mar = c(4,4,2,1), cex.lab = 1)

# Current status relative to historical average
plot(1950:2021, SS, "n", bty = "l", las = 1, ylab = "Spawner abundance (thousands)", xlab = "", ylim = c(10, 22))
lines(1950:2021, S, col = SWP_cols['stone1'])
abline(h = mean(SS), lty = 2, col = SWP_cols['stone3'])
points(2021, SS[72], pch = 21, bg = SWP_cols['stone1'], cex = 2)
arrows(2021, mean(SS), 2021, SS[72], lwd = 2, col = 1, length = 0.08)
text(2018, mean(SS), paste0("Current status:\n", round((SS[72] - mean(SS))/SS[72]*100, 1), "%"), col = 1, pos = 3, xpd = NA, font = 2)
# mtext(side = 3, adj = 0, "  a) Current status relative to historical average", cex = 0.8)

# 2. Change over the entire time series
fit <- lm(log(SS) ~ c(1950:2021))
predSS <- predict(fit, se.fit = TRUE)

# plot(1950:2021, SS, "l", bty = "l", lwd = 1.5, las = 1, ylab = "Spawner abundance (thousands)", xlab = "")
polygon(x = c(1950:2021, rev(1950:2021)),
        y = c(exp(predSS$fit - 1.96*predSS$se.fit), rev(exp(predSS$fit + 1.96*predSS$se.fit))),
        border = NA, col = paste0(SWP_cols['tidal2'], 30))
lines(1950:2021, exp(predSS$fit), col = SWP_cols['tidal2'], lwd = 2)

text(1992, 18, paste0("Long-term trend: ", round(coefficients(fit)[2]*100, 1), "% per year (declining)"), col = SWP_cols['tidal2'], xpd = NA, font = 2)
# mtext(side = 3, adj = 0, "  b) Long-term trend over the entire time series", cex = 0.8)

# 2. Change over 3 gens
fit2 <- lm(log(SS[(72 - 3*g + 1):72]) ~ c(1950:2021)[(72 - 3*g + 1):72])
predSS2 <- predict(fit2, se.fit = TRUE)

polygon(x = c(c(1950:2021)[(72 - 3*g + 1):72], rev(c(1950:2021)[(72 - 3*g + 1):72])),
        y = c(exp(predSS2$fit - 1.96*predSS2$se.fit), rev(exp(predSS2$fit + 1.96*predSS2$se.fit))),
        border = NA, col = paste0(SWP_cols['clay1'], 50))
lines(c(1950:2021)[(72 - 3*g + 1):72], exp(predSS2$fit), col = SWP_cols['clay1'], lwd = 2)
text(2007, 11, paste0("Short-term trend: ", round(coefficients(fit2)[2]*100, 1), "% per year (stable)"), col = SWP_cols['clay1'], xpd = NA, font = 2)

lines(1950:2021, SS, lwd = 1.5)
```

### Additional Considerations
The standardized approach to quantifying change described above was applied across regions and species. We then considered two additional attributes prior to reporting on the State of Salmon:

1. If the species and region had no data in the most recent generation, the current status was "Unknown" even if historical abundance information was available. This was the case for **Haida Gwaii Chinook**, for which there were no reliable estimates of spawners since 2006.

2. If there was reliable information that the absolute population size was < 1,500 spawners, then the species and region was considered at risk of extirpation regardless of the outcome of the metrics above. This was the case for **Columbia Chinook** which have been assessed as Endangered by COSEWIC due to low population size [@COSEWIC2017ChinookSalmonOncorhynchus] and **Columbia steelhead**. There were other regions and species that had spawner abundance < 1,500, but none of these had reliable estimates of **absolute** spawner abundance.
