---
title: "PSF's State of Salmon Reporting: Methods & Results"
author: "Salmon Watersheds Program, Steph Peacock"
date: "`r format(Sys.time(), '%B %d, %Y')`"
bibliography: references.bib
header-includes:
   - \usepackage{draftwatermark}
output: 
 html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: true
link-citations: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set colour palette for species
# sp_cols <- viridis(5)
sp_cols <- c("#332288", "#44AA99", "#88CCEE", "#CC6677", "#882255", "#DDCC77") #https://davidmathlogic.com/colorblind
names(sp_cols) <- c("Chinook", "Chum", "Coho", "Pink", "Sockeye", "Steelhead")


```

<head>
<meta name="robots" content="noindex">
</head>

<style type="text/css">

.watermark {
  opacity: 0.2;
  position: fixed;
  top: 50%;
  left: 50%;
  font-size: 500%;
  color: #000000;
}

.blackbox {
  padding: 1em;
  background: black;
  color: white;
  border: 2px solid orange;
  border-radius: 10px;
}
.center {
  text-align: center;
}

</style>


<div class="watermark">DRAFT</div>

# Overview 

Currently, there is no place where the public can readily find answers to general questions regarding the state of salmon in BC and the Yukon. It is difficult to find credible answers to questions such as:

* How are salmon populations doing? 
* What are the major factors influencing the state of salmon? 
* What is being done - and what could be done - to mitigate threats and recover salmon populations and the habitats they depend on?

This lack of publicly accessible information about the state of Pacific salmon and opens the door to misinformation that can hinder and harm our ability to conserve salmon. 

The [Pacific Salmon Foundation (PSF)](https://www.psf.ca) is developing an inaugural **State of Pacific Salmon in Canada** publication (henceforth referred to as *SPS*), which will address these questions and be tailored to the questions of a salmon-informed public audience. The outputs will be updated regularly (e.g. annually or biennially) to reflect new data and changes to the state of Pacific salmon populations.

As part of the SPS report, PSF's [Salmon Watersheds Program](https://www.salmonwatersheds.ca) is developing simple metrics to report on the status and trends in regional abundance for each species. This document outlines the data sources and analytical approach used to estimate time series of spawner abundance and, where available, run size for each species and region, as well as draft results. It is intended to be shared with external partners in order to garner feedback on preliminary results that will inform our approach moving forward. Our analyses are ongoing, and the information here is subject to change on a regular basis, but we will endeavor to keep it up-to-date.

# Methods

## General Approach

We report spawner abundance for each of nine regions that represent all major Pacific salmon-bearing watersheds in Canada: Yukon, Transboundary, Haida Gwaii, Nass, Skeena,  Central Coast, Vancouver Island & Mainland Inlets, Fraser, and Columbia. These regions are also used to organize data in the [Pacific Salmon Explorer](https://www.salmonexplorer.ca). There are a relatively small number of Pacific salmon that spawn in the MacKenzie River basin in Arctic Canada that are currently not considered here.

We separate five species of Pacific salmon: Chinook, chum, coho, pink, and sockeye. We also report status of steelhead trout for regions where spawner data are are available (Nass, Skeena, and interior Fraser). We note that when assessing biological status, pink are often separated into even- and odd-year lineages due to their consistent 2-year life cycle. However, for the general overviews provided in State of Salmon we consider generational averages, which take the running average of even- and odd-year lineages for pink salmon. This approach of using generation running averages also smooths over dominant lines for sockeye salmon, for which many populations display cyclic dominance. Shifts in dominance between even- and odd-year pink populations or declines in sub-dominant years of sockeye salmon are considered in a more nuanced way when discussing how changes in abundance have been reflected in the diversity and distribution of each species within the region. 

For each of these species, where data are available, we construct an index of spawner abundance at the regional scale. We focus on spawner abundance, rather than catch or run size (i.e., catch plus spawners), because data on spawner abundance are more readily summarized at different spatial scales. Further, spawners represents the abundance of salmon available for to meet cultural and ecological needs and thus provides a measure of status relevant to communities and ecosystems, rather than industry. We recognize that commercial catch has historically been a substantial portion of salmon that return to the coast, and that ignoring declines in catch will underestimate the declines in overall salmon abundance. Therefore, we include information on run size for species and regions with reliable data. **This assessment of the data landscape is ongoing, and one of the reasons why the results here should be considered preliminary.** 

For some species and regions, there is a reliable source of data for region-wide escapement and/or run size that we can directly use as an index of abundance. For example, stocks that are governed by international treaties may be monitored by the Pacific Salmon Commission, and tend to have reliable time series of abundance available at regional scales. These data sources are outlined in the [Region-Specific Data section](#region-specific-data). For species and regions where aggregate abundance is not reported at the scale needed, we adapted our approach to make the best use of available data. In most cases, this meant expanding spawner abundance from stream-level estimates to get a regional scale index of spawner abundance using two types of expansion factors [@English2016]. This expansion process is described in the [Expansion Factors section](#expansion-factors).  

R code for the analyses described here are available on GitHub at https://github.com/salmonwatersheds/state-of-salmon.

## Region-Specific Data

Here we describe specific data sources for the abundance of salmon and steelhead at the regional scale. If no data source is listed for a species and region, then we took an alternate approach of estimating regional-scale abundance from finer-scale data (see [Expansion Factors](#expansion-factors)). 

### Yukon

The Canadian portion of the Yukon River is home to Chinook, chum, and coho salmon. Border escapement and run size of Canadian-origin Chinook and fall chum salmon is estimated at the Eagle Sonar station on the Yukon/Alaska border, and are available from the [Yukon River Panel](https://www.yukonriverpanel.com/) in their Joint Technical Committee (JTC) Reports. Specifically, we used Chinook "Spawning escapement estimate" and "Canadian origin total run size estimate" from Appendix B11 of @jtcYukonRiverSalmon2023. Chum salmon data were from Appendix B16 of @jtcYukonRiverSalmon2023. There are also Chinook, chum, and coho salmon in the Canadian portion of the Porcupine River, which joins the mainstem Yukon River in Alaska. Data on escapement to the Porcupine River are more patchy and not currently included here.

<!-- ```{r, echo = FALSE, include = TRUE, fig.cap = "**Above:** Map showing the Canadian portino of the Yukon River drainage. Out estimates of spawner abundance and run size are based on border escapement estimated from Eagle sonar. Enuneration of salmon in the Porcupine river is less consistent and was not included in our analysis."} -->

<!-- ``` -->

### Transboundary

For Chinook, coho, and sockeye, we assessed regional abundance in the Transboundary using estimates of border escapement provided in the Pacific Salmon Commission's [Joint Transboundary Technical Committee Reports](https://www.psc.org/publications/technical-reports/technical-committee-reports/transboundary/). The regional spawner abundance and run size of Chinook was calculated as the sum of escapement and run size, respectively, to the Stikine, Alsek, and Taku watersheds, available from Table B2 of @ctcAnnualReportCatch2023 and provided to E. Hertz in Excel format following a data request. Sockeye spawner abundance and run size was the sum escapement or Terminal Run, respectively, from the Stikine River (Appendix B21 [@ttcFinalEstimatesTransboundary2022]) and the Taku River (Appendix D15; [@ttcFinalEstimatesTransboundary2022]). Coho spawner abundance and run size are available for the Taku River only, taken from Appendix D20 of @ttcFinalEstimatesTransboundary2022.

Pink and chum salmon abundance are less monitored in the Transboundary, with ongoing escapement available only from the Canyon Island fish wheel on the Taku River. We used the index of escapement from the Canyon Island fish wheel as an index of regional spawner abundance [@ttcFinalEstimatesTransboundary2022]. There are some historical data for pink salmon spawner abundance in the Nakina River of the Taku watershed, but this location has not reported data since 1998 and thus we did not include this stream survey in our index of regional abundance.

We recognize that our approach in the Transboundary lacks information from many unmonitored watersheds, in particular the smaller watersheds of the Chilkat, Unuk, and Whiting Rivers. In the absence of better monitoring, we choose to report available data as a proxy for regional abundance, and note that the contribution of these smaller watersheds to total regional abundance for each species is likely small.

### Haida Gwaii

For Haida Gwaii, we followed the Expansion Factors approach described below, with the exception of Chinook salmon for which the only data are from a single enumeration project on the Yakoun River. Thus we report Yakoun River spawner abundance for Haida Gwaii, but unfortunately there were no estimates since 2006. We compared our expansion approach to estimates of escapement to Area 1 reported in the @tcnbCanadaNorthernBoundary2023 (Appendix 30) but found the Area 1 estimates were much lower (except for Chinook, for which Appendix 30 matched the Yakoun River estimates exactly). 

### Nass

Nass sockeye escapement and run size were extracted from Pacific Salmon Commission reports [@tcnbCanadaNorthernBoundary2023,@SCSC2023]. 

The index of Nass steelhead spawner abundance and run size are for the Nass Summer CU, provided by the Nisga’a Fish and Wildlife Department and LGL Ltd. (publication forthcoming). There is another steelhead CU in the Nass region - Nass Winter - which is not well monitored and therefore not included in our index of spawner abundance.

### Skeena

Skeena sockeye escapement and run size were extracted from Pacific Salmon Commission reports [@tcnbCanadaNorthernBoundary2023,@SCSC2023]. 

For Skeena chum salmon, we used escapement to Area 4 reported in Table 32 of @tcnbCanadaNorthernBoundary2023, rather than expanding escapement from spawner surveys, because surveys of indicator stream have been increasingly patchy over the last 30 years, making expansions uncertain.

The index of Skeena steelhead spawner abundance is derived from estimated escapement of Skeena Summer steelhead at the Tyee Test Fishery (1956 - present), provided by the Province via email. As for other regions, these estimates may not capture winter-run steelhead, for which data are not available.

### Fraser

River-level spawner abundance for Fraser Chinook found in NuSEDS can be highly unreliable, thus we based our expansions on only the most intensively monitored streams that have been deemed as reliable estimates of abundance by @Brown2020. These streams are those shown in the [Pacific Salmon Explorer](https://www.salmonexplorer.ca), and the associated data can be accessed from the stream survey data in the [SWP Data Library](https://data.salmonwatersheds.ca/result?datasetid=1). We also truncated the Chinook index of abundance to 1984 and more recent data, as earlier years have been deemed unreliable [@Brown2020] and from 1984 onward spawner estimates mirrored those reported by @Atlas2023. For Chinook run size, we used the sum of total run for four Fraser populations (Harrison, Nicola, Lower Shuswap, and South Thompson - all) reported in @Atlas2023 and available from the associated [GitHub repository](https://github.com/willatlas/ChinookEscapement-RunSize).

For interior Fraser coho CUs, Spawner abundance and run size were provided by DFO on data request (pers. comm. Marissa Glavas, Data Manager, Fraser River Stock Assessment). We report run size as the sum of "Total Pre-Fishery Abundance" to all CUs and spawner abundance as the sum of "Total Return" (i.e., final spawner estimate plus fish removed from the system by DFO Salmon Enhancement Program or First Nations) to all CUs. We recognize that using these estimates ignores coastal coho populations, for which monitoring has been patchy through time.

Data on aggregate run size and spawner abundance of pink salmon and sockeye salmon in the Fraser region are provided by the Pacific Salmon Commission (PSC), and accessed through the [Fraser Panel Annual Report: Data Application](https://psc1.shinyapps.io/PSC_Annual_Fraser/). We note that Fraser River pink salmon are only counted in the dominant, odd-year run. 

Steelhead trout in the Fraser are monitored by the Province in at least 10 different streams, but these data are not readily available. Relatively reliable estimates of steelhead spawner abundance are available at the CU-level for interior Fraser steelhead from the Thompson Summer CU (monitored at the Thompson River) and Mid Fraser Summer CU (primarily monitored at the Chilcotin River). We used the sum of CU-level spawner abundance for these two CUs as an index of Fraser steelhead abundance. We note that this approach does not include more coastal populations, such as the Lower Fraser Summer steelhead monitored in the Coquihalla River or Boundary Bay Winter steelhead, which may not have declined to the same extent over the past decade. However, a lack of publicly accessible data on coastal Fraser steelhead has limited our ability to include these CUs in our index of abundance.

### Columbia

We used CU-level estimates of spawner abundance (run reconstructions) sourced from DFO (Ogden, pers. comm.) for Chinook and @Stockwell2003 and subsequent updates for sockeye. 

There is no monitoring of steelhead trout in the Canadian portion of the Columbia region, but the  Okanagan Nation Alliance does enumerate steelhead in akskwəkwant (Inkaneep Creek) and estimate a Canadian portion of steelhead spawning abundance. These data can be found in associated report (e.g. @OBMEP2023) and are available in the Pacific Salmon Explorer as the CU-level spawner abundance for the Mid Columbia Summer CU.

## Expansion Factors

For species and regions that lacked reliable data on spawner abundance and run size at the appropriate scale, we estimated regional-scale abundance from stream-level surveys. We started with spawner survey data shown in the [Pacific Salmon Explorer](https://www.salmonexplorer.ca). Spawner surveys were each assigned to one of the nine regions we considered based on their geographic location. We note that this is slightly different from how spawner survey data are organized in the Pacific Salmon Explorer, where data are organized by Conservation Units (CUs) that may span regional boundaries (e.g. for pink salmon that have relatively geographically large CUs). In cases of trans-regional CUs, the spawner surveys appear in both regions in the Pacific Salmon Explorer, whereas here we assign spawner surveys to the region in which they fall geographically, regardless of the CU boundary. 

Spawner survey data are largely derived from river-level estimates in  DFO’s [New Salmon Escapement Database System (NuSEDS)](https://open.canada.ca/data/en/dataset/c48669a3-045b-400d-b730-48aafe8c5ee6), but are cleaned up to address issues of, for example, inconsistent naming of streams through time or duplicate data. The spawner survey abundance is equal to the `MAX_ESTIMATE` in NUSEDS for each year and river population, calculated as the maximum of all fields containing spawner abundance data (e.g. natural adult spawners, natural jack spawners, total broodstock removals). Each of these river populations has been designated as an **indicator** stream or **non-indicator**. Indicator streams are observed more consistently in recent decades, tend to have higher spawner abundance, and tend to be monitored using more intensive methods that provide greater accuracy [@English2016]. For further information on the compilation of spawner survey data, see the [Pacific Salmon Explorer Technical Report](https://bookdown.org/salmonwatersheds/tech-report/analytical-approach.html#spawner-surveys).

Expansion Factor 1, $F_{1,y/d}$, expands the observed spawner abundances in indicator streams to account for indicator streams that are not monitored in a given year. It is calculated for each year $y$ of the spawner time series, and relies on a decadal contribution of each indicator stream to the total escapement to all indicator streams, $P_{d,i}$ in decade $d$ [@English2016]. The calculation of this decadal contribution requires at least one estimate from each indicator stream for the decade. If a decade does not contain sufficient information (i.e. one or more indicator streams are not monitored at all in a decade), then a reference decade is used to calculate $P_{d,i}$. This reference decade is chosen to be: (1) the closest decade (historical or future) with sufficient information, or failing (1), (2) the 20-year period from 1980-1999 [@Challenger2018].

For each decade (or reference decade if insufficient information) $d$, the average number of spawners returning to indicator stream $i$ is calculated as:

$$\bar{S}_{d,i} = \sum_{y = 1}^{Y_{d,i}} \frac{\hat{S}_{y/d, i}}{Y_{d,i}} $$
where $Y_{d,i}$ is the number of years for which spawner estimates are available within decade $d$ for stream $i$.  From the average number of spawners for all indicator streams, the decadal proportional contribution of each indicator stream is calculated as:

$$P_{d,i} = \frac{\bar{S}_{d,i}}{\sum_{i=1}^{I} \bar{S}_{d,i}}$$
where $I$ is the total number of indicator streams.

Expansion Factor 1 is then calculated for each year within the decade $y/d$ based on the decadal contributions and which streams were monitored or not in a given year:

$$F_{1,y/d}=\left( \sum_{i=1}^I P_{d,i} w_{y/d,i} \right)$$
where $w_{y/d,i}$ is 1 is stream $i$ was monitored in year $y$ and 0 if stream $i$ was not monitored in year $y$. Expansion Factor 1 is then multiplied by the sum of the observed spawners in all indicator stream to yield the expanded estimate of spawner abundances in all indicator streams in the region:

$$S'_{y} = F_{1,y/d} \sum_{i=1}^I \hat{S}_{y,i}$$

Expansion Factor 2 $F_{2,d}$ expands the spawner abundance to all indicator streams, $S'_{y}$, to account for non-indcator streams. Unlike Expansion Factor 1, this is calculated for each decade (rather than each year) and then applied to all years within a decade. Like Expansion Factor 1, there needs to be sufficient information within the given decade in order to calculate $F_{2,d}$, or else a reference decade is chosen. See @English2016 for detailed on how reference decades are chosen in that case.

Expansion Factor 2 is calculated as:

$$F_{2,d} = \frac{\sum_{i = 1}^I \bar{S}_{d,i} + \sum_{j = 1}^{J} \bar{S}_{d,j}}{\sum_{i = 1}^I \bar{S}_{d,i}}$$
where $\bar{S}_{d,i}$ and $\bar{S}_{d,j}$ are the deacdal average number of spawners in indicator and non-indicator streams, respectively, calculated above.  $J$ is the total number of non-indicator streams.  The adjusted total number of spawners in both indicator and non-indicator streams is then calculated as:
$$ S''_{y} = F_{2,d} S'_{y} $$

Note that when expanding spawner abundance for spawner-recruit analysis, a third expansion factor is applied to account for streams that are never monitored and for observer (in)efficiency [@Peacock2020]. We did not apply this third expansion factor because it is highly undertain and we are interested in relative changes in abundance through time, so we do not require to expand to absolute abundance.

## Smoothing
We smoothed time series of spawner abundance and run size using a right-aligned running geometric mean over the length of a generation. This reduces the influence of dominant years and produces an index that is less sensitive to stochastic interannual variability that is common in salmon population dynamics. The generation length was based on the dominant life-history type for each species in a particular region. The smoothed spawner abundance in year $y$ given a generation length $g$ was calculated as:
$$ \bar{S_{y}} = \left( \prod_{t = y-g+1}^y {S''_t} \right)^{1/g} $$

We smoothed abundance from the first year to the most recent year of raw abundance data. If there were intermediate years with missing data, the smoothed abundance was still calculated using the available years (i.e. ignoring the missing data, with the exponent $1/g$ adjusted so that $g$ reflected the number of years with data in the generation). If all years in a generation were missing data, than the smoothed abundance was `NA`.

When plotting, we show the smoothed abundance relative to the long-term historical average, so that species that have vastly different abundances within a region can be plotted on the same y-axis for comparison.

## Quantifying change
We summarize the time series of smoothed spawner abundance or run size at the regional scale using three different metrics:

1. **Current status** relative to historical average;
2. **Long-term trend** over the entire time series; and
3. **Short-term trend** over the most recent three generations.

The first metric provides information on how the most recent spawner abundance compares to past values, while the second two metrics provide information on the average direction of change and are less sensitive to the magnitude of current abundance.

The **current status** relative to historical average is calculated as:
$$ (\bar{S_{y}} - \bf{S}) / \bf{S} $$
where $y$ is the most recent year for which the index of spawner abundance could be calculated and $\bf{S}$ is the average smoothed spawner abundance over the entire time series.

The **long-term trend** is calculated following the recommendations of @DEon-Eggertson2015, who found that the correct identification of declines in salmon population abundance may be most reliable when considering the entire time series, and applying regression-based estimates of change calculated from log-transformed and smoothed spawner abundances. As such, we fit a simple linear model to the time series of $\log ( \bar{S}_{y} )$ over $y$. The resulting slope, $m$, gives an average estimate of annual change:

$$ m = \log \left( \frac{\bar{S}_{y}}{\bar{S}_{y-1}} \right)$$.

With some rearranging, we can calculate the average annual percent change as:
$$ \frac{\bar{S}_{y} - \bar{S}_{y-1}}{\bar{S}_{y-1}} = \frac{ e^m \bar{S}_{y-1} - \bar{S}_{y-1}}{\bar{S}_{y-1}} = e^m - 1$$.
If the slope $m$ is not significantly different from zero ($p \geq 0.05$), then the trend is classified as "stable" regardless of the magnitude of the estimated slope. If the slope is significantly different from zero ($p < 0.05$), then the trend is classified as increasing if $m > 0$ or decreasing if $m < 0$.

The **short-term trend** is calculated as described above, but only using the time series over the most recent three generations.

```{r, echo = FALSE, include = TRUE, fig.width = 8, fig.height = 3.5, fig.cap = "**Above:** Simulated time series of spawner abundance (grey line), smoothed with a running geometric mean over the generation length (4 years; black line). The three metrics are illustrated: Metric 1: **Current status** relative to historical average (red); Metric 2: **Long-term trend** over the entire time series (green); and Metric 3: **Short-term trend** over the most recent three generations (blue). The shaded regions of the regression lines indicate the 95% prediction intervals. In this example, the short-term trend is not statistically significant (p > 0.05), and so the trend is characterized as 'stable' despite the negative slope."}

# Simulate time series over 72 years (1950 - 2021)
S <- numeric(72)
set.seed(23435543)
S[1] <- 12
for(i in 2:72){
	S[i] <- rlnorm(n = 1, meanlog = log(S[i-1] * exp(1.3 - S[i-1]/seq(14, 10, length.out = 72)[i])), sd = 0.2)
}
S[72] <- 12

g <- 4 # generation length
SS <- numeric(72)
for(i in 1:72){
	SS[i] <- prod(S[max(1, (i - g + 1)):i])^(1/length(max(1, (i - g + 1)):i))
}

# Plot
par(mfrow = c(1,1), mar = c(4,4,2,1), cex.lab = 1)

# Current status relative to historical average
plot(1950:2021, SS, "n", bty = "l", las = 1, ylab = "Spawner abundance (thousands)", xlab = "", ylim = c(10, 22))
lines(1950:2021, S, col = "#00000030")
abline(h = mean(SS), lty = 2)
points(2021, SS[72], pch = 21, bg = 1, cex = 2)
arrows(2021, mean(SS), 2021, SS[72], lwd = 2, col = sp_cols[4], length = 0.08)
text(2018, mean(SS), paste0("Metric 1: ", round((SS[72] - mean(SS))/SS[72]*100, 1), "%"), col = sp_cols[4], pos = 3, xpd = NA, font = 2)
# mtext(side = 3, adj = 0, "  a) Current status relative to historical average", cex = 0.8)

# 2. Change over the entire time series
fit <- lm(log(SS) ~ c(1950:2021))
predSS <- predict(fit, se.fit = TRUE)

# plot(1950:2021, SS, "l", bty = "l", lwd = 1.5, las = 1, ylab = "Spawner abundance (thousands)", xlab = "")
polygon(x = c(1950:2021, rev(1950:2021)),
        y = c(exp(predSS$fit - 1.96*predSS$se.fit), rev(exp(predSS$fit + 1.96*predSS$se.fit))),
        border = NA, col = paste0(sp_cols[2], 30))
lines(1950:2021, exp(predSS$fit), col = sp_cols[2], lwd = 2)

text(1992, 18, paste0("Metric 2: ", round(coefficients(fit)[2]*100, 1), "% per year (declining)"), col = sp_cols[2], xpd = NA, font = 2)
# mtext(side = 3, adj = 0, "  b) Long-term trend over the entire time series", cex = 0.8)

# 2. Change over 3 gens
fit2 <- lm(log(SS[(72 - 3*g + 1):72]) ~ c(1950:2021)[(72 - 3*g + 1):72])
predSS2 <- predict(fit2, se.fit = TRUE)

polygon(x = c(c(1950:2021)[(72 - 3*g + 1):72], rev(c(1950:2021)[(72 - 3*g + 1):72])),
        y = c(exp(predSS2$fit - 1.96*predSS2$se.fit), rev(exp(predSS2$fit + 1.96*predSS2$se.fit))),
        border = NA, col = paste0(sp_cols[1], 50))
lines(c(1950:2021)[(72 - 3*g + 1):72], exp(predSS2$fit), col = sp_cols[1], lwd = 2)
text(2007, 11, paste0("Metric 3: ", round(coefficients(fit2)[2]*100, 1), "% per year (stable)"), col = sp_cols[1], xpd = NA, font = 2)

lines(1950:2021, SS, lwd = 1.5)
```

# Results 

:::: {.blackbox data-latex=""}
::: {.center data-latex=""}
**NOTICE!**
:::

**These are PRELIMINARY results and the following figures should not be taken as a statement on the status of salmon in these regions.** At this time, we are working to verify datasets and refine our methods.
::::



```{r, echo = FALSE, message = FALSE, warning = FALSE}
# library(viridis)
library(reactable)
library(reactablefmtr)
library(dplyr)
library(abind)
library(tools)

# Read in data for plots

sps_dat <- read.csv("output/sps-data.csv")
sps_metrics <- read.csv("output/sps-metrics.csv")

regions <- unique(unique(sps_dat$region))

# Source script with code for plotting functions
source("code/functions.R")

```

## All regions

**Below**: Summary of current status relative to historical average (Metric 1) for **spawner abundance**. Regions and species below historical average are in red and above historical average are in green. Black cells indicate species that are not thought to be present in the given region, and white cells indicate there were no available abundance data for that species and region combination. 
```{r, echo = FALSE, message = FALSE, warning = FALSE}
btn_table.all(sps_metrics[which(sps_metrics$type == "Spawners"), c("region", "species", "status")])
```

**Below**: Summary of current status relative to historical average (Metric 1) for **run size**. Regions and species below historical average are in red and above historical average are in green. Black cells indicate species that are not thought to be present in the given region, and white cells indicate there were no available abundance data for that species and region combination. 
```{r, echo = FALSE, message = FALSE, warning = FALSE}
btn_table.all(sps_metrics[which(sps_metrics$type == "Run Size"), c("region", "species", "status")])
```

## Yukon

**Highlights**

* Chinook and chum salmon spawners have both experienced precipitous declines in recent years.
* Though smoothed Chinook spawner abundance is only just below historical average, with no significant trend in the most recent 3 generations, raw data show an even sharper decline in the recent 3 years that is not yet reflected in the running geometric average due to the long generation time of Chinook (6 years).
* Trends in Chinook run size show statistically significant declines of 3% and 4% **per year** over long and short time frames, respectively.   

```{r, echo = FALSE, include = TRUE, fig.width = 10, fig.height = 4, fig.cap = "**Above:** Index of spawner abundance through time. Spawner abundance is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance. The percent shown is the percent change from the historical average to the most recent year.", message = FALSE, warning = FALSE}
plot.regional_abund(
  selected_region = "Yukon",
	abund = "spawners", # Select "spawners" or "runsize"
	sps_metrics = sps_metrics, 
	sps_dat = sps_dat,
	sp_cols = sp_cols)

```

```{r, echo = FALSE, include = TRUE, fig.width = 10, fig.height = 4, fig.cap = "**Above:** Index of run size through time. Run size  is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance. The percent shown is the percent change from the historical average to the most recent year.", message = FALSE, warning = FALSE}
plot.regional_abund(
  selected_region = "Yukon",
	abund = "runsize", # Select "spawners" or "runsize"
	sps_metrics = sps_metrics, 
	sps_dat = sps_dat,
	sp_cols = sp_cols)

```


**Below:** Summary of the three metrics for spawners and run size. For long-term (LT) trends and three-generation (3-Gen) trends, the percentage is the average annual change calculated over the respective time period. Grey values indicate the % change was not significantly different from zero (i.e., stable), red values indicate declining trends, and green values indicate increasing trends.
```{r, echo = FALSE, include = TRUE, message = FALSE, warning = FALSE}

btn_table(
  selected_region = "Yukon",
  sps_metrics = sps_metrics
)
```

## Transboundary

**Highlights**

* Most species in the Transboundary are below historical average spawner abundance, except pink salmon which have increased from a low in 2014.
* Although aggregate **sockeye** spawner abundance is still slightly below long-term average, Transboundary sockeye are doing better than anywhere else in the Province, with **stable** long-term and 3-generation trends in spawner abundance.
* As in many other regions, the most recent index of run size is lower than the index of spawner abundance for the three Transboundary species that have both types of data (Chinook: -63% compared to -58%, coho: -42% compared to -23%, and sockeye: -20% compared to -2%), indicating that **declines in catch** have also occurred in recent years.


```{r, echo = FALSE, include = TRUE, fig.width = 10, fig.height = 4, fig.cap = "**Above:** Index of spawner abundance through time. Spawner abundance is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance. The percent shown is the percent change from the historical average to the most recent year.", message = FALSE, warning = FALSE}
plot.regional_abund(
  selected_region = "Transboundary",
	abund = "spawners", # Select "spawners" or "runsize"
	sps_metrics = sps_metrics, 
	sps_dat = sps_dat,
	sp_cols = sp_cols)

```

```{r, echo = FALSE, include = TRUE, fig.width = 10, fig.height = 4, fig.cap = "**Above:** Index of run size through time. Run size  is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance. The percent shown is the percent change from the historical average to the most recent year.", message = FALSE, warning = FALSE}
plot.regional_abund(
  selected_region = "Transboundary",
	abund = "runsize", # Select "spawners" or "runsize"
	sps_metrics = sps_metrics, 
	sps_dat = sps_dat,
	sp_cols = sp_cols)

```


**Below:** Summary of the three metrics for spawners and run size. For long-term (LT) trends and three-generation (3-Gen) trends, the percentage is the average annual change calculated over the respective time period. Grey values indicate the % change was not significantly different from zero (i.e., stable), red values indicate declining trends, and green values indicate increasing trends.
```{r, echo = FALSE, include = TRUE, message = FALSE, warning = FALSE}

btn_table(
  selected_region = "Transboundary",
  sps_metrics = sps_metrics
)
```

## Haida Gwaii

**Highlights**

* All species for which data were available have declined over long- and short-time frames.
* Unlike in other regions, where pink salmon have been doing relatively well, pink salmon have shown the most dramatic declines, and most recent abundance was 90.9% lower than the previous generation and 92.3% lower than the historical average.
* There is only one indicator stream for Chinook in Haida Gwaii,the Yakoun River, which has not been monitored since 2006 and althjough abundance was relatively high at that time, current status is unknown.

```{r, echo = FALSE, include = TRUE, fig.width = 10, fig.height = 4, fig.cap = "**Above:** Index of spawner abundance through time. Spawner abundance is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance. The percent shown is the percent change from the historical average to the most recent year.", message = FALSE, warning = FALSE}
plot.regional_abund(
  selected_region = "Haida Gwaii",
	abund = "spawners", # Select "spawners" or "runsize"
	sps_metrics = sps_metrics, 
	sps_dat = sps_dat,
	sp_cols = sp_cols)

```

```{r, echo = FALSE, include = TRUE, fig.width = 10, fig.height = 4, fig.cap = "**Above:** Index of run size through time. Run size  is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance. The percent shown is the percent change from the historical average to the most recent year.", message = FALSE, warning = FALSE}
plot.regional_abund(
  selected_region = "Haida Gwaii",
	abund = "runsize", # Select "spawners" or "runsize"
	sps_metrics = sps_metrics, 
	sps_dat = sps_dat,
	sp_cols = sp_cols)

```


**Below:** Summary of the three metrics for spawners and run size. For long-term (LT) trends and three-generation (3-Gen) trends, the percentage is the average annual change calculated over the respective time period. Grey values indicate the % change was not significantly different from zero (i.e., stable), red values indicate declining trends, and green values indicate increasing trends.
```{r, echo = FALSE, include = TRUE, message = FALSE, warning = FALSE}

btn_table(
  selected_region = "Haida Gwaii",
  sps_metrics = sps_metrics
)
```

## Nass

**Highlights**

* While most species show declines over both long- and short-time frames, Nass pink salmon have rebounded to above historical average abundance in the most recent generation.
* Chinook salmon have shown the biggest declines from historical average abundance (-60%) with declines persisting in the most recent generation.
* Although sockeye spawner abundance and run size are below historical average, trends in spawner abundance are stable (long-term) or slightly increasing (3-gen).

```{r, echo = FALSE, include = TRUE, fig.width = 10, fig.height = 4, fig.cap = "**Above:** Index of spawner abundance through time. Spawner abundance is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance. The percent shown is the percent change from the historical average to the most recent year.", message = FALSE, warning = FALSE}
plot.regional_abund(
  selected_region = "Nass",
	abund = "spawners", # Select "spawners" or "runsize"
	sps_metrics = sps_metrics, 
	sps_dat = sps_dat,
	sp_cols = sp_cols)

```

```{r, echo = FALSE, include = TRUE, fig.width = 10, fig.height = 4, fig.cap = "**Above:** Index of run size through time. Run size  is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance. The percent shown is the percent change from the historical average to the most recent year.", message = FALSE, warning = FALSE}
plot.regional_abund(
  selected_region = "Nass",
	abund = "runsize", # Select "spawners" or "runsize"
	sps_metrics = sps_metrics, 
	sps_dat = sps_dat,
	sp_cols = sp_cols)

```


**Below:** Summary of the three metrics for spawners and run size. For long-term (LT) trends and three-generation (3-Gen) trends, the percentage is the average annual change calculated over the respective time period. Grey values indicate the % change was not significantly different from zero (i.e., stable), red values indicate declining trends, and green values indicate increasing trends.
```{r, echo = FALSE, include = TRUE, message = FALSE, warning = FALSE}

btn_table(
  selected_region = "Nass",
  sps_metrics = sps_metrics
)
```

## Skeena

**Highlights**

* All species are well below historical average spawner abundance, with chum salmon having declined by `r -round(sps_metrics$status[sps_metrics$region == "Skeena" & sps_metrics$species == "Chum" & sps_metrics$type == "Spawners"]*100, 1)`%.
* Run size is only available for Skeena sockeye, but shows signficant declines over the entire timeseries with current run size `r -round(sps_metrics$status[sps_metrics$region == "Skeena" & sps_metrics$species == "Sockeye" & sps_metrics$type == "Run Size"]*100, 1)`% below historical average. Other research has found that this decline has been accompanied by erosion of diversity, with enhanced stocks increasingly dominating returns [@Price2021].
* Estimates of regional spawner abundance have become increasingly uncertain since 2000, especially for Chinook and coho, due to a decline in monitoring of indicator streams.

```{r, echo = FALSE, include = TRUE, fig.width = 10, fig.height = 4, fig.cap = "**Above:** Index of spawner abundance through time. Spawner abundance is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance. The percent shown is the percent change from the historical average to the most recent year.", message = FALSE, warning = FALSE}
plot.regional_abund(
  selected_region = "Skeena",
	abund = "spawners", # Select "spawners" or "runsize"
	sps_metrics = sps_metrics, 
	sps_dat = sps_dat,
	sp_cols = sp_cols)

```

```{r, echo = FALSE, include = TRUE, fig.width = 10, fig.height = 4, fig.cap = "**Above:** Index of run size through time. Run size  is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance. The percent shown is the percent change from the historical average to the most recent year.", message = FALSE, warning = FALSE}
plot.regional_abund(
  selected_region = "Skeena",
	abund = "runsize", # Select "spawners" or "runsize"
	sps_metrics = sps_metrics, 
	sps_dat = sps_dat,
	sp_cols = sp_cols)

```


**Below:** Summary of the three metrics for spawners and run size. For long-term (LT) trends and three-generation (3-Gen) trends, the percentage is the average annual change calculated over the respective time period. Grey values indicate the % change was not significantly different from zero (i.e., stable), red values indicate declining trends, and green values indicate increasing trends.
```{r, echo = FALSE, include = TRUE, message = FALSE, warning = FALSE}

btn_table(
  selected_region = "Skeena",
  sps_metrics = sps_metrics
)
```


## Central Coast

**Highlights**

* All species are below historical average abundance except pink salmon, with dramatic declines of culturally and economically important sockeye, Chinook, and chum salmon.
* Pink salmon have increased only recently from an all-time low in 2019, more than doubling between 2019 and 2021.
* Expansion factors for regional-scale reconstructions have increased substantially since the mid-2000s, with a decline in monitoring creating uncertainties in our results for all species, but especially coho, sockeye, and Chinook.

```{r, echo = FALSE, include = TRUE, fig.width = 10, fig.height = 4, fig.cap = "**Above:** Index of spawner abundance through time. Spawner abundance is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance. The percent shown is the percent change from the historical average to the most recent year.", message = FALSE, warning = FALSE}
plot.regional_abund(
  selected_region = "Central Coast",
	abund = "spawners", # Select "spawners" or "runsize"
	sps_metrics = sps_metrics, 
	sps_dat = sps_dat,
	sp_cols = sp_cols)

```

```{r, echo = FALSE, include = TRUE, fig.width = 10, fig.height = 4, fig.cap = "**Above:** Index of run size through time. Run size  is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance. The percent shown is the percent change from the historical average to the most recent year.", message = FALSE, warning = FALSE}
plot.regional_abund(
  selected_region = "Central Coast",
	abund = "runsize", # Select "spawners" or "runsize"
	sps_metrics = sps_metrics, 
	sps_dat = sps_dat,
	sp_cols = sp_cols)

```


**Below:** Summary of the three metrics for spawners and run size. For long-term (LT) trends and three-generation (3-Gen) trends, the percentage is the average annual change calculated over the respective time period. Grey values indicate the % change was not significantly different from zero (i.e., stable), red values indicate declining trends, and green values indicate increasing trends.
```{r, echo = FALSE, include = TRUE, message = FALSE, warning = FALSE}

btn_table(
  selected_region = "Central Coast",
  sps_metrics = sps_metrics
)
```


## Vancouver Island & Mainland Inlets

**Highlights**

* All all species except Chinook are currently below historical average.
* Steelhead and coho were below average and declining for all 3 metrics.
* Recent (3-generation) trends are stable or increasing for Chinook, chum, pink, and sockeye, suggestion possibility for recovery.
* Dynamics through time are slightly more stable for the regional aggregate because there are many component populations in this large area, but some local trends are concerning (e.g., WCVI Chinook).

```{r, echo = FALSE, include = TRUE, fig.width = 10, fig.height = 4, fig.cap = "**Above:** Index of spawner abundance through time. Spawner abundance is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance. The percent shown is the percent change from the historical average to the most recent year.", message = FALSE, warning = FALSE}
plot.regional_abund(
  selected_region = "Vancouver Island & Mainland Inlets",
	abund = "spawners", # Select "spawners" or "runsize"
	sps_metrics = sps_metrics, 
	sps_dat = sps_dat,
	sp_cols = sp_cols)

```

```{r, echo = FALSE, include = TRUE, fig.width = 10, fig.height = 4, fig.cap = "**Above:** Index of run size through time. Run size  is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance. The percent shown is the percent change from the historical average to the most recent year.", message = FALSE, warning = FALSE}
plot.regional_abund(
  selected_region = "Vancouver Island & Mainland Inlets",
	abund = "runsize", # Select "spawners" or "runsize"
	sps_metrics = sps_metrics, 
	sps_dat = sps_dat,
	sp_cols = sp_cols)

```


**Below:** Summary of the three metrics for spawners and run size. For long-term (LT) trends and three-generation (3-Gen) trends, the percentage is the average annual change calculated over the respective time period. Grey values indicate the % change was not significantly different from zero (i.e., stable), red values indicate declining trends, and green values indicate increasing trends.
```{r, echo = FALSE, include = TRUE, message = FALSE, warning = FALSE}

btn_table(
  selected_region = "Vancouver Island & Mainland Inlets",
  sps_metrics = sps_metrics
)
```

## Fraser

**Highlights**

* Despite the concern over Fraser sockeye abundance since the crash of 2009, sockeye spawner abundance is still below the long-term average and has declined in the most recent generation.
* Other species are actually above historical average, **however** the these trends do not account for declining catch of the commercial fishing industry in this area. Because fishing pressure on Fraser salmon has historically been higher than in other regions, declines in overall run size (i.e., spawners + catch) mean that overall abundance has likely declined for all species. (**SP: We could include a plot of catch in Fraser areas for each species...)
* Although coho spawners have increased over both short- and long-time frames, coho fisheries have been decimated since concerns over Fraser coho stocks in the mid-1990s.
* Recent declines in Chinook, chum, pink, and sockeye may reflect the increasing pressures of climate change on these southern and interior populations.
```{r, echo = FALSE, include = TRUE, fig.width = 10, fig.height = 4, fig.cap = "**Above:** Index of spawner abundance through time. Spawner abundance is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance. The percent shown is the percent change from the historical average to the most recent year.", message = FALSE, warning = FALSE}
plot.regional_abund(
  selected_region = "Fraser",
	abund = "spawners", # Select "spawners" or "runsize"
	sps_metrics = sps_metrics, 
	sps_dat = sps_dat,
	sp_cols = sp_cols)

```

```{r, echo = FALSE, include = TRUE, fig.width = 10, fig.height = 4, fig.cap = "**Above:** Index of run size through time. Run size  is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance. The percent shown is the percent change from the historical average to the most recent year.", message = FALSE, warning = FALSE}
plot.regional_abund(
  selected_region = "Fraser",
	abund = "runsize", # Select "spawners" or "runsize"
	sps_metrics = sps_metrics, 
	sps_dat = sps_dat,
	sp_cols = sp_cols)

```


**Below:** Summary of the three metrics for spawners and run size. For long-term (LT) trends and three-generation (3-Gen) trends, the percentage is the average annual change calculated over the respective time period. Grey values indicate the % change was not significantly different from zero (i.e., stable), red values indicate declining trends, and green values indicate increasing trends.
```{r, echo = FALSE, include = TRUE, message = FALSE, warning = FALSE}

btn_table(
  selected_region = "Fraser",
  sps_metrics = sps_metrics
)
```
## Columbia

**Highlights**

* Columbia Chinook and sockeye show promise of rebuilding, with positive long-term trends. (We currently only have data to 2020, but record returns of Okanagan sockeye [were celebrated in 2022](https://www.cbc.ca/news/canada/british-columbia/sockeye-salmon-okangan-fishery-osoyoos-1.6571895).)
* The time series for steelhead and Chinook are short, making it difficult to establish a meaningful baseline. (For this reason, the biological status of Okanagan steelhead is not assessed in the [Pacific Salmon Explorer](https://www.salmonexplorer.ca).)
* Columbia steelhead have declined recently to `r `-round(sps_metrics$status[sps_metrics$region == "Columbia" & sps_metrics$species == "Steelhead" & sps_metrics$type == "Spawners"]*100, 1)`% below the 2006-2021 average.

```{r, echo = FALSE, include = TRUE, fig.width = 10, fig.height = 4, fig.cap = "**Above:** Index of spawner abundance through time. Spawner abundance is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance. The percent shown is the percent change from the historical average to the most recent year.", message = FALSE, warning = FALSE}
plot.regional_abund(
  selected_region = "Columbia",
	abund = "spawners", # Select "spawners" or "runsize"
	sps_metrics = sps_metrics, 
	sps_dat = sps_dat,
	sp_cols = sp_cols)

```

```{r, echo = FALSE, include = TRUE, fig.width = 10, fig.height = 4, fig.cap = "**Above:** Index of run size through time. Run size  is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance. The percent shown is the percent change from the historical average to the most recent year.", message = FALSE, warning = FALSE}
plot.regional_abund(
  selected_region = "Columbia",
	abund = "runsize", # Select "spawners" or "runsize"
	sps_metrics = sps_metrics, 
	sps_dat = sps_dat,
	sp_cols = sp_cols)

```


**Below:** Summary of the three metrics for spawners and run size. For long-term (LT) trends and three-generation (3-Gen) trends, the percentage is the average annual change calculated over the respective time period. Grey values indicate the % change was not significantly different from zero (i.e., stable), red values indicate declining trends, and green values indicate increasing trends.
```{r, echo = FALSE, include = TRUE, message = FALSE, warning = FALSE}

btn_table(
  selected_region = "Columbia",
  sps_metrics = sps_metrics
)
```



# References

<div id="refs"></div>

<!-- # Supplemental results -->

<!-- ## Expansion Factors -->

<!-- The following figures show the values of Expansion factors 1 (solid line) and 2 (dashed line) that were multiplied by the sum of observed spawners return to indicator stream to calculate the regional index of spawner abundance for each species. -->

<!-- ```{r, echo = FALSE, include = TRUE, fig.width = 8, fig.height = 4, fig.cap=""} -->
<!-- for(r in regions){ -->
<!--   plot.regional_expansion(selected_region = r, -->
<!--                        spawners = spawners, -->
<!--                        sp_cols = sp_cols) -->
<!-- } -->
<!-- ``` -->