---
title: "PSF's State of Salmon Reporting: Methods & Results"
author: "Salmon Watersheds Program, Steph Peacock"
date: "2023-08-28"
bibliography: references.bib
header-includes:
   - \usepackage{draftwatermark}
output: 
 html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<head>
<meta name="robots" content="noindex">
</head>

<style type="text/css">

.watermark {
  opacity: 0.2;
  position: fixed;
  top: 50%;
  left: 50%;
  font-size: 500%;
  color: #000000;
}

.blackbox {
  padding: 1em;
  background: black;
  color: white;
  border: 2px solid orange;
  border-radius: 10px;
}
.center {
  text-align: center;
}

</style>


<div class="watermark">DRAFT</div>

# Overview 

Currently, there is no place where the public can readily find answers to general questions regarding the state of salmon in BC and the Yukon. It is difficult to find credible answers to questions such as:

* How are salmon populations doing? 
* What are the major factors influencing the state of salmon? 
* What is being done - and what could be done - to mitigate threats and recover salmon populations and the habitats they depend on?

This lack of publicly accessible information about the state of Pacific salmon and opens the door to misinformation that can hinder and harm our ability to conserve salmon. 

The [Pacific Salmon Foundation (PSF)](https://www.psf.ca) is developing an inaugural **State of Pacific Salmon in Canada** publication (henceforth referred to as *SPS*), which will address these questions and be tailored to the questions of a salmon-informed public audience. The outputs will be updated annually to reflect new data and changes to the state of Pacific salmon populations.

As part of the SPS report, PSF's [Salmon Watersheds Program](https://www.salmonwatersheds.ca) has developed an approach to estimating the regional abundance of salmon spawners for each species as a simple metric that provides information on how the number of salmon returning to spawn has been changing. This document outlines the data sources and analytical approach used to estimate time series of spawner abundance for each species and region, as well as draft results. It is intended to be shared with external partners in order to garner feedback on preliminary results that will inform . Our analyses are ongoing, and the information here is subject to change on a regular basis, but we will endeavor to keep it up-to-date.

# Methods

## General Approach

We report spawner abundance for each of nine regions that represent all major Pacific salmon-bearing watersheds in Canada: Yukon, Transboundary, Nass, Skeena, Haida Gwaii, Central Coast, Vancouver Island & Mainland Inlets, Fraser, and Columbia. These regions are also used to organize data in the [Pacific Salmon Explorer](https://www.salmonexplorer.ca). There are a relatively small number of Pacific salmon that spawn in the MacKenzie River basin in Arctic Canada that are currently not considered here.

We separate five species of Pacific salmon: Chinook, chum, coho, pink, and sockeye. (**Note from SP: Need to discuss steelhead.**) We note that when assessing biological status, pink are often separated into even- and odd-year lineages due to their consistent 2-year life cycle. However, for the general overviews provided in State of Salmon we consider generational averages, which take the running average of even- and odd-year lineages for pink salmon. This approach of using generation running averages also smooths over dominant lines for sockeye salmon, for which many populations display cyclic dominance. Shifts in dominance between even- and odd-year pink populations or declines in sub-dominant years of sockeye salmon are considered in a more nuanced way when discussing how changes in abundance have been reflected in the diversity and distribution of each species within the region. 

For each of these species, we construct spawner abundance at the regional scale. We chose to use spawner abundance, rather than catch or run size (i.e., catch plus spawners), because data on spawner abundance is more readily summarized at different spatial scales. Further, spawners represents the abundance of salmon available for to meet cultural and ecological needs and thus provides a measure of status relevant to communities and ecosystems, rather than industry. We recognize that commercial catch has historical been a substantial portion of salmon that return to the coast, and that ignoring declines in catch will underestimate the declines in overall salmon abundance. Trends in catch will be discussed as part of the Harvest section under "Factors" in the SPS report.

For most species and regions, we expanded spawner abundance from river-level estimates to get a regional scale **index** of spawner abundance using two types of expansion factors [@English2016]. Expansion Factor 1 accounts for indicator streams that are not monitored in a given year, and is based on the mean proportional contribution of each indicator stream to the aggregate abundance. Expansion Factor 2 accounts for non-indicator streams that are monitored less rigorously, and is a constant through time calculated based on the average contribution of indicator versus non-indicator stream to aggregate abundance. This expansion process is described in more detail [below](#expansion-factors). For some regions and species, river-level spawner abundance data are not readily available or are not reliable. In particular, stocks that are governed by international treaties may be monitored by the Pacific Salmon Commission, and tend to have better data available at coarser management unit scales. For these regions and species, we may have adapted our approach to incorporate more reliable datasets (see [Region-Specific Deviations](#region-specific-deviations)).

## Expansion Factors
When expanding to regional scale abundance, we started with spawner survey data shown in the Pacific Salmon Explorer. These data are largely derived from river-level estimates in  DFO’s [New Salmon Escapement Database System (NuSEDS)](https://open.canada.ca/data/en/dataset/c48669a3-045b-400d-b730-48aafe8c5ee6), but are cleaned up to address issues of, for example, inconsistent naming of streams through time or duplicate data. The spawner survey abundance is equal to the "maximum estimate" in NUSEDS for each year and river population, calculated as the maximum of all fields containing spawner abundance data (e.g. natural adult spawners, natural jack spawners, total broodstock removals). Each of these river populations has been designated as an "indicator" stream or "non-indicator" stream by DFO. Indicator streams are observed more consistently in recent decades, tend to have higher spawner abundance, and tend to be monitored using more intensive methods that provide greater accuracy [@English2016]. For further information on the compilation of spawner survey data, see the [Pacific Salmon Explorer Technical Report](https://bookdown.org/salmonwatersheds/tech-report/analytical-approach.html#spawner-surveys).

Expansion Factor 1, $F_{1,y/d}$, expands the observed spawner abundances in indicator streams to account for indicator streams that are not monitored in a given year. It is calculated for each year $y$ of the spawner time series, and relies on a decadal contribution of each indicator stream to the total escapement to all indicator streams, $P_{d,i}$ in decade $d$ [@English2016]. The calculation of this decadal contribution requires at least one estimate from each indicator stream for the decade. If a decade does not contain sufficient information (i.e. one or more indicator streams are not monitored at all in a decade), then a reference decade is used to calculate $P_{d,i}$. This reference decade is chosen to be: (1) the closest decade (historical or future) with sufficient information, or failing (1), (2) the 20-year period from 1980-1999 [@Challenger2018].

For each decade (or reference decade if insufficient information) $d$, the average number of spawners returning to indicator stream $i$ is calculated as:

$$\bar{S}_{d,i} = \sum_{y = 1}^{Y_{d,i}} \frac{\hat{S}_{y/d, i}}{Y_{d,i}} $$
where $Y_{d,i}$ is the number of years for which spawner estimates are available within decade $d$ for stream $i$.  From the average number of spawners for all indicator streams, the decadal proportional contribution of each indicator stream is calculated as:

$$P_{d,i} = \frac{\bar{S}_{d,i}}{\sum_{i=1}^{I} \bar{S}_{d,i}}$$
where $I$ is the total number of indicator streams.

Expansion Factor 1 is then calculated for each year within the decade $y/d$ based on the decadal contributions and which streams were monitored or not in a given year:

$$F_{1,y/d}=\left( \sum_{i=1}^I P_{d,i} w_{y/d,i} \right)$$
where $w_{y/d,i}$ is 1 is stream $i$ was monitored in year $y$ and 0 if stream $i$ was not monitored in year $y$. Expansion Factor 1 is then multiplied by the sum of the observed spawners in all indicator stream to yield the expanded estimate of spawner abundances in all indicator streams in the region:

$$S'_{y} = F_{1,y/d} \sum_{i=1}^I \hat{S}_{y,i}$$

Expansion Factor 2 $F_{2,d}$ expands the spawner abundance to all indicator streams, $S'_{y}$, to account for non-indcator streams. Unlike Expansion Factor 1, this is calculated for each decade (rather than each year) and then applied to all years within a decade. Like Expansion Factor 1, there needs to be sufficient information within the given decade in order to calculate $F_{2,d}$, or else a reference decade is chosen. See @English2016 for detailed on how reference decades are chosen in that case.

Expansion Factor 2 is calculated as:

$$F_{2,d} = \frac{\sum_{i = 1}^I \bar{S}_{d,i} + \sum_{j = 1}^{J} \bar{S}_{d,j}}{\sum_{i = 1}^I \bar{S}_{d,i}}$$
where $\bar{S}_{d,i}$ and $\bar{S}_{d,j}$ are the deacdal average number of spawners in indicator and non-indicator streams, respectively, calculated above.  $J$ is the total number of non-indicator streams.  The adjusted total number of spawners in both indicator and non-indicator streams is then calculated as:
$$ S''_{y} = F_{2,d} S'_{y} $$.

Note that when expanding spawner abundance for spawner-recruit analysis, a third expansion factor is applied to account for streams that are never monitored and for observer (in)efficiency [@Peacock2020]. We did not apply this third expansion factor because it is highly undertain and we are interested in relative changes in abundance through time, so we do not require to expand to absolute abundance.

## Region-Specific Deviations

### Fraser
<!-- From More Information for Fraser spawner abundance: https://www.salmonexplorer.ca/#!/fraser/pink/fraser-river-odd&pop=ABUNDANCE_ESTIMATE&pop-detail=1 -->

River-level spawner abundance for Fraser Chinook found in NuSEDS can be highly unreliable, thus we based our expansions on only the most intensively monitored streams that have been deemed as reliable estimates of abundance by @Brown2020. These streams are those shown in the [Pacific Salmon Explorer](https://www.salmonexplorer.ca), and the associated data can be accessed from the stream survey data in the [SWP Data Library](https://data.salmonwatersheds.ca/result?datasetid=1).

In the Fraser region, there are only pink salmon returning in any significant abundance in odd-years, constituting the **Fraser River (odd)** Conservation Unit (CU). There are no spawner estimates for Fraser pink in NuSEDS since 2001. Therefore, spawner abundance is estimated for the Fraser River (odd) pink salmon using three different methods: (1) Up to 2001, estimated spawner abundance was calculated as the sum of observed spawner abundance in NuSEDS, (2) from 2003 to 2007 estimated spawner abundance was based on test fishing expansion, and (3) from 2009 to present estimated spawner abundance is based on [hydroacoustic estimates at Mission](https://www.psc.org/publications/fraser-panel-in-season-information/fish-passage-past-the-psc-hydroacoustic-counting-station-near-mission-bc/). 

### Northern Transboundary

The northern Transboundary watersheds that cross the Canada-US border into Alaska are managed under the Pacific Salmon Treaty. There is considerable effort in assessing the abundance of salmon that cross the border of major watersheds in this region in order to monitor whether escapement goals to Canadian rivers are met, but data on individual river population in NuSEDS can be patchy. As such, we assessed regional abundance in the Transboundary using estimates of border escapement provided by the Pacific Salmon Commission's [Joint Transboundary Technical Committee Reports](https://www.psc.org/publications/technical-reports/technical-committee-reports/transboundary/). The regional abundance of Chinook, coho, and sockeye salmon were calculated as the sum of border escapement to the Stikine, Alsek, and Taku watersheds. We recognize that this approach ignores spawners in the smaller watersheds, specifically the Chilkat, Unuk, and Whiting, but the contribution of these smaller watersheds to total regional abundance for each species is likely small.

<!-- Are we using all three watersheds for all species??  -->

# Results 

:::: {.blackbox data-latex=""}
::: {.center data-latex=""}
**NOTICE!**
:::

**These are PRELIMINARY results and the following figures should not be taken as a statement on the status of salmon in these regions.** At this time, we are working to verify datasets and refine our methods.
::::



```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(viridis)
library(reactable)
library(reactablefmtr)
library(dplyr)

# Set wd as root for project, rather than docs folder
setwd("/Users/stephaniepeacock/Salmon Watersheds Dropbox/Stephanie Peacock/X Drive/1_PROJECTS/1_Active/State of Salmon/Data and Analysis/state-of-salmon")

# Read in data for plots

spawners <- read.csv("output/spawners_all-regions.csv")
regions <- c("Skeena", "Nass", "Haida Gwaii", "Central Coast", "Vancouver Island & Mainland Inlets", "Fraser")

genLength <-read.csv("data/gen_length_regions.csv")

# Set colour palette for species
sp_cols <- viridis(5); names(sp_cols) <- c("Chinook", "Chum", "Coho", "Pink", "Sockeye")

# Source script with code for plotting functions
source("code/functions.R")

```

## Yukon

Coming soon!

## Transboundary

Coming soon!

## Nass

**Highlights**

* While most species show declines over both long- and short-time frames, Nass pink salmon have rebounded to above historical average abundance in the most recent generation.
* Chinook salmon have shown the biggest declines from historical average abundance (-61.6%) with declines persisting in the most recent generation.
* Although sockeye have had a declining trend over the most recent generation, they are above historical average with an uptick in the most recent year.

```{r, echo = FALSE, include = TRUE, fig.width = 8, fig.height = 4, fig.cap = "**Above:** Index of spawner abundance through time for Pacific salmon in the Nass region. Spawner abundance is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance and open points indicate the value of the index one generation prior (note: generation length differs by species). The percent shown is the percent change from the historical average to the most recent year."}
plot.regional_spawners(selected_region = "Nass",
                       spawners = spawners,
                       genLength = genLength,
                       baseline = "average",
                       sp_cols = sp_cols)
```

**Below:** Summary of the index spawner abundance (expanded from river-level spawner estimates) and short- and long-term trends by species.
```{r, echo = FALSE, include = TRUE}

btn_table(
  selected_region = "Nass",
  spawners = spawners,
  genLength = genLength
)
```


## Skeena

**Highlights**

* All species are well below historical average spawner abundance, with chum salmon having declined by 96.5%.
* Pink salmon are the only species showing an increase in the most recent generation, though they are still well below historical average.
* Estimates of regional spawner abundance have become increasingly uncertain since 2000, especially for Chinook and coho, due to a decline in monitoring of indicator streams.

<!-- ```{r, echo = FALSE, include = TRUE, fig.width = 8, fig.height = 4, fig.cap="Index of spawner abundance through time for Pacific salmon in the Skeena region. Spawner abundance is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the smoothed abundance in 1980 when escapement monitoring effort was generally high. Closed points highlight the most recent estimate and the percent change is from the spawner abundance in the previous generation (open point)."} -->
<!-- plot.regional_spawners(selected_region = "Skeena", -->
<!--                        spawners = spawners, -->
<!--                        genLength = genLength, -->
<!--                        baseline = 1980, -->
<!--                        sp_cols = sp_cols) -->

<!-- ``` -->

```{r, echo = FALSE, include = TRUE, fig.width = 8, fig.height = 4, fig.cap="**Above:** Index of spawner abundance through time for Pacific salmon in the Skeena region. Spawner abundance is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance and open points indicate the value of the index one generation prior (note: generation length differs by species). The percent shown is the percent change from the historical average to the most recent year."}

plot.regional_spawners(selected_region = "Skeena",
                       spawners = spawners,
                       genLength = genLength,
                       baseline = "average",
                       sp_cols = sp_cols)
```


**Below:** Summary of the index spawner abundance (expanded from river-level spawner estimates) and short- and long-term trends by species.
```{r, echo = FALSE, include = TRUE}

btn_table(
  selected_region = "Skeena",
  spawners = spawners,
  genLength = genLength
)
```
## Haida Gwaii

**Highlights**

* All species for which data were available have declined over long- and short-time frames.
* Unlike in other regions, where pijnk salmon have been doing relatively well, pink salmon have shown the most dramatic declines, and most recent abundance was 90.9% lower than the previous generation and 92.3% lower than the historical average.
* There is only one indicator stream for Chinook in Haida Gwaii, which has not been consistently monitored, making it impossible to reconstruct Chinook abundance at the regional scale.

```{r, echo = FALSE, include = TRUE, warning=FALSE, message=FALSE, fig.width = 8, fig.height = 4, fig.cap = "**Above:** Index of spawner abundance through time for Pacific salmon in the Haida Gwaii region. Spawner abundance is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance and open points indicate the value of the index one generation prior (note: generation length differs by species). The percent shown is the percent change from the historical average to the most recent year."}
plot.regional_spawners(selected_region = "Haida Gwaii",
                       spawners = spawners,
                       genLength = genLength,
                       baseline = "average",
                       sp_cols = sp_cols)
```

**Below:** Summary of the index spawner abundance (expanded from river-level spawner estimates) and short- and long-term trends by species.
```{r, echo = FALSE, include = TRUE}

btn_table(
  selected_region = "Haida Gwaii",
  spawners = spawners,
  genLength = genLength
)
```

## Central Coast

**Highlights**

* All species are below historical average abundance, with dramatic declines of culturally and economically important sockeye, Chinook, and chum salmon.
* Pink salmon are the only species that has increased from the previous generation, more than doubling between 2019 and 2021.
* Expansion factors for regional-scale reconstructions have increased substantially since the mid-2000s, with a decline in monitoring creating uncertainties in our results for all species, but especially coho, sockeye, and Chinook.

```{r, echo = FALSE, include = TRUE, fig.width = 8, fig.height = 4, fig.cap = "**Above:** Index of spawner abundance through time for Pacific salmon in the Central Coast region. Spawner abundance is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance and open points indicate the value of the index one generation prior (note: generation length differs by species). The percent shown is the percent change from the historical average to the most recent year."}
plot.regional_spawners(selected_region = "Central Coast",
                       spawners = spawners,
                       genLength = genLength,
                       baseline = "average",
                       sp_cols = sp_cols)
```

**Below:** Summary of the index spawner abundance (expanded from river-level spawner estimates) and short- and long-term trends by species.
```{r, echo = FALSE, include = TRUE}

btn_table(
  selected_region = "Central Coast",
  spawners = spawners,
  genLength = genLength
)
```

## Vancouver Island & Mainland Inlets

**Highlights**

* This region has seen some of the more promising changes in recent years, with two species (coho and pink) increasing in the most recent generation (although both these species are still below historical average).
* Dynamics through time are slightly more stable for the regional aggregate because there are many component  populations in this large area, but some local trends are concerning (e.g., WCVI Chinook).
* (Something about recent rebound of pink salmon continuing to 2023 year..?)

```{r, echo = FALSE, include = TRUE, fig.width = 8, fig.height = 4, fig.cap = "**Above:** Index of spawner abundance through time for Pacific salmon in the Vancouver Island & Mainland Inlets region. Spawner abundance is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance and open points indicate the value of the index one generation prior (note: generation length differs by species). The percent shown is the percent change from the historical average to the most recent year."}
plot.regional_spawners(selected_region = "Vancouver Island & Mainland Inlets",
                       spawners = spawners,
                       genLength = genLength,
                       baseline = "average",
                       sp_cols = sp_cols)
```


**Below:** Summary of the index spawner abundance (expanded from river-level spawner estimates) and short- and long-term trends by species.
```{r, echo = FALSE, include = TRUE}

btn_table(
  selected_region = "Vancouver Island & Mainland Inlets",
  spawners = spawners,
  genLength = genLength
)

```
## Fraser

**Highlights**

* Despite the concern over Fraser sockeye abundance since the crash of 2009, sockeye spawner abundance is still below the long-term average and has declined in the most recent generation.
* Other species are actually above historical average, **however** the these trends do not account for declining catch of the commercial fishing industry in this area. Because fishing pressure on Fraser salmon has historically been higher than in other regions, declines in overall run size (i.e., spawners + catch) mean that overall abundance has likely declined for all species. (**SP: We could include a plot of catch in Fraser areas for each species...)
* Although coho spawners have increased over both short- and long-time frames, coho fisheries have been decimated since concerns over Fraser coho stocks in the mid-1990s.
* Recent declines in Chinook, chum, pink, and sockeye may reflect the increasing pressures of climate change on these southern and interior populations.

```{r, echo = FALSE, include = TRUE, fig.width = 8, fig.height = 4, fig.cap = "**Above:** Index of spawner abundance through time for Pacific salmon in the Fraser region. Spawner abundance is smoothed using a one-generation geometric running average (right-aligned), and plotted relative to the historical average for each species. Closed points highlight the most recent index of spawner abundance and open points indicate the value of the index one generation prior (note: generation length differs by species). The percent shown is the percent change from the historical average to the most recent year."}
plot.regional_spawners(selected_region = "Fraser",
                       spawners = spawners,
                       genLength = genLength,
                       baseline = "average",
                       sp_cols = sp_cols)
```

**Below:** Summary of the index spawner abundance (expanded from river-level spawner estimates) and short- and long-term trends by species.
```{r, echo = FALSE, include = TRUE}

btn_table(
  selected_region = "Fraser",
  spawners = spawners,
  genLength = genLength
)
```

## Columbia

Coming soon!


# References

<div id="refs"></div>

# Supplemental results

## Expansion Factors

The following figures show the values of Expansion factors 1 (solid line) and 2 (dashed line) that were multiplied by the sum of observed spawners return to indicator stream to calculate the regional index of spawner abundance for each species.

```{r, echo = FALSE, include = TRUE, fig.width = 8, fig.height = 4, fig.cap=""}
for(r in regions){
  plot.regional_expansion(selected_region = r,
                       spawners = spawners,
                       sp_cols = sp_cols)
}
```